---
title: "PHS EDA"
author: "Natalia Montalvo Cabornero"
date: "2026-08-14"
output: html_document
---

```{r setup, include=FALSE}

# prep.R handles package loading, helper functions, data import and cleaning.
# risk_model_data.R sources prep.R internally, then builds SIMD aggregations,
# shared context variables, and the risk model datasets.
source("R/prep.R")

# Make sure the figure output folders exist
dir.create("figures/eda", recursive = TRUE, showWarnings = FALSE)
```

# 1. Monthly demographics data set

```{r}
summarise_columns(month_demo_data)
head(month_demo_data)
```

## 1.1 Monthly and long term attendance trends

```{r}
# Monthly demographics year trends
month_trend_data <- month_demo_data %>%
  group_by(year, m_num) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop")

# Plot attendances by month over years
m_demo_trend_plt <- ggplot(month_trend_data, 
                           aes(x = m_num, y = total, color = year, 
                               group = year)) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(x = "Month", 
       y = "Total Attendances",
       title = "A&E Attendances by Month Across Years") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

m_demo_trend_plt

ggsave("figures/eda/m_demo_trend_by_year.pdf", plot = m_demo_trend_plt,
       width = 8, height = 5)

```

```{r}
# Overall monthly attendance trend (all demographics combined)
overall_monthly_trend <- month_demo_data %>%
  group_by(m_date) %>%
  summarise(
    total_attendances = sum(NumberOfAttendances, na.rm = TRUE),
    .groups = "drop")

overall_monthly_trend_plt <- ggplot(
  overall_monthly_trend,
  aes(x = m_date, y = total_attendances)) +
  geom_line(linewidth = 0.8) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    x = NULL,
    y = "Total Attendances",
    title = "Monthly A&E Attendances Over Time"
  ) +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))
  

overall_monthly_trend_plt

ggsave("figures/eda/overall_monthly_trend.pdf", plot = overall_monthly_trend_plt,
       width = 9, height = 5)
```

## 1.2 Patterns by sex

```{r}
# Divide data by sex
sex_data <- month_demo_data %>%
  filter(Sex %in% c("Male", "Female")) %>%
  group_by(Sex) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop")

# Plot sex differences in attendance
sex_plt <- ggplot(sex_data, aes(x = Sex, y = total, fill = Sex)) +
  geom_col() +
  scale_fill_discrete_phs(palette = "main") +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Attendances by Sex", 
       x = "Sex", 
       y = "Total Attendances") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

sex_plt

ggsave("figures/eda/attendances_by_sex.pdf", plot = sex_plt, width = 6, 
       height = 5)
```

```{r}
# Monthly demographics year trends by sex
month_trend_sex_data <- month_demo_data %>%
  filter(Sex %in% c("Male", "Female")) %>%
  group_by(year, m_num, Sex) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop")

month_trend_sex_plt <- ggplot(month_trend_sex_data, 
                              aes(x = m_num, y = total,
                                  color = year, group = year)) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  facet_wrap(~ Sex, ncol = 2) +
  labs(x = "Month", y = "Total Attendances",
       title = "A&E Attendances by Month, Year, and Sex") +
  theme_phs() +
  theme(axis.text.x = element_text(size = 11, angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))

month_trend_sex_plt

ggsave("figures/eda/montrend_by_year_sex.pdf", plot = month_trend_sex_plt,
       width = 10, height = 5)
```


## 1.3 Patterns by deprivation

```{r}
# Divide attendances by deprivation index
deprivation_data <- month_demo_data %>%
  filter(!is.na(Deprivation)) %>%
  group_by(Deprivation) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop") %>%
  mutate(Deprivation = factor(Deprivation, levels = 1:5,
           labels = c("1 (most deprived)", "2", "3", "4", "5 (least deprived)")))

# Plotting attendances by deprivation quintile
deprivation_plt <- ggplot(deprivation_data, aes(x = Deprivation, y = total)) +
  geom_col(fill = phs_colours("phs-graphite")) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Deprivation Quintile", y = "Total Attendances",
       title = "Attendances by Deprivation Quintile") +
  theme_phs() +
  theme(axis.text.x = element_text(size = 11, angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))

deprivation_plt

ggsave("figures/eda/attendances_by_deprivation.pdf", plot = deprivation_plt, width = 8, height = 5)
```

```{r}
# Divide attendances by deprivation index and type of department
deprivation_dept_data <- month_demo_data %>%
  filter(!is.na(Deprivation)) %>%
  group_by(Deprivation, DepartmentType) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), 
            .groups = "drop") %>%
  mutate(Deprivation = factor(Deprivation))

# Plotting attendances by deprivation quintile and department type
deprivation_dept_plt <- ggplot(deprivation_dept_data, 
                               aes(x = Deprivation, y = total)) +
  geom_col(fill = phs_colours("phs-graphite")) +
  facet_wrap(~ DepartmentType, scales = "free_y") +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Deprivation Quintile", 
       y = "Total Attendances",
       title = "Attendances by Deprivation Quintile and Department Type") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

deprivation_dept_plt

ggsave("figures/eda/deprivation_by_dept.pdf", plot = deprivation_dept_plt, 
       width = 10, height = 6)
```

```{r}
# Divide attendances by deprivation index and sex
deprivation_sex_data <- month_demo_data %>%
  filter(!is.na(Deprivation), Sex %in% c("Male", "Female")) %>%
  group_by(Deprivation, Sex) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), 
            .groups = "drop") %>%
  mutate(Deprivation = factor(Deprivation))

# Plotting attendances by deprivation quintile and sex
deprivation_sex_plt <- ggplot(deprivation_sex_data, 
                              aes(x = Deprivation, y = total, fill = Sex)) +
  geom_col(position = "dodge") +
  scale_fill_discrete_phs(palette = "main") +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Deprivation Quintile", y = "Total Attendances",
       title = "Attendances by Deprivation Quintile and Sex") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

deprivation_sex_plt

ggsave("figures/eda/deprivation_by_sex.pdf", plot = deprivation_sex_plt, 
       width = 8, height = 5)
```

## 1.4 Patterns by age

```{r}
# Data by age groups
age_data <- month_demo_data %>%
  filter(!is.na(Age)) %>%
  group_by(Age) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop")

# Plot attendances by age groups
age_plt <- ggplot(age_data, aes(x = Age, y = total)) +
  geom_col(fill = phs_colours("phs-teal")) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Age Group", y = "Total Attendances", title = "Attendances by Age Group") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

age_plt

ggsave("figures/eda/attendances_by_age.pdf", plot = age_plt, width = 8, height = 5)
```

```{r}
# Attendances by age group and sex (raw counts)
age_sex_attendance <- month_demo_data %>%
  filter(!is.na(Age), Sex %in% c("Male", "Female")) %>%
  group_by(Age, Sex) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), 
            .groups = "drop")

age_sex_attendance_plt <- ggplot(age_sex_attendance,
                                 aes(x = Age, y = total, fill = Sex)) +
  geom_col(position = "dodge") +
  scale_fill_discrete_phs(palette = "main") +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Age Group",
       y = "Total Attendances",
       title = "A&E Attendances by Age Group and Sex") +
  theme_phs() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        plot.title = element_text(hjust = 0.5))

age_sex_attendance_plt

ggsave("figures/eda/attendances_by_age_and_sex.pdf", 
       plot = age_sex_attendance_plt, width = 9, height = 5)
```

```{r}
# Deprivation converted to a labelled factor here 
age_deprivation_data <- month_demo_data %>%
  filter(!is.na(Age),!is.na(Deprivation)) %>%
  group_by(Age, Deprivation) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), 
            .groups = "drop") %>%
  mutate(Deprivation = factor(Deprivation, levels = 1:5,
           labels = c("1 (most deprived)", "2", "3", "4", "5 (least deprived)")))

age_deprivation_plt <- ggplot(age_deprivation_data, 
                              aes(x = Age, y = total, fill = Deprivation)) +
  geom_col(position = "fill") +
  scale_fill_manual(values = phs_colours(c("phs-purple", "phs-magenta", 
                                           "phs-blue", "phs-green", 
                                           "phs-graphite"))) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Age Group",
       y = "Share of Attendances",
       fill = "SIMD Quintile",
       title = "Deprivation Composition of Attendances by Age Group") +
  theme_phs() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        plot.title = element_text(hjust = 0.5))

age_deprivation_plt

ggsave("figures/eda/age_deprivation_composition.pdf", 
       plot = age_deprivation_plt, width = 9, height = 5)
```


## 1.5 Distribution of attendance counts

```{r}
# Check for zeros first
cat("Rows with zero attendances:", 
    sum(month_demo_data$NumberOfAttendances == 0, na.rm = TRUE), "\n")

# Distribution of attendances
attendances_hist_plt <- ggplot(month_demo_data, 
                               aes(x = NumberOfAttendances)) +
  geom_histogram(bins = 50, fill = phs_colours("phs-purple"), alpha = 0.85) +
  labs(x = "Number of Attendances", 
       y = "Count",
       title = "Distribution of Number Of Attendances") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

attendances_hist_plt

ggsave("figures/eda/attendances_histogram.pdf", plot = attendances_hist_plt, 
       width = 8, height = 5)
```


```{r}
# Distribution of attendances log scale
attendances_hist_log_plt <- ggplot(month_demo_data, 
                                   aes(x = NumberOfAttendances)) +
  geom_histogram(bins = 50, fill = phs_colours("phs-purple"), alpha = 0.85) +
  scale_x_log10() +
  labs(x = "Number of Attendances (log scale)", 
       y = "Count",
       title = "Distribution of Number Of Attendances (log scale)") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

attendances_hist_log_plt

ggsave("figures/eda/attendances_histogram_log.pdf", 
       plot = attendances_hist_log_plt, width = 8, height = 5)
```

# 2. Activity data

This section inspects the weekly and monthly activity datasets used to describe
overall A&E service activity.

## 2.1 Weekly data

The weekly activity data set contains Type 1 emergency departments only. It
covers 32 treatment locations across 14 Health Boards.

```{r}
summarise_columns(week_activity_data)
head(week_activity_data)
tail(week_activity_data)
```

```{r}
# Confirm the number of treatment locations and Health Boards
n_distinct(week_activity_data$TreatmentLocation)
n_distinct(week_activity_data$HBT)

# Check the available attendance categories
distinct(week_activity_data, AttendanceCategory)
```

## 2.2 Monthly data

```{r}
summarise_columns(month_activity_data)
head(month_activity_data)
tail(month_activity_data)
```

```{r monthly-activity-coverage}
# Check the department types, treatment locations and Health Boards
distinct(month_activity_data, DepartmentType)

n_distinct(month_activity_data$TreatmentLocation)
n_distinct(month_activity_data$HBT)

# Confirm the available monthly date range
range(month_activity_data$m_date, na.rm = TRUE)
```

# 3. Treatment location data

This section links the weekly Type 1 activity data to the treatment location 
lookup. It then checks reporting coverage and compares average weekly activity 
between sites.

```{r}
summarise_columns(trt_loc)
head(trt_loc)
```

```{r}
# Add location info to week activity data set
week_activity_data <- week_activity_data %>%
  left_join(trt_loc, by = c("TreatmentLocation" = "TreatmentLocationCode"))

week_activity_data %>%
  filter(is.na(TreatmentLocationName)) %>%
  distinct(TreatmentLocation)

week_activity_data %>%
  distinct(TreatmentLocation, TreatmentLocationName, HB, CurrentDepartmentType, 
           Status)
```

```{r}
# How many weeks per location in data set
location_nweeks <- week_activity_data %>%
  filter(AttendanceCategory == "All") %>%
  count(TreatmentLocation, name = "n_weeks_reported") %>%
  left_join(trt_loc, by = c("TreatmentLocation" = "TreatmentLocationCode")) %>%
  select(TreatmentLocation, TreatmentLocationName, HB, 
         Status, n_weeks_reported) %>%
  arrange(n_weeks_reported)

location_nweeks
```

- G516H: West Glasgow Ambulatory Care Hospital S08000031 Closed (15 weeks)
- G306H: New Victoria Hospital S08000031Ope n (13 weeks)

## 3.1 Average weekly attendance by location

```{r}
# Average attendance per location
location_attendance <- week_activity_data %>%
  filter(AttendanceCategory == "All") %>%
  group_by(TreatmentLocation) %>%
  summarise(
    n_weeks_reported = n(),
    total_attendance = sum(NumberOfAttendancesEpisode, na.rm = TRUE),
    avg_weekly_attendance = mean(NumberOfAttendancesEpisode, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(trt_loc, by = c("TreatmentLocation" = "TreatmentLocationCode")) %>%
  select(TreatmentLocation, TreatmentLocationName, HB, Status,
         n_weeks_reported, total_attendance, avg_weekly_attendance) %>%
  arrange(desc(total_attendance))

location_attendance

# Plot locations by number of attendances
location_attendance_plt <- ggplot(location_attendance, 
                                    aes(x = fct_reorder(TreatmentLocationName, 
                                                        avg_weekly_attendance), 
                                        y = avg_weekly_attendance)) +
  geom_col(fill = phs_colours("phs-blue")) +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  labs(x = NULL, y = "Average Weekly Attendances",
       title = "NHS Scotland A&E Locations by Average Weekly Attendance") +
  theme_phs() +
  theme(axis.text.y = element_text(size = 9), 
        panel.grid.major.y = element_blank())

location_attendance_plt

ggsave("figures/eda/location_avg_attendance.pdf", 
       plot = location_attendance_plt, width = 8, height = 10)
```


```{r}
# Average attendance per location per year (kept for the trend plot below)
location_year_attendance <- week_activity_data %>%
  filter(AttendanceCategory == "All") %>%
  group_by(year, TreatmentLocation) %>%
  summarise(
    n_weeks_reported = n(),
    total_attendance = sum(NumberOfAttendancesEpisode, na.rm = TRUE),
    avg_weekly_attendance = mean(NumberOfAttendancesEpisode, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(trt_loc, by = c("TreatmentLocation" = "TreatmentLocationCode")) %>%
  select(year, TreatmentLocation, TreatmentLocationName, HB, Status,
         n_weeks_reported, total_attendance, avg_weekly_attendance) %>%
  arrange(desc(total_attendance))

location_year_attendance

# Top 10 locations — weighted by actual weeks reported
top10_locations <- week_activity_data %>%
  filter(AttendanceCategory == "All",
         !is.na(TreatmentLocationName)) %>%
  group_by(TreatmentLocation, TreatmentLocationName) %>%
  summarise(
    n_weeks_reported = n(),
    overall_avg = mean(NumberOfAttendancesEpisode, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  slice_max(order_by = overall_avg, n = 10, with_ties = FALSE) %>%
  arrange(desc(overall_avg))

top10_names <- top10_locations$TreatmentLocationName
top10_codes <- top10_locations$TreatmentLocation

# Trend plot for top 10 locations
top10_trend_overlay_plt <- location_year_attendance %>%
  filter(TreatmentLocationName %in% top10_names) %>%
  mutate(TreatmentLocationName = factor(TreatmentLocationName, 
                                        levels = top10_names)) %>%
  ggplot(aes(x = year, 
             y = avg_weekly_attendance, 
             colour = TreatmentLocationName, group = TreatmentLocationName)) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Year", y = "Average Weekly Attendances", colour = "Location",
       title = "Top 10 A&E Locations: Attendance Trend by Year") +
  theme_phs() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
        legend.position = "right",
        legend.text = element_text(size = 9),
        legend.title = element_text(size = 11),
        legend.key.size = unit(0.4, "cm"))

top10_trend_overlay_plt

ggsave("figures/eda/top10_location_trend_overlay.pdf", 
       plot = top10_trend_overlay_plt, width = 11, height = 6)
```


# 4. Population data

This section examines the population data used to describe demographic 
structure and construct population adjusted attendance rates.

```{r}
head(pop_data)
summarise_columns(pop_data)
```

```{r}
# Confirm the available sex categories
distinct(pop_data, Sex)

# Count the number of unique Data Zones 
n_distinct(pop_data$DataZone) 

# Compare Data Zone coverage across the population and SIMD datasets
n_distinct(simd_ind$Data_Zone)             

# Population Data Zones with no corresponding SIMD classification
setdiff(unique(pop_data$DataZone), unique(simd_main$DataZone))

# SIMD Data Zones with no corresponding population record
setdiff(unique(simd_main$DataZone), unique(pop_data$DataZone))
```


<!-- ```{r} -->
<!-- # Compare totals -->
<!-- sum(pop_hb_density$population, na.rm = TRUE) -->
<!-- sum(simd_hb$quintile_population, na.rm = TRUE)   -->
<!-- ``` -->


## 4.1 National population structure by age and sex

```{r}
# Population by age group and sex, nationally
pop_age_sex_plt <- pop_hb_age_sex %>%
  group_by(Age, Sex) %>%
  summarise(population = sum(population, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = Age, y = population, fill = Sex)) +
  geom_col(position = "dodge") +
  scale_fill_discrete_phs(palette = "main") +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Age Group", y = "Population", 
       title = "Scotland Population by Age Group and Sex") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

pop_age_sex_plt

ggsave("figures/eda/population_by_age_sex.pdf", plot = pop_age_sex_plt, 
       width = 8, height = 5)
```
## 4.2 Population density by Health Board

```{r}
# Population density by health board
pop_hb_density <- pop_hb_density %>%
  left_join(hb_names, by = c("HB" = "HBT"))

pop_density_plt <- ggplot(pop_hb_density, aes(x = fct_reorder(HBName, density),
                                              y = density)) +
  geom_col(fill = phs_colours("phs-liberty")) +
  coord_flip() +
  labs(x = "Health Board", y = "Population Density (per km\u00b2)",
       title = "Population Density by Health Board") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

pop_density_plt

ggsave("figures/eda/population_density_by_hb.pdf", plot = pop_density_plt, 
       width = 8, height = 6)
```

## 4.3 Population adjusted attendance rates by age and sex

```{r}
# Attendances by year, age and sex
attendance_age_sex_year <- month_demo_data %>%
  filter(!is.na(Age), Sex %in% c("Male", "Female")) %>%
  group_by(year, Age, Sex) %>%
  summarise(total_attendances = sum(NumberOfAttendances, na.rm = TRUE),
            .groups = "drop")

# Year-varying population by age and sex 
pop_age_sex_year <- pop_dz %>%
  filter(Sex %in% c("Male", "Female"), !is.na(Age)) %>%
  group_by(Year, Age, Sex) %>%
  summarise(population = sum(population, na.rm = TRUE), .groups = "drop")

# Join and compute annual rates, then average across years
age_sex_rates <- attendance_age_sex_year %>%
  mutate(pop_year = pmin(as.integer(year), max(pop_age_sex_year$Year))) %>%
  left_join(pop_age_sex_year, by = c("pop_year" = "Year", "Age", "Sex")) %>%
  mutate(rate_per_1000 = total_attendances / population * 1000) %>%
  group_by(Age, Sex) %>%
  summarise(attendance_rate = mean(rate_per_1000, na.rm = TRUE),
            .groups = "drop")

age_sex_rate_plt <- ggplot(age_sex_rates, aes(x = Age, y = attendance_rate,
                                               colour = Sex, group = Sex)) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_colour_discrete_phs(palette = "main") +
  labs(x = "Age group", y = "Attendances per 1,000 Population (Annual Average)",
       title = "Population-adjusted A&E Attendance Rates by Age and Sex") +
  theme_phs() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))

age_sex_rate_plt

ggsave("figures/eda/age_sex_attendance_rate.pdf", plot = age_sex_rate_plt,
       width = 8, height = 5)
```

## 4.4 Population adjusted attendance rates by Health Board

```{r}
# Aggregate annual attendance volume by Health Board
hb_attendance_year <- month_demo_data %>%
  group_by(year, HBT) %>%
  summarise(total_attendance = sum(NumberOfAttendances, na.rm = TRUE),
            .groups = "drop")

# Link Data Zone population records to Health Boards
pop_hb_year <- pop_dz %>%
  left_join(simd_main %>% select(DataZone, HB), by = "DataZone") %>%
  group_by(Year, HB) %>%
  summarise(population = sum(population, na.rm = TRUE), .groups = "drop")

# Calculate annual Health Board attendance rates
hb_attendance_rates <- hb_attendance_year %>%
  mutate(pop_year = pmin(as.integer(year), max(pop_hb_year$Year))) %>%
  left_join(pop_hb_year, by = c("pop_year" = "Year", "HBT" = "HB")) %>%
  mutate(rate_per_1000 = total_attendance / population * 1000) %>%
  group_by(HBT) %>%
  summarise(rate_per_1000 = mean(rate_per_1000, na.rm = TRUE),
            .groups = "drop") %>%
  left_join(pop_hb_density %>% select(HB, density), by = c("HBT" = "HB")) %>%
  left_join(hb_names, by = "HBT")

hb_rate_plt <- ggplot(
  hb_attendance_rates,
  aes(x = fct_reorder(HBName, rate_per_1000), y = rate_per_1000)) +
  geom_col(fill = phs_colours("phs-blue")) +
  coord_flip() +
  labs(x = NULL, y = "Attendances per 1,000 Population",
       title = "Population-adjusted Attendance Rates by Health Board") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

hb_rate_plt

ggsave("figures/eda/hb_attendance_rate.pdf", plot = hb_rate_plt, 
       width = 8, height = 6)
```

# 5. Referrals data

This section examines how patients were referred to A&E overall, across age groups, and within the ten highest volume treatment locations.

```{r}
summarise_columns(referrals_data)
head(referrals_data)
```

## 5.1 Overall attendance volume by referral source

```{r}
# Attendance by referral type in general
referral_types <- referrals_data %>%
  mutate(Referral = case_when(
    is.na(Referral) | trimws(Referral) %in% c("", ":") ~ "Not Known",
    TRUE ~ Referral
  )) %>%
  group_by(Referral) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop")

referrals_type_plt <- ggplot(referral_types, 
                              aes(x = fct_reorder(Referral, total), 
                                  y = total)) +
  geom_col(fill = phs_colours("phs-graphite")) +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  labs(x = NULL, y = "Total Attendances",
       title = "Attendances by Referral Source") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

referrals_type_plt

ggsave("figures/eda/attendances_by_referral.pdf", plot = referrals_type_plt, 
       width = 8, height = 5)
```

## 5.2 Referral source patterns by age group

```{r}
# Attendance by referral type and age
referral_age_data <- referrals_data %>%
  filter(!is.na(Age)) %>%
  mutate(Referral = case_when(
    is.na(Referral) | trimws(Referral) %in% c("", ":") ~ "Not Known",
    TRUE ~ Referral
  )) %>%
  group_by(Referral, Age) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop")

referral_age_plt <- ggplot(referral_age_data, 
                            aes(x = Age, y = total, fill = Referral)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = phs_colours(c("phs-purple", "phs-magenta", 
                                           "phs-blue", "phs-green",
                                           "phs-graphite", "phs-teal", 
                                           "phs-liberty", "phs-rust"))) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Age Group", y = "Total Attendances", fill = "Referral Source",
       title = "Attendances by Age Group and Referral Source") +
  theme_phs() +
  theme(legend.text = element_text(size = 9), legend.key.size = unit(0.4, "cm"),
        plot.title = element_text(hjust = 0.5))

referral_age_plt

ggsave("figures/eda/referral_by_age.pdf", plot = referral_age_plt, width = 11, 
       height = 6)
```

## 5.3 Referral composition in the ten highest volume locations

```{r}
referral_top10_data <- referrals_data %>%
  filter(!is.na(Referral), Referral != "", 
         TreatmentLocation %in% top10_codes) %>%
  left_join(trt_loc, by = c("TreatmentLocation" = "TreatmentLocationCode")) %>%
  group_by(TreatmentLocationName, Referral) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop")

referral_top10_pct <- referral_top10_data %>%
  group_by(TreatmentLocationName) %>%
  mutate(pct = total / sum(total) * 100) %>%
  ungroup()

referral_table_plt <- ggplot(referral_top10_pct, 
                                aes(x = Referral, 
                                    y = TreatmentLocationName, fill = pct)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = round(pct, 0)), size = 3, colour = "black") +
  scale_fill_gradient(low = "white", high = phs_colours("phs-blue"), 
                      name = "% of\nattendances") +
  labs(x = "Referral Source", y = NULL,
       title = "Referral Source Percentages by Top 10 A&E Locations") +
  theme_phs() +
  theme(axis.text.x = element_text(size = 9, angle = 45, hjust = 1),
        axis.text.y = element_text(size = 9),
        panel.grid = element_blank())

referral_table_plt

ggsave("figures/eda/referral_table_top10.pdf", plot = referral_table_plt, width = 10, height = 7)
```


# 6. Discharges

This section examines recorded discharge destinations and defines the admission outcome used in the admission model. 

```{r}
summarise_columns(discharges_data)
head(discharges_data)
```

## 6.1 Overall attendance volume by discharge destination

```{r}
# Attendance by discharge type in general
discharge_types <- discharges_data %>%
  mutate(Discharge = case_when(
    is.na(Discharge) | trimws(Discharge) %in% c("", ":") ~ "Not Known",
    TRUE ~ Discharge
  )) %>%
  group_by(Discharge) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop")

discharge_type_plt <- ggplot(discharge_types, 
                              aes(x = fct_reorder(Discharge, total), 
                                  y = total)) +
  geom_col(fill = phs_colours("phs-graphite")) +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  labs(x = NULL, y = "Total Attendances",
       title = "Attendances by Discharge Type") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

discharge_type_plt

ggsave("figures/eda/attendances_by_discharge.pdf", plot = discharge_type_plt, 
       width = 8, height = 5)
```

## 6.2 Discharge patterns by age group

```{r}
discharge_age_data <- discharges_data %>%
  filter(!is.na(Age)) %>%
  mutate(Discharge = case_when(
    is.na(Discharge) | trimws(Discharge) %in% c("", ":") ~ "Not Known",
    TRUE ~ Discharge
  )) %>%
  group_by(Discharge, Age) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop")

discharge_age_plt <- ggplot(discharge_age_data, 
                            aes(x = Age, y = total, fill = Discharge)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = phs_colours(c("phs-purple", "phs-magenta", 
                                           "phs-blue", "phs-green",
                                           "phs-teal"))) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Age Group", y = "Total Attendances", fill = "Discharge Type",
       title = "Attendances by Age Group and Discharge Type") +
  theme_phs() +
  theme(legend.text = element_text(size = 9), legend.key.size = unit(0.4, "cm"),
        plot.title = element_text(hjust = 0.5))

discharge_age_plt

ggsave("figures/eda/discharge_by_age.pdf", plot = discharge_age_plt, width = 11, 
       height = 6)
```

## 6.2 Admission data

For this analysis, an admission was defined as an attendance recorded with the discharge destination "Admission to same Hospital".

```{r}
# Admission rate over time
admission_rate_trend <- discharges_data %>%
  group_by(m_date) %>%
  summarise(
    total_attendances = sum(NumberOfAttendances, na.rm = TRUE),
    admitted = sum(NumberOfAttendances[
      Discharge == "Admission to same Hospital"], 
      na.rm = TRUE), 
    .groups = "drop"
  ) %>%
  mutate(admission_rate = admitted / total_attendances * 100)

admission_rate_plt <- ggplot(admission_rate_trend, aes(x = m_date, 
                                                       y = admission_rate)) +
  geom_line(linewidth = 0.8, colour = phs_colours("phs-rust")) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1, 
              colour = phs_colours("phs-blue")) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(x = NULL, y = "Admission Rate (%)",
       title = "A&E Admission Rate Over Time (All Department Types)") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

admission_rate_plt

ggsave("figures/eda/admission_rate_trend.pdf", plot = admission_rate_plt,
       width = 9, height = 5)
```

## 6.4 Admission rates by department type

```{r}
# Admission rates by department type
discharges_data %>%
  group_by(DepartmentType) %>%
  summarise(
    total_attendances = sum(NumberOfAttendances, na.rm = TRUE),
    admitted = sum(NumberOfAttendances[Discharge == "Admission to same Hospital"],
                   na.rm = TRUE),
    admission_rate = admitted / total_attendances * 100,
    .groups = "drop"
  ) %>%
  mutate(pct_of_all_attendances = total_attendances / sum(total_attendances) * 100)
```


The overall admission rate includes both Type 1 (major ED, 27.3% admission 
rate) and Type 3 (minor injury, 8.9% admission rate).

```{r}
# Admission rate over time Type 1
admission_rate_t1 <- discharges_data %>%
  filter(DepartmentType == "Type 1") %>%
  group_by(m_date) %>%
  summarise(
    total_attendances = sum(NumberOfAttendances, na.rm = TRUE),
    admitted = sum(NumberOfAttendances[
      Discharge == "Admission to same Hospital"], 
      na.rm = TRUE), 
    .groups = "drop"
  ) %>%
  mutate(admission_rate = admitted / total_attendances * 100)

admission_rate_t1_plt <- ggplot(admission_rate_t1, aes(x = m_date, 
                                                    y = admission_rate)) +
  geom_line(linewidth = 0.8, colour = phs_colours("phs-rust")) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1, 
              colour = phs_colours("phs-blue")) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(x = NULL, y = "Admission Rate for Type 1 (%)",
       title = "A&E Admission Rate Over Time (Type 1 Only)") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

admission_rate_t1_plt

ggsave("figures/eda/admission_rate_type1.pdf", plot = admission_rate_t1_plt,
       width = 9, height = 5)
```


# 7. Multiple attendances

This section examines how frequently patients attended A&E within a 12-month period

```{r}
summarise_columns(mul_att_data)
head(mul_att_data)
tail(mul_att_data)
```

## 7.1 Overall distribution of attendance frequency

`mul_att_long` is created in prep.R, reshaping from wide to long so it's 
easier to compute the summaries.

```{r}
freq_totals <- mul_att_long %>%
  group_by(attendance_freq) %>%
  summarise(n_patients = sum(n_patients, na.rm = TRUE), .groups = "drop")

att_freq_plot <- ggplot(freq_totals, aes(x = attendance_freq, 
                                         y = n_patients)) +
  geom_col(fill = phs_colours("phs-blue")) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Number of Attendances (12-month period)", 
       y = "Number of Patients",
       title = "Distribution of A&E Attendance Frequency per Patient") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

att_freq_plot 

ggsave("figures/eda/attendance_frequency.pdf", plot = att_freq_plot, width = 8,
       height = 5)
```

## 7.2 Attendance frequency by deprivation

```{r}
# Attendance frequency by deprivation
freq_by_deprivation <- mul_att_long %>%
  filter(!is.na(Deprivation)) %>%
  group_by(Deprivation, attendance_freq) %>%
  summarise(n_patients = sum(n_patients, na.rm = TRUE), .groups = "drop") %>%
  mutate(Deprivation = factor(Deprivation, levels = 1:5,
           labels = c("1 (most deprived)", "2", "3", "4", "5 (least deprived)")))

att_freq_dep_plot <- ggplot(freq_by_deprivation, 
                             aes(x = Deprivation, y = n_patients, 
                                 fill = attendance_freq)) +
  geom_col(position = "fill") +
  scale_fill_manual(values = phs_colours(c("phs-blue", "phs-teal", "phs-green",
                                             "phs-graphite", "phs-rust"))) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Deprivation Quintile", y = "Share of Patients", fill = "Attendances",
       title = "Attendance Frequency Composition by Deprivation Quintile") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

att_freq_dep_plot

ggsave("figures/eda/attendance_frequency_by_deprivation.pdf", 
       plot = att_freq_dep_plot, width = 8, height = 5)
```



```{r}
# Attendance frequency by age
att_freq_by_age <- mul_att_long %>%
  filter(!is.na(Age)) %>%
  mutate(Age = droplevels(Age)) %>%
  group_by(Age, attendance_freq) %>%
  summarise(n_patients = sum(n_patients, na.rm = TRUE), .groups = "drop")

att_freq_age_plot <- ggplot(att_freq_by_age, 
                             aes(x = Age, y = n_patients, 
                                 fill = attendance_freq)) +
  geom_col(position = "fill") +
  scale_fill_manual(values = phs_colours(c("phs-blue", "phs-teal", "phs-green",
                                             "phs-graphite", "phs-rust"))) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Age Group", y = "Share of Patients", fill = "Attendances",
       title = "Attendance Frequency Composition by Age Group") +
  theme_phs() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))

att_freq_age_plot

ggsave("figures/eda/attendance_frequency_by_age.pdf", plot = att_freq_age_plot,
       width = 9, height = 5)
```


```{r}
frequent_attender_by_deprivation <- freq_by_deprivation %>%
  group_by(Deprivation) %>%
  mutate(total_patients = sum(n_patients, na.rm = TRUE)) %>%
  filter(attendance_freq == "5+") %>%
  transmute(Deprivation, pct_five_plus = 100 * n_patients / total_patients)

frequent_attender_by_deprivation
```

```{r}
frequent_attender_by_age <- att_freq_by_age %>%
  group_by(Age) %>%
  mutate(total_patients = sum(n_patients, na.rm = TRUE)) %>%
  filter(attendance_freq == "5+") %>%
  transmute(Age, pct_five_plus = 100 * n_patients / total_patients)

frequent_attender_by_age
```


# 8. Timing of attendance

This section examines when A&E attendances occur by day, hour, out of hours status and weekday/weekend classification.

```{r}
summarise_columns(when_data)
head(when_data)
```

## 8.1 Attendance patterns by day and hour

```{r}
# Aggregate attendance volume by day and hour of the day
when_heatmap_data <- when_data %>%
  group_by(Day, Hour) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), 
            .groups = "drop") %>%
  mutate(Day = factor(Day, 
                      levels = c("Monday", "Tuesday", "Wednesday", 
                                 "Thursday", "Friday", "Saturday", "Sunday")))

# Display attendance intensity
when_heatmap_plt <- ggplot(when_heatmap_data, aes(x = Hour, y = Day, 
                                                  fill = total)) +
  geom_tile(colour = "white") +
  scale_fill_gradient(low = "white", high = phs_colours("phs-blue"), 
                      labels = scales::comma, name = "Total\nAttendances") +
  labs(x = "Hour of Day", y = NULL, 
       title = "A&E Attendances by Day of Week and Hour") +
  theme_phs() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5))

when_heatmap_plt

ggsave("figures/eda/when_heatmap.pdf", plot = when_heatmap_plt, 
       width = 12, height = 6)
```

## 8.2 Out of hours attendance share over time

```{r}
# Out of hours attendance trend over the years
oohours_trend <- when_data %>%
  group_by(m_date) %>%
  summarise(
    total = sum(NumberOfAttendances, na.rm = TRUE),
    out_of_hours = sum(NumberOfAttendances[InOut == "Out of Hours"], 
                       na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(oohours_pct = out_of_hours / total * 100)

# Plot monthly variation 
oohours_trend_plt <- ggplot(oohours_trend, aes(x = m_date, y = oohours_pct)) +
  geom_line(linewidth = 0.8, colour = phs_colours("phs-rust")) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1, 
              colour = phs_colours("phs-blue")) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(x = NULL, y = "Out of Hours Share (%)",
       title = "Out of Hours A&E Attendance Share Over Time") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

oohours_trend_plt

ggsave("figures/eda/oohours_trend.pdf", plot = oohours_trend_plt, 
       width = 9, height = 5)
```

## 8.3 Out of hours attendance share by Health Board

```{r}
# Out of hours by health board
oohours_by_hb <- when_data %>%
  group_by(HBT) %>%
  summarise(
    total = sum(NumberOfAttendances, na.rm = TRUE),
    out_of_hours = sum(NumberOfAttendances[InOut == "Out of Hours"], 
                       na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(oohours_pct = out_of_hours / total * 100)

oohours_by_hb <- oohours_by_hb %>%
  left_join(hb_names, by = "HBT")

oohours_by_hb_plt <- ggplot(oohours_by_hb, 
                            aes(x = fct_reorder(HBName, oohours_pct), 
                                y = oohours_pct)) +
  geom_col(fill = phs_colours("phs-rust")) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(x = NULL, y = "Out of Hours Share (%)",
       title = "Out of Hours A&E Attendance Share by Health Board") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

oohours_by_hb_plt

ggsave("figures/eda/oohours_by_hb.pdf", plot = oohours_by_hb_plt, 
       width = 8, height = 6)
```

## 8.4 Average daily attendance on weekdays and weekends

```{r}
# Average daily attendance: weekday vs weekend 
weekend_trend <- when_data %>%
  group_by(m_date, Week) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(n_days = if_else(Week == "Weekend", 2, 5),
         avg_daily = total / n_days)

weekend_trend_plt <- ggplot(weekend_trend, 
                             aes(x = m_date, y = avg_daily, colour = Week)) +
  geom_line(linewidth = 1) +
  scale_colour_discrete_phs(palette = "main") +
  scale_y_continuous(labels = scales::comma) +
  labs(x = NULL, y = "Average Daily Attendances", colour = NULL,
       title = "A&E Attendances: Weekday vs Weekend (per day)") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

weekend_trend_plt

ggsave("figures/eda/weekend_vs_weekday_daily.pdf", 
       plot = weekend_trend_plt, width = 9, height = 5)
```

```{r}
# Weekend attendance share over time
weekend_share <- when_data %>%
  group_by(m_date) %>%
  summarise(
    total = sum(NumberOfAttendances, na.rm = TRUE),
    weekend = sum(NumberOfAttendances[Week == "Weekend"], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(weekend_pct = 100 * weekend / total)

weekend_share_plt <- ggplot(weekend_share, 
                             aes(x = m_date, y = weekend_pct)) +
  geom_line(linewidth = 0.8, colour = phs_colours("phs-rust")) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1,
              colour = phs_colours("phs-blue")) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(x = NULL, y = "Weekend Share (%)",
       title = "Weekend Share of A&E Attendances Over Time") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

weekend_share_plt

ggsave("figures/eda/weekend_share_trend.pdf",
       plot = weekend_share_plt, width = 9, height = 5)
```


# 9. SIMD data 

This section examines the SIMD classification and indicator datasets used to represent area level deprivation and geographic access.

## 9.1 SIMD main

```{r}
summarise_columns(simd_main)
head(simd_main)
```

## 9.2 SIMD indicators

```{r}
summarise_columns(simd_ind)
head(simd_ind)
```

Income_rate and Employment_rate are proportions (0-1 scale), not percentages.

## 9.3 Aggregated SIMD by HB and quintile

Full SIMD aggregation validation is documented in build_risk_data.R. Summary checks retained here.

```{r}
simd_hb %>%
  group_by(Quintile) %>%
  summarise(across(c(Income_rate, Employment_rate, crime_rate, ALCOHOL, DRUG),
                    \(x) mean(x, na.rm = TRUE)))
```

```{r}
# Correlation matrix 
cor_matrix <- simd_hb %>%
  select(Income_rate, Employment_rate, EMERG, crime_rate, 
         ALCOHOL, DRUG, drive_GP) %>%
  cor(use = "pairwise.complete.obs")
round(cor_matrix, 2)
```

Income_rate, Employment_rate, EMERG, DRUG, and ALCOHOL are all correlated 
at r = 0.85–0.99, reflecting the occurrence of multiple deprivation dimensions 
in the same areas. 
drive_GP is the only indicator negatively correlated with the others 
(r = -0.19 to -0.38), capturing a distinct geographic access dimension. 

Employment_rate (r = 0.99 with Income_rate) was excluded from all specifications.


## 9.4 Deprivation indicators by quintile

```{r}
# Income deprivation rate by quintile and HB
income_by_quintile_plt <- simd_hb %>%
  mutate(Quintile = factor(Quintile)) %>%
  ggplot(aes(x = Quintile, y = Income_rate)) +
  geom_boxplot(fill = phs_colours("phs-blue"), alpha = 0.7) +
  labs(x = "Deprivation Quintile (1 = most deprived)",
       y = "Income Deprivation Rate",
       title = "Income Deprivation Rate by SIMD Quintile") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

income_by_quintile_plt

ggsave("figures/eda/income_rate_by_quintile.pdf", 
       plot = income_by_quintile_plt, width = 7, height = 5)
```


```{r}
# Key indicators faceted by quintile — shows how each domain varies
# across the deprivation gradient
simd_long <- simd_hb %>%
  select(HB, Quintile, Income_rate, EMERG, crime_rate, drive_GP) %>%
  pivot_longer(cols = c(Income_rate, EMERG, crime_rate, drive_GP),
               names_to = "indicator", values_to = "value") %>%
  mutate(Quintile = factor(Quintile))

indicators_by_quintile_plt <- ggplot(simd_long, 
                                      aes(x = Quintile, y = value)) +
  geom_boxplot(fill = phs_colours("phs-teal"), alpha = 0.7) +
  facet_wrap(~ indicator, scales = "free_y") +
  labs(x = "Deprivation Quintile",
       y = "Indicator value",
       title = "SIMD Domain Indicators by Deprivation Quintile") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

indicators_by_quintile_plt

ggsave("figures/eda/simd_indicators_by_quintile.pdf",
       plot = indicators_by_quintile_plt, width = 10, height = 7)
```

The boxplots show within-quintile variation across health boards — even 
within the same national quintile, boards differ in their domain-specific 
deprivation profiles. 

## 9.5 Geographic variation in deprivation

```{r}
# Income deprivation rate by health board
simd_hb_income_plt <- simd_hb_overall %>%
  left_join(hb_names, by = c("HB" = "HBT")) %>%
  ggplot(aes(x = fct_reorder(HBName, Income_rate), y = Income_rate)) +
  geom_col(fill = phs_colours("phs-purple")) +
  coord_flip() +
  labs(x = NULL, y = "Population weighted Income Deprivation Rate",
       title = "Income Deprivation Rate by Health Board") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

simd_hb_income_plt

ggsave("figures/eda/income_rate_by_hb.pdf", 
       plot = simd_hb_income_plt, width = 8, height = 6)
```

```{r}
# drive_GP by health board — shows the rurality dimension
simd_hb_gp_plt <- simd_hb_overall %>%
  left_join(hb_names, by = c("HB" = "HBT")) %>%
  ggplot(aes(x = fct_reorder(HBName, drive_GP), y = drive_GP)) +
  geom_col(fill = phs_colours("phs-rust")) +
  coord_flip() +
  labs(x = NULL, y = "Average Drive Time to GP (minutes)",
       title = "GP Access by Health Board") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

simd_hb_gp_plt

ggsave("figures/eda/drive_gp_by_hb.pdf",
       plot = simd_hb_gp_plt, width = 8, height = 6)
```

We can see the contrast between income deprivation (highest in urban boards like 
Greater Glasgow and Clyde) and GP drive time (highest in rural/island 
boards like Highland, Shetland and Western Isles).

## 9.6 Island-board quintile gaps

```{r}
# Island board quintile gaps
missing_combos <- expand_grid(
  HB = sort(unique(month_demo_data$HBT)), 
  Quintile = 1:5
) %>% 
  anti_join(simd_hb, by = c("HB", "Quintile")) %>%
  left_join(hb_names, by = c("HB" = "HBT"))

missing_combos
```

Three island health boards (Orkney, Shetland, Western Isles) have no data 
zones in certain national SIMD quintiles.  These 
combinations are excluded from the risk model (around 5% of rows), documented in 
build_risk_data.R.







