---
title: "PHS EDA"
author: "Natalia Montalvo Cabornero"
date: "2026-08-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)

# All package loading, helper functions, data import and cleaning happens in
# prep.R file 
source("R/prep.R")
source("R/risk_model_data.R")
source("R/discharge_model_data.R")

# Make sure the figure output folders exist
dir.create("figures/eda", recursive = TRUE, showWarnings = FALSE)
```

# 1. Monthly demographics data set

```{r}
summarise_columns(month_demo_data)
head(month_demo_data)
```

```{r}
# Monthly demographics year trends
month_trend_data <- month_demo_data %>%
  group_by(year, m_num) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop")

# Plot attendances by month over years
m_demo_trend_plt <- ggplot(month_trend_data, 
                           aes(x = m_num, y = total, color = year, 
                               group = year)) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(x = "Month", 
       y = "Total Attendances",
       title = "A&E Attendances by Month Across Years") +
  theme_minimal(base_size = 14) +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

m_demo_trend_plt

ggsave("figures/eda/m_demo_trend_by_year.pdf", plot = m_demo_trend_plt,
       width = 8, height = 5)

```

```{r}
overall_monthly_trend <- month_demo_data %>%
  group_by(m_date) %>%
  summarise(
    total_attendances = sum(NumberOfAttendances, na.rm = TRUE),
    .groups = "drop"
  )

overall_monthly_trend_plt <- ggplot(
  overall_monthly_trend,
  aes(x = m_date, y = total_attendances)
) +
  geom_line(linewidth = 0.8) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    x = NULL,
    y = "Total attendances",
    title = "Monthly A&E attendances over time"
  ) +
  theme_phs()

overall_monthly_trend_plt

ggsave("figures/eda/overall_monthly_trend.pdf", plot = overall_monthly_trend_plt,
       width = 9, height = 5)
```

```{r}
# Divide data by sex
sex_data <- month_demo_data %>%
  filter(Sex %in% c("Male", "Female")) %>%
  group_by(Sex) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop")

# Plot sex differences in attendance
sex_plt <- ggplot(sex_data, aes(x = Sex, y = total, fill = Sex)) +
  geom_col() +
  scale_fill_discrete_phs(palette = "main") +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Attendances by Sex", 
       x = "Sex", 
       y = "Total Attendances") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

sex_plt

ggsave("figures/eda/attendances_by_sex.pdf", plot = sex_plt, width = 6, 
       height = 5)
```

```{r}
# Monthly demographics year trends by sex
month_trend_sex_data <- month_demo_data %>%
  filter(Sex %in% c("Male", "Female")) %>%
  group_by(year, m_num, Sex) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop")

month_trend_sex_plt <- ggplot(month_trend_sex_data, 
                              aes(x = m_num, y = total,
                                  color = year, group = year)) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  facet_wrap(~ Sex, ncol = 2) +
  labs(x = "Month", y = "Total Attendances",
       title = "A&E Attendances by Month, Year, and Sex") +
  theme_phs() +
  theme(axis.text.x = element_text(size = 11, angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))

month_trend_sex_plt

ggsave("figures/eda/montrend_by_year_sex.pdf", plot = month_trend_sex_plt,
       width = 10, height = 5)
```

```{r}
# Divide attendances by deprivation index
deprivation_data <- month_demo_data %>%
  filter(!is.na(Deprivation)) %>%
  group_by(Deprivation) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop") %>%
  mutate(Deprivation = factor(Deprivation, levels = 1:5,
           labels = c("1 (most deprived)", "2", "3", "4", "5 (least deprived)")))

# Plotting attendances by deprivation quintile
deprivation_plt <- ggplot(deprivation_data, aes(x = Deprivation, y = total)) +
  geom_col(fill = phs_colours("phs-graphite")) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Deprivation Quintile", y = "Total Attendances",
       title = "Attendances by Deprivation Quintile") +
  theme_phs() +
  theme(axis.text.x = element_text(size = 11, angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))

deprivation_plt

ggsave("figures/eda/attendances_by_deprivation.pdf", plot = deprivation_plt, width = 8, height = 5)
```

```{r}
# Divide attendances by deprivation index and type of department
deprivation_dept_data <- month_demo_data %>%
  filter(!is.na(Deprivation)) %>%
  group_by(Deprivation, DepartmentType) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop") %>%
  mutate(Deprivation = factor(Deprivation, levels = 1:5,
           labels = c("1 (most deprived)", "2", "3", "4", "5 (least deprived)")))

# Plotting attendances by deprivation quintile and department type
deprivation_dept_plt <- ggplot(deprivation_dept_data, 
                               aes(x = Deprivation, y = total)) +
  geom_col(fill = phs_colours("phs-graphite")) +
  facet_wrap(~ DepartmentType, scales = "free_y") +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Deprivation Quintile", 
       y = "Total Attendances",
       title = "Attendances by Deprivation Quintile and Department Type") +
  theme_phs() +
  theme(axis.text.x = element_text(size = 10, angle = 30, hjust = 1),
        plot.title = element_text(hjust = 0.5))

deprivation_dept_plt

ggsave("figures/eda/deprivation_by_dept.pdf", plot = deprivation_dept_plt, 
       width = 10, height = 6)
```

```{r}
# Divide attendances by deprivation index and sex
deprivation_sex_data <- month_demo_data %>%
  filter(!is.na(Deprivation), Sex %in% c("Male", "Female")) %>%
  group_by(Deprivation, Sex) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop") %>%
  mutate(Deprivation = factor(Deprivation, levels = 1:5,
           labels = c("1 (most deprived)", "2", "3", "4", "5 (least deprived)")))

# Plotting attendances by deprivation quintile and sex
deprivation_sex_plt <- ggplot(deprivation_sex_data, aes(x = Deprivation, y = total, fill = Sex)) +
  geom_col(position = "dodge") +
  scale_fill_discrete_phs(palette = "main") +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Deprivation Quintile", y = "Total Attendances",
       title = "Attendances by Deprivation Quintile and Sex") +
  theme_phs() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))

deprivation_sex_plt

ggsave("figures/eda/deprivation_by_sex.pdf", plot = deprivation_sex_plt, width = 8, height = 5)
```


```{r}
# Data by age groups
age_data <- month_demo_data %>%
  filter(!is.na(Age), Age != "") %>%
  group_by(Age) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop")

# Plot attendances by age groups
age_plt <- ggplot(age_data, aes(x = fct_reorder(Age, total), y = total)) +
  geom_col(fill = phs_colours("phs-teal")) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Age Group", y = "Total Attendances", title = "Attendances by Age Group") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

age_plt

ggsave("figures/eda/attendances_by_age.pdf", plot = age_plt, width = 8, height = 5)
```

```{r}
age_sex_attendance <- month_demo_data %>%
  filter(
    !is.na(Age),
    Sex %in% c("Male", "Female")
  ) %>%
  group_by(Age, Sex) %>%
  summarise(
    total = sum(NumberOfAttendances, na.rm = TRUE),
    .groups = "drop"
  )

age_sex_attendance_plt <- ggplot(
  age_sex_attendance,
  aes(x = Age, y = total, fill = Sex)
) +
  geom_col(position = "dodge") +
  scale_fill_discrete_phs(palette = "main") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    x = "Age group",
    y = "Total attendances",
    title = "A&E attendances by age group and sex"
  ) +
  theme_phs() +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )

age_sex_attendance_plt
```

```{r}
# Deprivation converted to a labelled factor here 
age_deprivation_data <- month_demo_data %>%
  filter(
    !is.na(Age),
    !is.na(Deprivation)
  ) %>%
  group_by(Age, Deprivation) %>%
  summarise(
    total = sum(NumberOfAttendances, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(Deprivation = factor(Deprivation, levels = 1:5,
           labels = c("1 (most deprived)", "2", "3", "4", "5 (least deprived)")))

age_deprivation_plt <- ggplot(age_deprivation_data, 
                              aes(x = Age, y = total, fill = Deprivation)) +
  geom_col(position = "fill") +
  scale_fill_manual(values = phs_colours(c("phs-purple", "phs-magenta", 
                                           "phs-blue", "phs-green", 
                                           "phs-graphite"))) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Age group",
       y = "Share of attendances",
       fill = "SIMD quintile",
       title = "Deprivation composition of attendances by age group") +
  theme_phs() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

age_deprivation_plt

ggsave("figures/eda/age_deprivation_composition.pdf", plot = age_deprivation_plt,
       width = 9, height = 5)
```

```{r}
# Distribution of attendances
attendances_hist_plt <- ggplot(month_demo_data, 
                               aes(x = NumberOfAttendances)) +
  geom_histogram(bins = 50, fill = phs_colours("phs-purple"), alpha = 0.85) +
  labs(x = "Number of Attendances", 
       y = "Count",
       title = "Distribution of Number Of Attendances") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

attendances_hist_plt

ggsave("figures/eda/attendances_histogram.pdf", plot = attendances_hist_plt, width = 8, height = 5)

# Distribution of attendances log scale
attendances_hist_log_plt <- ggplot(month_demo_data, 
                                   aes(x = NumberOfAttendances)) +
  geom_histogram(bins = 50, fill = phs_colours("phs-purple"), alpha = 0.85) +
  scale_x_log10() +
  labs(x = "Number of Attendances (log scale)", 
       y = "Count",
       title = "Distribution of Number Of Attendances (log scale)") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

attendances_hist_log_plt

ggsave("figures/eda/attendances_histogram_log.pdf", plot = attendances_hist_log_plt, width = 8, height = 5)
```

# 2. Activity data

## 2.1 Weekly data

This data set is only Type 1 (ED). 
32 treatment locations and 14 Health boards.

`week_activity_data` is already cleaned by `prep.R` (`WeekEndingDate` parsed into `w_date`/`year`/`m_num`/`week_num`).

```{r}
summarise_columns(week_activity_data)
head(week_activity_data)
tail(week_activity_data)
```

## 2.2 Monthly data

`month_activity_data` is already cleaned by `prep.R`: "All" rows have been
dropped (kept only "Unplanned" and "New planned", to avoid double counting if summed), so `AttendanceCategory` below will only show those two values.

```{r}
summarise_columns(month_activity_data)
head(month_activity_data)
tail(month_activity_data)
```

# 3. Treatment location data

`trt_loc` is already loaded by `prep.R`

```{r}
summarise_columns(trt_loc)
head(trt_loc)
```

This join attaches location names/status onto the weekly dataset for display

```{r}
# Add location info to week activity data set
week_activity_data <- week_activity_data %>%
  left_join(trt_loc, by = c("TreatmentLocation" = "TreatmentLocationCode"))

week_activity_data %>%
  filter(is.na(TreatmentLocationName)) %>%
  distinct(TreatmentLocation)

week_activity_data %>%
  distinct(TreatmentLocation, TreatmentLocationName, HB, CurrentDepartmentType, Status)
```

```{r}
# How many weeks per location in data set
location_nweeks <- week_activity_data %>%
  filter(AttendanceCategory == "All") %>%
  count(TreatmentLocation, name = "n_weeks_reported") %>%
  left_join(trt_loc, by = c("TreatmentLocation" = "TreatmentLocationCode")) %>%
  select(TreatmentLocation, TreatmentLocationName, HB, Status, n_weeks_reported) %>%
  arrange(n_weeks_reported)

location_nweeks
```

- G516H: West Glasgow Ambulatory Care Hospital S08000031 Closed (15 weeks)
- G306H: New Victoria Hospital S08000031Ope n (13 weeks)

```{r}
# Average attendance per location
location_attendance <- week_activity_data %>%
  filter(AttendanceCategory == "All") %>%
  group_by(TreatmentLocation) %>%
  summarise(
    n_weeks_reported = n(),
    total_attendance = sum(NumberOfAttendancesEpisode, na.rm = TRUE),
    avg_weekly_attendance = mean(NumberOfAttendancesEpisode, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(trt_loc, by = c("TreatmentLocation" = "TreatmentLocationCode")) %>%
  select(TreatmentLocation, TreatmentLocationName, HB, Status,
         n_weeks_reported, total_attendance, avg_weekly_attendance) %>%
  arrange(desc(total_attendance))

location_attendance

# Plot locations by number of attendances
location_attendance_plt <- ggplot(location_attendance, 
                                    aes(x = fct_reorder(TreatmentLocationName, avg_weekly_attendance), 
                                        y = avg_weekly_attendance)) +
  geom_col(fill = phs_colours("phs-blue")) +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  labs(x = NULL, y = "Average Weekly Attendances",
       title = "NHS Scotland A&E Locations by Average Weekly Attendance") +
  theme_phs() +
  theme(
    axis.text.y = element_text(size = 9),
    panel.grid.major.y = element_blank()
  )

location_attendance_plt

ggsave("figures/eda/location_avg_attendance.pdf", plot = location_attendance_plt, 
       width = 8, height = 10)
```

```{r}
# Average attendance per location per year
location_year_attendance <- week_activity_data %>%
  filter(AttendanceCategory == "All") %>%
  group_by(year, TreatmentLocation) %>%
  summarise(
    n_weeks_reported = n(),
    total_attendance = sum(NumberOfAttendancesEpisode, na.rm = TRUE),
    avg_weekly_attendance = mean(NumberOfAttendancesEpisode, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(trt_loc, by = c("TreatmentLocation" = "TreatmentLocationCode")) %>%
  select(year, TreatmentLocation, TreatmentLocationName, HB, Status,
         n_weeks_reported, total_attendance, avg_weekly_attendance) %>%
  arrange(desc(total_attendance))

location_year_attendance

# Top 10 locations with most attendances
top10_locations <- location_year_attendance %>%
  group_by(TreatmentLocationName) %>%
  summarise(overall_avg = mean(avg_weekly_attendance, na.rm = TRUE),
            .groups = "drop") %>%
  slice_max(overall_avg, n = 10) %>%
  arrange(desc(overall_avg)) %>%
  pull(TreatmentLocationName)

top10_locations

# Convert the top-10 names to their location codes 
top10_codes <- trt_loc %>%
  filter(TreatmentLocationName %in% top10_locations) %>%
  pull(TreatmentLocationCode)

# Plot locations by number of attendances
top10_trend_overlay_plt <- location_year_attendance %>%
  filter(TreatmentLocationName %in% top10_locations) %>%
  mutate(TreatmentLocationName = factor(TreatmentLocationName, levels = top10_locations)) %>%
  ggplot(aes(x = year, 
             y = avg_weekly_attendance, 
             colour = TreatmentLocationName, group = TreatmentLocationName)) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Year", y = "Average Weekly Attendances", colour = "Location",
       title = "Top 10 A&E Locations: Attendance Trend by Year") +
  theme_phs() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
        legend.position = "right",
        legend.text = element_text(size = 9),
        legend.title = element_text(size = 11),
        legend.key.size = unit(0.4, "cm"))

top10_trend_overlay_plt

ggsave("figures/eda/top10_location_trend_overlay.pdf", plot = top10_trend_overlay_plt, width = 11, height = 6)

```



# 4. Population data

```{r}
head(pop_data)
summarise_columns(pop_data)
```

```{r}
distinct(pop_data, Sex)

n_distinct(pop_data$DataZone)   # should be 6976, per the dataset description
# Confirm
pop_data %>% filter(DataZoneQF != "" | !is.na(DataZoneQF)) %>% 
  distinct(DataZone, DataZoneQF)
n_distinct(simd_ind$Data_Zone)             # compare against what you already validated

setdiff(unique(pop_data$DataZone), unique(simd_main$DataZone))
setdiff(unique(simd_main$DataZone), unique(pop_data$DataZone))
```


```{r}
national_total_new <- pop_data %>%
  mutate(Sex = trimws(Sex)) %>%
  filter(Year == 2020, Sex != "All", DataZone != "S92000003")

sum(national_total_new$AllAges, na.rm = TRUE)

# Compare against:
sum(pop_hb_density$population)      # your NRS mid-2024 estimate
sum(simd_hb$quintile_population)    # your SIMD 2020 population
```






```{r}
head(pop_hb_age_sex)
head(pop_hb_density)
head(simd_hb)
```

Sex: person (all), female and male
Area type: Scotland (all), council area, HB

```{r}
population_pyramid_data <- pop_hb_age_sex %>%
  group_by(Age, Sex) %>%
  summarise(population = sum(population, na.rm = TRUE), .groups = "drop") %>%
  mutate(plot_population = if_else(Sex == "Male", -population, population))

population_pyramid_plt <- ggplot(
  population_pyramid_data,
  aes(x = Age, y = plot_population, fill = Sex)
) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(
    labels = ~ scales::comma(abs(.x))
  ) +
  scale_fill_discrete_phs(palette = "main") +
  labs(
    x = "Age group",
    y = "Population",
    title = "Population distribution by age and sex"
  ) +
  theme_phs()

population_pyramid_plt

ggsave("figures/eda/population_pyramid.pdf", plot = population_pyramid_plt,
       width = 8, height = 5)
```

```{r}
# Population by age group and sex, nationally
pop_age_sex_plt <- pop_hb_age_sex %>%
  group_by(Age, Sex) %>%
  summarise(population = sum(population, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = Age, y = population, fill = Sex)) +
  geom_col(position = "dodge") +
  scale_fill_discrete_phs(palette = "main") +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Age Group", y = "Population", title = "Scotland Population by Age Group and Sex") +
  theme_phs()

pop_age_sex_plt

ggsave("figures/eda/population_by_age_sex.pdf", plot = pop_age_sex_plt, width = 8, height = 5)
```

```{r}
# Population density by health board
pop_density_plt <- ggplot(pop_hb_density, aes(x = fct_reorder(HB, density), y = density)) +
  geom_col(fill = phs_colours("phs-liberty")) +
  coord_flip() +
  labs(x = "Health Board", y = "Population Density (per km\u00b2)",
       title = "Population Density by Health Board") +
  theme_phs()

pop_density_plt

ggsave("figures/eda/population_density_by_hb.pdf", plot = pop_density_plt, width = 8, height = 6)
```

```{r}
attendance_age_sex <- month_demo_data %>%
  filter(!is.na(Age), Sex %in% c("Male", "Female")) %>%
  group_by(Age, Sex) %>%
  summarise(total_attendances = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop")

population_age_sex <- pop_hb_age_sex %>%
  group_by(Age, Sex) %>%
  summarise(population = sum(population, na.rm = TRUE), .groups = "drop")

age_sex_rates <- attendance_age_sex %>%
  left_join(population_age_sex, by = c("Age", "Sex")) %>%
  mutate(attendance_rate = total_attendances / population * 1000)

age_sex_rate_plt <- ggplot(age_sex_rates, aes(x = Age, y = attendance_rate, colour = Sex, group = Sex)) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_colour_discrete_phs(palette = "main") +
  labs(x = "Age group", y = "Attendances per 1,000 population",
       title = "Population-adjusted A&E attendance rates by age and sex") +
  theme_phs() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

age_sex_rate_plt

ggsave("figures/eda/age_sex_attendance_rate.pdf", plot = age_sex_rate_plt,
       width = 8, height = 5)
```

```{r}
hb_attendance <- month_demo_data %>%
  group_by(HBT) %>%
  summarise(total_attendance = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop")

hb_attendance_rates <- hb_attendance %>%
  left_join(
    pop_hb_density %>%
      select(HB, population, density),
    by = c("HBT" = "HB")
  ) %>%
  mutate(
    rate_per_1000 =
      total_attendance / population * 1000
  )

hb_rate_plt <- ggplot(
  hb_attendance_rates,
  aes(
    x = fct_reorder(HBT, rate_per_1000),
    y = rate_per_1000
  )
) +
  geom_col(fill = phs_colours("phs-blue")) +
  coord_flip() +
  labs(
    x = NULL,
    y = "Attendances per 1,000 population",
    title = "Population-adjusted attendance rates by health board"
  ) +
  theme_phs()

hb_rate_plt

ggsave("figures/eda/hb_attendance_rate.pdf", plot = hb_rate_plt, width = 8, height = 6)
```


```{r}
density_rate_plt <- ggplot(
  hb_attendance_rates,
  aes(
    x = density,
    y = rate_per_1000,
    label = HBT
  )
) +
  geom_point(size = 3, colour = phs_colours("phs-blue")) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    colour = phs_colours("phs-rust")
  ) +
  labs(
    x = "Population density per km²",
    y = "Attendances per 1,000 population",
    title = "Population density and A&E attendance rates"
  ) +
  theme_phs()

density_rate_plt

ggsave("figures/eda/density_attendance_rate.pdf", plot = density_rate_plt,
       width = 8, height = 5)
```


# 5. Referrals data

`referrals_data` is already cleaned by `prep.R` (QF dropped, date parsed,
`Referral` recoded so blanks/":" show as "Not Known") - used directly here.

```{r}
summarise_columns(referrals_data)
head(referrals_data)
```

```{r}
# Attendance by referral type in general
referral_types <- referrals_data %>%
  mutate(Referral = case_when(
    is.na(Referral) | trimws(Referral) %in% c("", ":") ~ "Not Known",
    TRUE ~ Referral
  )) %>%
  group_by(Referral) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop")

referrals_type_plt <- ggplot(referral_types, 
                              aes(x = fct_reorder(Referral, total), y = total)) +
  geom_col(fill = phs_colours("phs-graphite")) +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  labs(x = NULL, y = "Total Attendances",
       title = "Attendances by Referral Source") +
  theme_phs()

referrals_type_plt

ggsave("figures/eda/attendances_by_referral.pdf", plot = referrals_type_plt, width = 8, height = 5)
```

```{r}
# Attendance by referral type and age
referral_age_data <- referrals_data %>%
  filter(!is.na(Age), Age != "") %>%
  mutate(Referral = case_when(
    is.na(Referral) | trimws(Referral) %in% c("", ":") ~ "Not Known",
    TRUE ~ Referral
  )) %>%
  group_by(Referral, Age) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop")

referral_age_plt <- ggplot(referral_age_data, 
                            aes(x = Age, y = total, fill = Referral)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = phs_colours(c("phs-purple", "phs-magenta", 
                                           "phs-blue", "phs-green",
                                           "phs-graphite", "phs-teal", 
                                           "phs-liberty", "phs-rust"))) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Age Group", y = "Total Attendances", fill = "Referral Source",
       title = "Attendances by Age Group and Referral Source") +
  theme_phs() +
  theme(legend.text = element_text(size = 9), legend.key.size = unit(0.4, "cm"))

referral_age_plt

ggsave("figures/eda/referral_by_age.pdf", plot = referral_age_plt, width = 11, 
       height = 6)
```

```{r}
referral_top10_data <- referrals_data %>%
  filter(!is.na(Referral), Referral != "", 
         TreatmentLocation %in% top10_codes) %>%
  left_join(trt_loc, by = c("TreatmentLocation" = "TreatmentLocationCode")) %>%
  group_by(TreatmentLocationName, Referral) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop")

referral_top10_pct <- referral_top10_data %>%
  group_by(TreatmentLocationName) %>%
  mutate(pct = total / sum(total) * 100) %>%
  ungroup()

referral_table_plt <- ggplot(referral_top10_pct, 
                                aes(x = Referral, y = TreatmentLocationName, fill = pct)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = round(pct, 0)), size = 3, colour = "black") +
  scale_fill_gradient(low = "white", high = phs_colours("phs-blue"), name = "% of\nattendances") +
  labs(x = "Referral Source", y = NULL,
       title = "Referral Source Percentages by Top 10 A&E Locations") +
  theme_phs() +
  theme(
    axis.text.x = element_text(size = 9, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 9),
    panel.grid = element_blank()
  )

referral_table_plt

ggsave("figures/eda/referral_table_top10.pdf", plot = referral_table_plt, width = 10, height = 7)
```

Unir monthly activity with referrals:



# 6. Discharges

```{r}
summarise_columns(discharges_data)
head(discharges_data)
```


```{r}
# Attendance by discharge type in general
discharge_types <- discharges_data %>%
  mutate(Discharge = case_when(
    is.na(Discharge) | trimws(Discharge) %in% c("", ":") ~ "Not Known",
    TRUE ~ Discharge
  )) %>%
  group_by(Discharge) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop")

discharge_type_plt <- ggplot(discharge_types, 
                              aes(x = fct_reorder(Discharge, total), y = total)) +
  geom_col(fill = phs_colours("phs-graphite")) +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  labs(x = NULL, y = "Total Attendances",
       title = "Attendances by Discharge Type") +
  theme_phs()

discharge_type_plt

ggsave("figures/eda/attendances_by_discharge.pdf", plot = discharge_type_plt, width = 8, height = 5)
```


```{r}
discharge_age_data <- discharges_data %>%
  filter(!is.na(Age), Age != "") %>%
  mutate(Discharge = case_when(
    is.na(Discharge) | trimws(Discharge) %in% c("", ":") ~ "Not Known",
    TRUE ~ Discharge
  )) %>%
  group_by(Discharge, Age) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop")

discharge_age_plt <- ggplot(discharge_age_data, 
                            aes(x = Age, y = total, fill = Discharge)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = phs_colours(c("phs-purple", "phs-magenta", 
                                           "phs-blue", "phs-green",
                                           "phs-teal"))) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Age Group", y = "Total Attendances", fill = "Discharge Type",
       title = "Attendances by Age Group and Discharge Type") +
  theme_phs() +
  theme(legend.text = element_text(size = 9), legend.key.size = unit(0.4, "cm"))

discharge_age_plt

ggsave("figures/eda/discharge_by_age.pdf", plot = discharge_age_plt, width = 11, 
       height = 6)
```

```{r}
# Admission rate over time
admission_rate_trend <- discharges_data %>%
  group_by(m_date) %>%
  summarise(
    total_attendances = sum(NumberOfAttendances, na.rm = TRUE),
    admitted = sum(NumberOfAttendances[Discharge == "Admission to same Hospital"], na.rm = TRUE),  # adjust label
    .groups = "drop"
  ) %>%
  mutate(admission_rate = admitted / total_attendances * 100)

admission_rate_plt <- ggplot(admission_rate_trend, aes(x = m_date, y = admission_rate)) +
  geom_line(linewidth = 0.8, colour = phs_colours("phs-rust")) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1, colour = phs_colours("phs-blue")) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(x = NULL, y = "Admission Rate (%)",
       title = "A&E Admission Rate Over Time") +
  theme_phs()

admission_rate_plt

ggsave("figures/eda/admission_rate_trend.pdf", plot = admission_rate_plt, width = 9, height = 5)
```


# 7. Multiple attendances

```{r}
summarise_columns(mul_att_data)
head(mul_att_data)
tail(mul_att_data)
```

```{r}
mul_att_long <- mul_att_data %>%
  pivot_longer(cols = OneAttendance:FivePlusAttendances,
               names_to = "attendance_freq", values_to = "n_patients") %>%
  mutate(attendance_freq = factor(attendance_freq,
           levels = c("OneAttendance", "TwoAttendances", "ThreeAttendances",
                       "FourAttendances", "FivePlusAttendances"),
           labels = c("1", "2", "3", "4", "5+")))

freq_totals <- mul_att_long %>%
  group_by(attendance_freq) %>%
  summarise(n_patients = sum(n_patients, na.rm = TRUE), .groups = "drop")

att_freq_plot <- ggplot(freq_totals, aes(x = attendance_freq, y = n_patients)) +
  geom_col(fill = phs_colours("phs-blue")) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Number of Attendances (12-month period)", y = "Number of Patients",
       title = "Distribution of A&E Attendance Frequency per Patient") +
  theme_phs()

att_freq_plot 
```

```{r}
# attendance frequency by deprivation
freq_by_deprivation <- mul_att_long %>%
  filter(!is.na(Deprivation)) %>%
  group_by(Deprivation, attendance_freq) %>%
  summarise(n_patients = sum(n_patients, na.rm = TRUE), .groups = "drop") %>%
  group_by(Deprivation) %>%
  mutate(pct = n_patients / sum(n_patients) * 100) %>%
  mutate(Deprivation = factor(Deprivation, levels = 1:5,
           labels = c("1 (most deprived)", "2", "3", "4", "5 (least deprived)")))

att_freq_dep_plot <- ggplot(freq_by_deprivation, aes(x = Deprivation, y = pct, fill = attendance_freq)) +
  geom_col(position = "fill") +
  scale_fill_manual(values = phs_colours(c("phs-blue", "phs-teal", "phs-green",
                                             "phs-graphite", "phs-rust"))) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(x = "Deprivation Quintile", y = "Share of Patients", fill = "Attendances",
       title = "Attendance Frequency Composition by Deprivation Quintile") +
  theme_phs()

att_freq_dep_plot
```
Most derpvied attend more times

```{r}
att_freq_by_age <- mul_att_long %>%
  filter(!is.na(Age), Age != "") %>%
  group_by(Age, attendance_freq) %>%
  summarise(n_patients = sum(n_patients, na.rm = TRUE), .groups = "drop") %>%
  group_by(Age) %>%
  mutate(pct = n_patients / sum(n_patients) * 100)

att_freq_age_plot <- ggplot(att_freq_by_age, aes(x = Age, y = pct, fill = attendance_freq)) +
  geom_col(position = "fill") +
  scale_fill_manual(values = phs_colours(c("phs-blue", "phs-teal", "phs-green",
                                             "phs-graphite", "phs-rust"))) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(x = "Age Group", y = "Share of Patients", fill = "Attendances",
       title = "Attendance Frequency Composition by Age Group") +
  theme_phs() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

att_freq_age_plot 
```

Similar accross age groups

```{r}
# Self-referral share vs. frequent-attendance share, by health board


# mul_att_data covers a single 12-month window ending April 2026 - referrals
# filtered to roughly the same period so the comparison is like-for-like
freq_rate_by_hb <- mul_att_long %>%
  group_by(HBT) %>%
  summarise(
    total_patients = sum(n_patients, na.rm = TRUE),
    freq_patients = sum(n_patients[attendance_freq == "5+"], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(freq_pct = freq_patients / total_patients * 100)

referral_mix_by_hb <- referrals_data %>%
  filter(m_date >= as.Date("2025-05-01"), m_date <= as.Date("2026-04-01")) %>%
  group_by(HBT) %>%
  summarise(
    total = sum(NumberOfAttendances, na.rm = TRUE),
    self_referral = sum(NumberOfAttendances[Referral == "Self Referral"], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(self_referral_pct = self_referral / total * 100)

ecological_compare <- freq_rate_by_hb %>%
  left_join(referral_mix_by_hb, by = "HBT")

self_referral_freq_plt <- ggplot(ecological_compare,
                                  aes(x = self_referral_pct, y = freq_pct, label = HBT)) +
  geom_point(size = 3, colour = phs_colours("phs-blue")) +
  geom_smooth(method = "lm", se = TRUE, colour = phs_colours("phs-rust")) +
  labs(x = "Self-Referral Share (%)", y = "Frequent Attender Share (%)",
       title = "Self-Referral Rate vs. Frequent Attendance Rate, by Health Board",
       caption = "Ecological (area-level) comparison across 14 health boards - not a patient-level relationship") +
  theme_phs() +
  theme(plot.caption = element_text(size = 8, hjust = 0, colour = "grey40"))

self_referral_freq_plt

ggsave("figures/eda/self_referral_vs_frequent_attendance.pdf",
       plot = self_referral_freq_plt, width = 8, height = 5)
```

not a strong relationship, wide CI and poitns scattered. 

# 8. When attendance
```{r}
summarise_columns(when_data)
head(when_data)
```


```{r}
when_heatmap_data <- when_data %>%
  group_by(Day, Hour) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop") %>%
  mutate(Day = factor(Day, levels = c("Monday", "Tuesday", "Wednesday", "Thursday",
                                        "Friday", "Saturday", "Sunday")))

when_heatmap_plt <- ggplot(when_heatmap_data, aes(x = Hour, y = Day, fill = total)) +
  geom_tile(colour = "white") +
  scale_fill_gradient(low = "white", high = phs_colours("phs-blue"), labels = scales::comma,
                       name = "Total\nAttendances") +
  labs(x = "Hour of Day", y = NULL, title = "A&E Attendances by Day of Week and Hour") +
  theme_phs() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
        panel.grid = element_blank())

when_heatmap_plt

ggsave("figures/eda/when_heatmap.pdf", plot = when_heatmap_plt, width = 12, height = 6)
```


```{r}
# out of hours attendance trend over the years
oohours_trend <- when_data %>%
  group_by(m_date) %>%
  summarise(
    total = sum(NumberOfAttendances, na.rm = TRUE),
    out_of_hours = sum(NumberOfAttendances[InOut == "Out of Hours"], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(oohours_pct = out_of_hours / total * 100)

oohours_trend_plt <- ggplot(oohours_trend, aes(x = m_date, y = oohours_pct)) +
  geom_line(linewidth = 0.8, colour = phs_colours("phs-rust")) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1, colour = phs_colours("phs-blue")) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(x = NULL, y = "Out-of-Hours Share (%)",
       title = "Out-of-Hours A&E Attendance Share Over Time") +
  theme_phs()

oohours_trend_plt

ggsave("figures/eda/oohours_trend.pdf", plot = oohours_trend_plt, width = 9, height = 5)
```

```{r}
# out of hours by health board
oohours_by_hb <- when_data %>%
  group_by(HBT) %>%
  summarise(
    total = sum(NumberOfAttendances, na.rm = TRUE),
    out_of_hours = sum(NumberOfAttendances[InOut == "Out of Hours"], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(oohours_pct = out_of_hours / total * 100)

oohours_by_hb_plt <- ggplot(oohours_by_hb, aes(x = fct_reorder(HBT, oohours_pct), y = oohours_pct)) +
  geom_col(fill = phs_colours("phs-rust")) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +
  labs(x = NULL, y = "Out-of-Hours Share (%)",
       title = "Out-of-Hours A&E Attendance Share by Health Board") +
  theme_phs()

oohours_by_hb_plt

ggsave("figures/eda/oohours_by_hb.pdf", plot = oohours_by_hb_plt, width = 8, height = 6)
```


```{r}
weekend_trend <- when_data %>%
  group_by(m_date, Week) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop")

ggplot(weekend_trend, aes(x = m_date, y = total, colour = Week)) +
  geom_line(linewidth = 1) +
  scale_colour_discrete_phs(palette = "main") +
  scale_y_continuous(labels = scales::comma) +
  labs(x = NULL, y = "Total Attendances", colour = NULL,
       title = "A&E Attendances Over Time: Weekday vs. Weekend") +
  theme_phs()
```

# 9. SIMD data 

## 9.1 SIMD main

```{r}
summarise_columns(simd_main)
```

## 9.2 SIMD indicators

Income_rate and Employment_rate are proportions (0-1 scale), not percentages

```{r}
summarise_columns(simd_ind)
```

## 9.3 Checks

 Check after agregagating by HB both simd datasets
```{r}
simd_hb %>%
  group_by(Quintile) %>%
  summarise(across(c(Income_rate, Employment_rate, crime_rate, ALCOHOL, DRUG),
                    \(x) mean(x, na.rm = TRUE)))
```

