---
title: "PHS Risk Analysis"
author: "Natalia Montalvo Cabornero"
date: "2026-08-14"
output: html_document
---

```{r setup, include=FALSE}

source("R/risk_model_data.R")

# Make sure the figure output folders exist
dir.create("figures/risk_model", recursive = TRUE, showWarnings = FALSE)
```

# 1. Data recap

The combined risk model uses `combined_model_data`, built from:

- **Outcome**: `NumberOfAttendances` — monthly count of A&E attendances per
  demographic group
  
- **Offset**: `population` — transforms the model from predicting raw counts 
  (which are driven by group size as much as by risk) into predicting a 
  population adjusted rate. Mathematically, including log(population) as an 
  offset with a coefficient fixed at 1 is equivalent to modelling 
  NumberOfAttendances / population as the outcome, but done within the count
  model framework so the correct variance structure is preserved.
  
- **Demographic predictors**: `Age` (6 groups), `Sex`, `Deprivation` (SIMD
  quintile 1-5)
  
- **SIMD domain indicators** (HB x Quintile level): `Income_rate`,
  `Employment_rate`, `crime_rate`, `ALCOHOL`, `DRUG`, `EMERG`, `drive_GP`
  
- **Operational context** (HB x Month level): `oohours_pct`, `weekend_pct`,
  referral mix percentages
  
- **Structural terms**: 
    - `s(t)`: long-term trend, non-linear trajectory of attendance over the 
              2018–2026 study period
    - `s(m_num, bs = "cc")`: seasonal smooth, captures the repeating within-year
              seasonal cycle (e.g. winter peaks) separately from
              the long-term trend.
    - `s(HBT, bs = "re")`: health board random intercept, accounts for baseline 
              differences in attendance rates across Scotland's 14 health boards
  
- **Distribution**: negative binomial distribution was chosen over Poisson 
  because A&E attendance counts show overdispersion. The variance exceeds the 
  mean, as confirmed by the right-skewed distribution observed in the EDA

```{r}
dim(combined_model_data)  # 79505 x 30
summary(combined_model_data$NumberOfAttendances)
summary(combined_model_data$population)

# Confirm no zero/NA populations that would break the log offset
sum(is.na(combined_model_data$population) | combined_model_data$population <= 0)

summarise_columns(combined_model_data )
```

5% missing from the merged data with the simd indicators and population from those 
missing quintile info

```{r}
# Filter to rows with valid population (excludes island-board quintile gaps)
risk_data <- combined_model_data %>%
  filter(!is.na(population), population > 0) %>%
  mutate(
    HBT = factor(HBT),
    Age = factor(Age),
    Sex = factor(Sex),
    Deprivation = factor(Deprivation)
  )
```


# 2. Modelling section 

## 2.1 Baseline model 

```{r}
baseline_model2 <- gam(NumberOfAttendances ~ 
                        Age + Sex + Deprivation +
                        s(t, k = 25) +
                        s(m_num, bs = "cc", k = 12) +
                        s(HBT, bs = "re") +
                        offset(log(population)),
                      family = nb(),
                      data = risk_data,
                      method = "REML")

summary(baseline_model2)
```

```{r}
baseline_model <- bam(NumberOfAttendances ~ 
                        Age + Sex + Deprivation +
                        s(t, k = 30) +
                        s(m_num, bs = "cc", k = 12) +
                        s(HBT, bs = "re") +
                        offset(log(population)),
                      family = nb(),
                      data = risk_data,
                      method = "REML")

summary(baseline_model)

# Baseline AIC
baseline_aic <- AIC(baseline_model)
baseline_aic

# Create table for AIC comparison
aic_log <- tribble(
  ~model, ~aic, ~delta_aic, ~note,
  "baseline", baseline_aic, 0, "Age + Sex + Deprivation + smooths"
)
```

```{r}
par(mfrow = c(2, 2))
gam.check(baseline_model)
par(mfrow = c(1, 1))
```

```{r}
model_time2 <- system.time({
  baseline_model <- gam(
    NumberOfAttendances ~
      Age + Sex + Deprivation +
      s(t, k = 25) +
      s(m_num, bs = "cc", k = 12) +
      s(HBT, bs = "re") +
      offset(log(population)),
    family = nb(),
    data = risk_data,
    method = "REML"
  )
})

model_time
#  user  system elapsed 
# 97.67    4.22  105.31 


model_time1 <- system.time({
  baseline_model <- bam(
    NumberOfAttendances ~
      Age + Sex + Deprivation +
      s(t, k = 25) +
      s(m_num, bs = "cc", k = 12) +
      s(HBT, bs = "re") +
      offset(log(population)),
    family = nb(),
    data = risk_data,
    method = "REML"
  )
})

model_time2
  #  user  system elapsed 
  # 15.74    1.75   18.08 
```



```{r}
baseline_coefs <- summary(baseline_model)$p.table
baseline_coefs <- as.data.frame(baseline_coefs)
baseline_coefs$rate_ratio <- exp(baseline_coefs$Estimate)
baseline_coefs$rr_lower <- exp(baseline_coefs$Estimate - 1.96 * baseline_coefs$`Std. Error`)
baseline_coefs$rr_upper <- exp(baseline_coefs$Estimate + 1.96 * baseline_coefs$`Std. Error`)
baseline_coefs
```

```{r}
# Smooth term plots
plot(baseline_model, pages = 1, shade = TRUE)
```

### Check NB over Poisson

```{r}
poisson_model <- bam(
  NumberOfAttendances ~
    Age + Sex + Deprivation +
    s(t, k = 30) + s(m_num, bs = "cc", k = 12) + s(HBT, bs = "re") +
    offset(log(population)),
  family = poisson(),
  data = risk_data,
  method = "REML"
)

# Dispersion ratio: residual deviance / residual df
# Should be close to 1 for Poisson; >>1 indicates overdispersion
dispersion_ratio <- sum(residuals(poisson_model, type = "deviance")^2) / poisson_model$df.residual
dispersion_ratio

# Compare AIC
AIC(poisson_model)
AIC(baseline_model)

# The difference should be large and favour the NB model (lower AIC)
AIC(poisson_model) - AIC(baseline_model)
```

## 2.2 Baseline + income

```{r}
model_income <- bam(NumberOfAttendances ~ 
                        Age + Sex + Deprivation + Income_rate +
                        s(t, k = 30) +
                        s(m_num, bs = "cc", k = 12) +
                        s(HBT, bs = "re") +
                        offset(log(population)),
                      family = nb(),
                      data = risk_data,
                      method = "REML")

summary(model_income)
```

Change of direction of coeffs -- maybe due to circularity?
option maybe replace "deprivation" by only indicators

```{r}
aic_log <- aic_log %>% add_row(
  model = "+ Income_rate", aic = AIC(model_income),
  delta_aic = AIC(model_income) - baseline_aic,
  note = "Income deprivation rate (HB x Quintile level)"
)
aic_log
```

## 2.3 Baseline + EMERG
```{r}
model_emerg <- bam(
  NumberOfAttendances ~
    Age + Sex + Deprivation + EMERG +
    s(t, k = 30) + s(m_num, bs = "cc", k = 12) + s(HBT, bs = "re") +
    offset(log(population)),
  family = nb(),
  data = risk_data,
  method = "REML"
)

summary(model_emerg)
```

similar issue: dep index is positive now

```{r}
aic_log <- aic_log %>% add_row(
  model = "+ EMERG", aic = AIC(model_emerg),
  delta_aic = AIC(model_emerg) - baseline_aic,
  note = "Historical emergency stays ratio (SIMD health domain)"
)

aic_log
```

## 2.4 Baseline + External (when data)

```{r}
model_external <- bam(
  NumberOfAttendances ~
    Age + Sex + Deprivation + oohours_pct + weekend_pct +
    s(t, k = 30) + s(m_num, bs = "cc", k = 12) + s(HBT, bs = "re") +
    offset(log(population)),
  family = nb(),
  data = risk_data,
  method = "REML"
)

summary(model_external)
```

- HB with higher oohours share have lower attendacne rates --> areas with better 
primary care alternatives divert some demand?
- HB with higher weekend share have higher rates --> maybe areas where weekend 
alternatives GP or pharmacies are less available

```{r}
aic_log <- aic_log %>% add_row(
  model = "+ oohours_pct + weekend_pct",
  aic = AIC(model_external),
  delta_aic = AIC(model_external) - baseline_aic,
  note = "External operational context only (no SIMD indicators)"
)

aic_log
```

some improvement

## 2.5 Baseline no Deprivation, only indicators

```{r}
model_indicators <- bam(
  NumberOfAttendances ~
    Age + Sex + Income_rate + EMERG + crime_rate + drive_GP +
    oohours_pct + weekend_pct +
    s(t, k = 30) + s(m_num, bs = "cc", k = 12) + s(HBT, bs = "re") +
    offset(log(population)),
  family = nb(),
  data = risk_data,
  method = "REML"
)

summary(model_indicators)
```

- Income_rate: strongest effect by far. Higher income deprivation --> higher 
attendance rate.
- EMERG: higher historical emergency burden --> higher attendance. 
- crime_rate: negative (higher crime --> lower attendance?). maybe crime rate 
correlates with income deprivation
- drive_GP: negative, longer drive to GP --> lower attendance. could reflect 
rurality: remote areas have long GP drive times and long A&E drive times
- oohours_pct and weekend_pct — same direction and magnitude as model_external.


```{r}
aic_log <- aic_log %>% add_row(
  model = "indicators(no Deprivation)",
  aic = AIC(model_indicators),
  delta_aic = AIC(model_indicators) - baseline_aic,
  note = "SIMD domain indicators replace Deprivation quintile"
)
aic_log
```

```{r}

```


## 2.6 Baseline no Deprivation, only indicators (no crime)

```{r}
model_indicators_nocrime <- bam(
  NumberOfAttendances ~
    Age + Sex + Income_rate + EMERG + drive_GP +
    oohours_pct + weekend_pct +
    s(t, k = 30) + s(m_num, bs = "cc", k = 12) + s(HBT, bs = "re") +
    offset(log(population)),
  family = nb(),
  data = risk_data,
  method = "REML"
)

summary(model_indicators_nocrime)
```

```{r}
aic_log <- aic_log %>% add_row(
  model = "indicators(no Deprivation/crime rate)",
  aic = AIC(model_indicators_nocrime),
  delta_aic = AIC(model_indicators_nocrime) - baseline_aic,
  note = "SIMD domain indicators replace Deprivation quintile (no crime rate)"
)
aic_log
```

worse AIC, so crime does add info

maybe among areas with similar income deprivation levels, those with higher 
crime rates tend to have slightly lower attendance, possibly because high-crime 
areas in Scotland tend to be dense urban locations with better access to both 
A&E and alternative services, while lower-crime areas at the same income level 
tend to be more dispersed/rural with fewer options.


## 2.7 interactions to explore

- Age * sex

```{r}
model_interaction_age_sex <- bam(
  NumberOfAttendances ~
    Age * Sex + Deprivation + oohours_pct + weekend_pct +
    s(t, k = 30) + s(m_num, bs = "cc", k = 12) + s(HBT, bs = "re") +
    offset(log(population)),
  family = nb(),
  data = risk_data,
  method = "REML"
)

summary(model_interaction_age_sex)
AIC(model_interaction_age_sex) - baseline_aic
```


- Age * deprivation

```{r}
model_interaction_age_dep <- bam(
  NumberOfAttendances ~
    Age * Deprivation + Sex + oohours_pct + weekend_pct +
    s(t, k = 30) + s(m_num, bs = "cc", k = 12) + s(HBT, bs = "re") +
    offset(log(population)),
  family = nb(),
  data = risk_data,
  method = "REML"
)

summary(model_interaction_age_dep)
AIC(model_interaction_age_dep) - baseline_aic
```

```{r}
model_both_interactions <- bam(
  NumberOfAttendances ~
    Age * Sex + Age * Deprivation + oohours_pct + weekend_pct +
    s(t, k = 30) + s(m_num, bs = "cc", k = 12) + s(HBT, bs = "re") +
    offset(log(population)),
  family = nb(),
  data = risk_data,
  method = "REML"
)
summary(model_both_interactions)

AIC(model_both_interactions) - baseline_aic

aic_log <- aic_log %>% add_row(
  model = "Age*Sex + Age*Dep + external",
  aic = AIC(model_both_interactions),
  delta_aic = AIC(model_both_interactions) - baseline_aic,
  note = "Both interactions + operational context"
)
```

```{r}
gam.check(model_both_interactions)
```

