---
title: "PHS Risk Analysis"
author: "Natalia Montalvo Cabornero"
date: "2026-08-14"
output: html_document
---

```{r setup, include=FALSE}

source("R/risk_model_data.R")

# Make sure the figure output folders exist
dir.create("figures/risk_model", recursive = TRUE, showWarnings = FALSE)
```

# 1. Data recap

The combined risk model uses `combined_model_data`, built from:

- **Outcome**: `NumberOfAttendances` — monthly count of A&E attendances per
  demographic group
  
- **Offset**: `population` — exact HB x Age x Sex x Deprivation joint
  population, each attendance year matched to its own population year; 2025/2026
  capped to 2024, the most recent available)
  
- **Demographic predictors**: `Age` (6 groups), `Sex`, `Deprivation` (SIMD
  quintile 1-5)
  
- **SIMD domain indicators** (HB x Quintile level): `Income_rate`,
  `Employment_rate`, `crime_rate`, `ALCOHOL`, `DRUG`, `EMERG`, `drive_GP`

- **Operational context** (HB x Month level): `oohours_pct`, `weekend_pct`,
  referral mix percentages

- **Grain**: Month x HBT x Age x Sex x Deprivation

```{r}
dim(combined_model_data)  # 79505 x 30
summary(combined_model_data$NumberOfAttendances)
summary(combined_model_data$population)

# Confirm no zero/NA populations that would break the log offset
sum(is.na(combined_model_data$population) | combined_model_data$population <= 0)

summarise_columns(combined_model_data)
```

5% missing from the merged data with the simd indicators and population from those 
missing quintile info

```{r}
# Filter to rows with valid population (excludes island-board quintile gaps)
risk_data <- combined_model_data %>%
  filter(!is.na(population), population > 0) %>%
  rename(pct_ref_self = `pct_ref_Self Referral`) %>%
  mutate(
    HBT = factor(HBT),
    Age = factor(Age),
    Sex = factor(Sex),
    Deprivation = factor(Deprivation))

cat("Rows after filtering:", nrow(risk_data), "\n")
cat("Rows excluded:", nrow(combined_model_data) - nrow(risk_data), "\n")
```


# 2. Baseline model

```{r}
baseline_model <- bam(NumberOfAttendances ~ 
                        Age + Sex + Deprivation +
                        s(t, k = 35) +
                        s(m_num, bs = "cc", k = 12) +
                        s(HBT, bs = "re") +
                        offset(log(population)),
                      family = nb(),
                      data = risk_data,
                      method = "ML")

summary(baseline_model)
```

## 2.1 Diagnostics

```{r}
pdf("figures/risk_model/gam_check_baseline.pdf", width = 10, height = 8)
par(mfrow = c(2, 2))
gam.check(baseline_model)
par(mfrow = c(1, 1))
dev.off()
```

The k-index test is non-significant for both smooth terms


## 2.2 Overdispersion check

```{r}
poisson_model <- bam(NumberOfAttendances ~
                       Age + Sex + Deprivation +
                       s(t, k = 35) + 
                       s(m_num, bs = "cc", k = 12) + 
                       s(HBT, bs = "re") +
                       offset(log(population)),
                     family = poisson(),
                     data = risk_data,
                     method = "ML")

# Dispersion ratio: should be ~1 for Poisson; >>1 indicates overdispersion
dispersion_ratio <- sum(residuals(poisson_model, type = "deviance")^2) /
  poisson_model$df.residual
cat("Dispersion ratio:", round(dispersion_ratio, 2), "\n")

# AIC comparison
cat("Poisson AIC:", AIC(poisson_model), "\n")
cat("NB AIC:", AIC(baseline_model), "\n")

rm(poisson_model)
```
A Poisson model was initially considered as the natural choice for count
data, but diagnostic checks revealed substantial overdispersion.

## 2.3 Coefficient interpretation

```{r}
# Exponentiate parametric coefficients to get rate ratios
baseline_coefs <- summary(baseline_model)$p.table
baseline_coefs <- as.data.frame(baseline_coefs)
baseline_coefs$rate_ratio <- exp(baseline_coefs$Estimate)
baseline_coefs$rr_lower <- exp(baseline_coefs$Estimate - 1.96 * baseline_coefs$`Std. Error`)
baseline_coefs$rr_upper <- exp(baseline_coefs$Estimate + 1.96 * baseline_coefs$`Std. Error`)
baseline_coefs
```

Cross-check against EDA findings:

- Deprivation show a monotonic gradient (1 --> 5)
- Age75 plus show higher risk than other age bands
- Males show modestly higher risk than females

## 2.4 Smooth term plots

```{r}
# Smooth term plots
pdf("figures/risk_model/smooth_terms_baseline.pdf", width = 10, height = 4)
plot(baseline_model, pages = 1, shade = TRUE)
dev.off()
```

## 2.5 Record baseline AIC

```{r}
baseline_aic <- AIC(baseline_model)

aic_log <- tribble(
  ~model, ~aic, ~delta_aic, ~note,
  "baseline", baseline_aic, 0, "Age + Sex + Deprivation + smooths")

aic_log
```


# 3. Adding contextual predictors

Starting from the baseline, we add one predictor at a time. For each
addition: check whether AIC improves, whether the new coefficient is
significant, and whether existing coefficients change notably.

SIMD domain indicators are components of the SIMD score that determines
the Deprivation quintile, so including them alongside `Deprivation` risks
collinearity, tested below.

## 3.1 Baseline + Income rate

```{r}
model_income <- bam(NumberOfAttendances ~ 
                        Age + Sex + Deprivation + Income_rate +
                        s(t, k = 35) +
                        s(m_num, bs = "cc", k = 12) +
                        s(HBT, bs = "re") +
                        offset(log(population)),
                      family = nb(),
                      data = risk_data,
                      method = "ML")

summary(model_income)

aic_log <- aic_log %>% add_row(
  model = "+ Income_rate", aic = AIC(model_income),
  delta_aic = AIC(model_income) - baseline_aic,
  note = "Income deprivation rate (HB x Quintile level)")

aic_log
```

Deprivation coefficients changed direction, a sign of collinearity. Income 
deprivation is a direct component of the SIMD quintile, so cannot use it 
alongside Deprivation.

## 3.2 Baseline + EMERG
```{r}
model_emerg <- bam(NumberOfAttendances ~
                     Age + Sex + Deprivation + EMERG +
                     s(t, k = 35) + 
                     s(m_num, bs = "cc", k = 12) + 
                     s(HBT, bs = "re") +
                     offset(log(population)),
                   family = nb(),
                   data = risk_data,
                  method = "ML")

summary(model_emerg)

aic_log <- aic_log %>% add_row(
  model = "+ EMERG", aic = AIC(model_emerg),
  delta_aic = AIC(model_emerg) - baseline_aic,
  note = "Historical emergency stays ratio (SIMD health domain)")

aic_log
```

Similar issue, so cannot use with Deprivation either.

## 3.3 Baseline + External operational context (when data)

```{r}
model_external <- bam(NumberOfAttendances ~
                        Age + Sex + Deprivation + oohours_pct + weekend_pct +
                        s(t, k = 35) + 
                        s(m_num, bs = "cc", k = 12) + 
                        s(HBT, bs = "re") +
                        offset(log(population)),
                      family = nb(),
                      data = risk_data,
                      method = "ML")

summary(model_external)

aic_log <- aic_log %>% add_row(
  model = "+ oohours_pct + weekend_pct",
  aic = AIC(model_external),
  delta_aic = AIC(model_external) - baseline_aic,
  note = "External operational context only (no SIMD indicators)")

aic_log
```

Deprivation coefficients stayed perfectly monotonic, these predictors are
external to SIMD and both significant:

- `oohours_pct` negative: HB with higher out-of-hours share have lower
  attendance rates --> areas with better out-of-hours primary care
  alternatives may divert some demand away from A&E
- `weekend_pct` positive: HB with higher weekend share have higher rates, 
  could reflect areas where weekend alternatives (GP, pharmacy) are less
  available
  
Out-of-hours and weekend attendance shares were included as HB level operational
context variables, capturing differences in the timing structure of A&E demand. 
Because these measures are derived from attendance data, they may be partially 
endogenous to the outcome-generating process. However, they may also reflect
broader structural differences between areas, including the availability and 
use of alternative urgent-care services outside standard working hours. They 
were therefore interpreted as contextual indicators rather than individual level
causal predictors.

## 3.4 Domain indicators replacing Deprivation

As shown in the EDA (Section 9.3), SIMD domain indicators are highly 
correlated.

```{r}
model_indicators <- bam(NumberOfAttendances ~
                          Age + Sex + Income_rate + EMERG + crime_rate + 
                          drive_GP + oohours_pct + weekend_pct +
                          s(t, k = 35) + 
                          s(m_num, bs = "cc", k = 12) + 
                          s(HBT, bs = "re") +
                          offset(log(population)),
                        family = nb(),
                        data = risk_data,
                        method = "ML")

summary(model_indicators)

aic_log <- aic_log %>% add_row(
  model = "indicators (no Deprivation)",
  aic = AIC(model_indicators),
  delta_aic = AIC(model_indicators) - baseline_aic,
  note = "SIMD domain indicators replace Deprivation quintile")

aic_log
```

All predictors significant:
- Income_rate is the strongest driver. 
- EMERG shows a small positive association, indicating areas with higher 
  historical emergency burden continue to generate higher A&E demand.
- crime_rate shows a negative sign, among areas with similar income deprivation 
  levels, those with higher crime rates tend to have slightly lower attendance,
  possibly because high-crime areas in Scotland tend to be dense urban
  locations with better access to both A&E and alternative services.
- drive_GP negative, same rurality reasoning: remote areas have long GP 
  drive times but also long A&E drive times.
  
  
```{r}
# Testing whether additional indicators add predictive value
model_indicators_full <- bam(NumberOfAttendances ~
                          Age + Sex + Income_rate + EMERG + crime_rate + 
                          ALCOHOL + DRUG + drive_GP + 
                          oohours_pct + weekend_pct +
                          s(t, k = 35) + 
                          s(m_num, bs = "cc", k = 12) + 
                          s(HBT, bs = "re") +
                          offset(log(population)),
                        family = nb(),
                        data = risk_data,
                        method = "ML")

summary(model_indicators_full)

aic_log <- aic_log %>% add_row(
  model = "indicators + alcohol/drug",
  aic = AIC(model_indicators_full),
  delta_aic = AIC(model_indicators_full) - baseline_aic,
  note = "SIMD domain indicators + ALCOHOL and DRUG")

aic_log
```

AIC improved by a further 1703 points, but coefficient instability 
consistent with the high correlations seen above:

- Income_rate more than doubled (1.50 --> 3.33), inflated by the correlated 
  negative ALCOHOL/DRUG effects.
- ALCOHOL and DRUG both negative, same suppression pattern as crime_rate: at a 
  given income deprivation level, areas with higher alcohol/drug burden tend to 
  be dense urban areas with better service access.
- crime_rate shrank and became barely significant (p = 0.002).
  
The simpler specification was retained, its coefficients are individually 
interpretable and the key finding (income deprivation dominates) is unaffected. 

```{r}
# Sensitivity check: removing crime_rate
model_indicators_nocrime <- bam(NumberOfAttendances ~
                                  Age + Sex + Income_rate + EMERG + drive_GP +
                                  oohours_pct + weekend_pct +
                                  s(t, k = 35) + 
                                  s(m_num, bs = "cc", k = 12) + 
                                  s(HBT, bs = "re") +
                                  offset(log(population)),
                                family = nb(),
                                data = risk_data,
                                method = "ML")

summary(model_indicators_nocrime)

aic_log <- aic_log %>% add_row(
  model = "indicators (no crime)",
  aic = AIC(model_indicators_nocrime),
  delta_aic = AIC(model_indicators_nocrime) - baseline_aic,
  note = "Domain indicators without crime_rate"
)

aic_log
```

AIC worsened by around 222 points when crime_rate was removed, confirming it
contributes genuine predictive value despite the counter intuitive sign.

## 3.5 Referral mix

Note: referral mix is partly a consequence of attendance volume (if
attendance surges, the proportion arriving via different routes may shift),
so any significant coefficient should be interpreted as associational,
not causal.

```{r}
# Self referral 
model_self_ref <- bam(NumberOfAttendances ~
                        Age + Sex + Deprivation + pct_ref_self +
                        s(t, k = 35) + 
                        s(m_num, bs = "cc", k = 12) + 
                        s(HBT, bs = "re") +
                        offset(log(population)),
                      family = nb(),
                      data = risk_data,
                      method = "ML")

summary(model_self_ref)

aic_log <- aic_log %>% add_row(
  model = "+ Self Referral %",
  aic = AIC(model_self_ref),
  delta_aic = AIC(model_self_ref) - baseline_aic,
  note = "Self-referral share"
)
aic_log
```

Referral mix (self-referral share) was tested as a predictor but excluded from 
the final specification despite statistical significance, because it describes 
how attenders arrived at A&E rather than who is at risk of attending, an 
operational characteristic of the attendance itself, not a population level 
risk factor. 

# 4. Interactions

## 4.1 Age x sex

```{r}
model_interaction_age_sex <- bam(NumberOfAttendances ~
                                   Age * Sex + Deprivation + oohours_pct + 
                                   weekend_pct +
                                   s(t, k = 35) + 
                                   s(m_num, bs = "cc", k = 12) + 
                                   s(HBT, bs = "re") +
                                   offset(log(population)),
                                 family = nb(),
                                 data = risk_data,
                                 method = "ML")

summary(model_interaction_age_sex)

aic_log <- aic_log %>% add_row(
  model = "baseline + Age*Sex",
  aic = AIC(model_interaction_age_sex),
  delta_aic = AIC(model_interaction_age_sex) - baseline_aic,
  note = "Interaction: does age pattern differ by sex?"
)
aic_log
```


## 4.2 Age × Deprivation

```{r}
model_interaction_age_dep <- bam(NumberOfAttendances ~
                                   Age * Deprivation + Sex + oohours_pct + 
                                   weekend_pct +
                                   s(t, k = 35) + 
                                   s(m_num, bs = "cc", k = 12) + 
                                   s(HBT, bs = "re") +
                                   offset(log(population)),
                                 family = nb(),
                                 data = risk_data,
                                 method = "ML")

summary(model_interaction_age_dep)

aic_log <- aic_log %>% add_row(
  model = "baseline + Age*Deprivation",
  aic = AIC(model_interaction_age_dep),
  delta_aic = AIC(model_interaction_age_dep) - baseline_aic,
  note = "Interaction: does deprivation gap differ by age?"
)
aic_log
```

## 4.3 Both interactions combined
```{r}
model_both_interactions <- bam(NumberOfAttendances ~
                                 Age * Sex + Age * Deprivation + oohours_pct + 
                                 weekend_pct +
                                 s(t, k = 35) + 
                                 s(m_num, bs = "cc", k = 12) + 
                                 s(HBT, bs = "re") +
                                 offset(log(population)),
                               family = nb(),
                               data = risk_data,
                               method = "ML")

summary(model_both_interactions)

aic_log <- aic_log %>% add_row(
  model = "Age*Sex + Age*Dep + external",
  aic = AIC(model_both_interactions),
  delta_aic = AIC(model_both_interactions) - baseline_aic,
  note = "Both interactions + operational context"
)
aic_log
```


# 5. Final Model 

All model comparisons above used ML estimation, which provides a valid 
likelihood for comparing models with different fixed-effect structures. The 
final selected specifications are refitted below using REML, which gives less 
biased estimates of variance components and smoothing parameters for inference.

```{r}
# Final model with Deprivation quintile
final_risk_model_A <- bam(
  NumberOfAttendances ~
    Age * Sex + Age * Deprivation + oohours_pct + weekend_pct +
    s(t, k = 40) + s(m_num, bs = "cc", k = 12) + s(HBT, bs = "re") +
    offset(log(population)),
  family = nb(),
  data = risk_data,
  method = "REML")

summary(final_risk_model_A)
```

Model A uses k = 40 for s(t) rather than k = 35 used during model selection. 
The Age × Deprivation interaction (20 additional parameters) altered the 
residual structure, causing s(t) at k = 35 to absorb seasonal oscillation into 
the trend. Increasing to k = 40 resolved this.

```{r}
# Final model with Deprivation indicators
final_risk_model_B <- bam(
  NumberOfAttendances ~
    Age * Sex + Income_rate + EMERG + crime_rate + drive_GP +
    oohours_pct + weekend_pct +
    s(t, k = 35) + s(m_num, bs = "cc", k = 12) + s(HBT, bs = "re") +
    offset(log(population)),
  family = nb(),
  data = risk_data,
  method = "REML")

summary(final_risk_model_B)
```

## 5.1 Diagnostics

```{r}
pdf("figures/risk_model/gam_check_final_A.pdf", width = 10, height = 8)
par(mfrow = c(2, 2))
gam.check(final_risk_model_A)
par(mfrow = c(1, 1))
dev.off()
```

```{r}
pdf("figures/risk_model/gam_check_final_B.pdf", width = 10, height = 8)
par(mfrow = c(2, 2))
gam.check(final_risk_model_B)
par(mfrow = c(1, 1))
dev.off()
```

## 5.2 Smooth term plots

```{r}
pdf("figures/risk_model/smooth_terms_final_A.pdf", width = 10, height = 4)
plot(final_risk_model_A, pages = 1, shade = TRUE)
dev.off()
```

```{r}
pdf("figures/risk_model/smooth_terms_final_B.pdf", width = 10, height = 4)
plot(final_risk_model_B, pages = 1, shade = TRUE)
dev.off()
```

## 5.3 Coefficient interpretation

```{r}
final_coefs_A <- as.data.frame(summary(final_risk_model_A)$p.table)
final_coefs_A$rate_ratio <- exp(final_coefs_A$Estimate)
final_coefs_A$rr_lower <- exp(final_coefs_A$Estimate - 1.96 * final_coefs_A$`Std. Error`)
final_coefs_A$rr_upper <- exp(final_coefs_A$Estimate + 1.96 * final_coefs_A$`Std. Error`)
final_coefs_A
```

Because Model A includes Age × Sex and Age × Deprivation interactions, the 
main-effect coefficients are conditional: Age rate ratios apply to females 
in quintile 1, the Sex rate ratio applies to 18-24 year olds, and the 
Deprivation rate ratios apply to 18-24 year olds. The interaction terms 
show how these effects differ for other combinations. 

```{r}
final_coefs_B <- as.data.frame(summary(final_risk_model_B)$p.table)
final_coefs_B$rate_ratio <- exp(final_coefs_B$Estimate)
final_coefs_B$rr_lower <- exp(final_coefs_B$Estimate - 1.96 * final_coefs_B$`Std. Error`)
final_coefs_B$rr_upper <- exp(final_coefs_B$Estimate + 1.96 * final_coefs_B$`Std. Error`)
final_coefs_B
```

The rate ratios above are per 1-unit increase, but the continuous SIMD 
indicators are on different scales, so raw rate ratios are not directly 
comparable across predictors. The table below shows effects scaled to each 
predictor's observed range.

```{r}
continuous_predictors <- c("Income_rate", "EMERG", "crime_rate", "drive_GP",
                            "oohours_pct", "weekend_pct")

# Check scales first
risk_data %>%
     summarise(across(all_of(continuous_predictors),
                     list(min = ~ min(.x, na.rm = TRUE),
                          q25 = ~ quantile(.x, 0.25, na.rm = TRUE),
                          q75 = ~ quantile(.x, 0.75, na.rm = TRUE),
                          max = ~ max(.x, na.rm = TRUE))))

interpretation_increments <- c(Income_rate = 0.05, EMERG = 10, crime_rate = 100,
                               drive_GP = 5, oohours_pct = 5, weekend_pct = 5)

practical_interpretation <- tibble(
  predictor = continuous_predictors,
  estimate = final_coefs_B[continuous_predictors, "Estimate"],
  increment = interpretation_increments,
  q25 = sapply(continuous_predictors,
               \(x) quantile(risk_data[[x]], 0.25, na.rm = TRUE)),
  q75 = sapply(continuous_predictors,
               \(x) quantile(risk_data[[x]], 0.75, na.rm = TRUE))
) %>%
  mutate(
    iqr = q75 - q25,
    rr_per_increment = exp(estimate * increment),
    pct_change_increment = round(100 * (rr_per_increment - 1), 1),
    rr_per_iqr = exp(estimate * iqr),
    pct_change_iqr = round(100 * (rr_per_iqr - 1), 1)
  )

practical_interpretation
```

Key findings (scaled to observed ranges):

- **Income_rate** (0.02–0.30): a 10-percentage-point increase in area 
  income deprivation is associated with a 16% higher attendance rate. 
  Across the full observed range, the most income-deprived areas have 
  ~51% higher attendance than the least (RR = 1.51).
- **EMERG** (59–165): across the full range of emergency hospitalisation 
  ratios, attendance is ~38% higher in the most burdened areas (RR = 1.38).
- **crime_rate** (78–891): across the full range, areas with highest crime 
  have ~15% lower attendance (RR = 0.85) — the conditional urban/rural 
  effect discussed in Section 3.4.
- **drive_GP** (2.5–11.0 minutes): areas with the longest GP drive times 
  have ~32% lower attendance (RR = 0.68) — the rurality dimension.
- **oohours_pct** (42–65%): areas with the highest out-of-hours share 
  have ~33% lower attendance (RR = 0.67).
- **weekend_pct** (18–38%): areas with the highest weekend share have 
  ~22% higher attendance (RR = 1.22).



```{r}
concurvity(final_risk_model_B, full = TRUE)
```


## 5.4 Predicted rates and interaction visualisation


```{r}
# Create prediction grid
pred_grid <- expand.grid(
  Age = levels(risk_data$Age),
  Sex = levels(risk_data$Sex),
  Deprivation = levels(risk_data$Deprivation)
) %>%
  mutate(
    oohours_pct = mean(risk_data$oohours_pct, na.rm = TRUE),
    weekend_pct = mean(risk_data$weekend_pct, na.rm = TRUE),
    t = median(risk_data$t),
    m_num = 6,
    HBT = levels(risk_data$HBT)[1],
    population = 10000   # per 10,000 population
  )

pred_grid$predicted <- predict(final_risk_model_A, newdata = pred_grid,
                                type = "response")
pred_grid$rate_per_10k <- pred_grid$predicted
```


```{r}
# Deprivation gradient by age group — shows the interaction directly
dep_by_age_plt <- pred_grid %>%
  group_by(Age, Deprivation) %>%
  summarise(rate_per_10k = mean(rate_per_10k), .groups = "drop") %>%
  ggplot(aes(x = Deprivation, y = rate_per_10k, colour = Age, group = Age)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(x = "Deprivation Quintile (1 = most deprived)",
       y = "Predicted attendances per 10,000 population",
       title = "Deprivation gradient varies by age group",
       colour = "Age group") +
  theme_phs()

dep_by_age_plt

ggsave("figures/risk_model/dep_gradient_by_age.pdf", 
       plot = dep_by_age_plt, width = 8, height = 6)
```


```{r}
sex_by_age_plt <- pred_grid %>%
  group_by(Age, Sex) %>%
  summarise(rate_per_10k = mean(rate_per_10k), .groups = "drop") %>%
  ggplot(aes(x = Age, y = rate_per_10k, fill = Sex)) +
  geom_col(position = "dodge") +
  labs(x = "Age group",
       y = "Predicted attendances per 10,000 population",
       title = "Male-female gap varies by age group") +
  theme_phs()

sex_by_age_plt

ggsave("figures/risk_model/sex_gap_by_age.pdf",
       plot = sex_by_age_plt, width = 8, height = 6)
```


```{r}
heatmap_plt <- pred_grid %>%
  filter(Sex == "Female") %>%   # one sex to keep it readable
  ggplot(aes(x = Deprivation, y = Age, fill = rate_per_10k)) +
  geom_tile() +
  geom_text(aes(label = round(rate_per_10k, 0)), size = 3) +
  scale_fill_gradient(low = "white", high = phs_colours("phs-blue")) +
  labs(x = "Deprivation Quintile", y = "Age group",
       fill = "Rate per\n10,000",
       title = "Predicted A&E attendance rate (females)") +
  theme_phs()

heatmap_plt

ggsave("figures/risk_model/rate_heatmap_female.pdf",
       plot = heatmap_plt, width = 7, height = 5)
```


# 6. Supplementary analyses

## 6.1 Fixed-effect HB version

Run the final Model A specification with HBT as a fixed effect instead of 
a random effect, to extract and rank individual health board coefficients.

```{r}
# Set reference HB for this analysis only
risk_data_hb <- risk_data %>%
  mutate(HBT = relevel(HBT, ref = "S08000031"))

final_risk_model_hb <- bam(
  NumberOfAttendances ~
    Age * Sex + Age * Deprivation + oohours_pct + weekend_pct + HBT +
    s(t, k = 40) + 
    s(m_num, bs = "cc", k = 12) +
    offset(log(population)),
  family = nb(),
  data = risk_data_hb,
  method = "REML"
)

# Extract and rank HB rate ratios
hb_coefs <- as.data.frame(summary(final_risk_model_hb)$p.table)
hb_coefs$term <- rownames(hb_coefs)
hb_coefs <- hb_coefs %>%
  filter(grepl("^HBT", term)) %>%
  mutate(
    HBT = gsub("^HBT", "", term),
    rate_ratio = exp(Estimate),
    rr_lower = exp(Estimate - 1.96 * `Std. Error`),
    rr_upper = exp(Estimate + 1.96 * `Std. Error`)
  ) %>%
  left_join(hb_names, by = "HBT") %>%
  # Add reference board back at RR = 1
  bind_rows(tibble(
    HBT = "S08000031",
    HBName = "Greater Glasgow and Clyde",
    rate_ratio = 1, rr_lower = 1, rr_upper = 1
  )) %>%
  arrange(desc(rate_ratio))

hb_coefs
```


```{r}
hb_coefs_plt <- ggplot(hb_coefs, 
                        aes(x = fct_reorder(HBName, rate_ratio), 
                            y = rate_ratio)) +
  geom_point(aes(colour = rate_ratio == 1), size = 3, show.legend = FALSE) +
  scale_colour_manual(values = c("TRUE" = "grey50", 
                                  "FALSE" = phs_colours("phs-blue"))) +
  geom_errorbar(aes(ymin = rr_lower, ymax = rr_upper), width = 0.3,
                colour = phs_colours("phs-blue")) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey30") +
  coord_flip() +
  labs(x = NULL, 
       y = "Rate Ratio (vs. Greater Glasgow and Clyde)",
       title = "Health Board Attendance Rate Ratios\n
       (adjusted for Age, Sex, Deprivation, interactions)") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

hb_coefs_plt
  

ggsave("figures/risk_model/hb_rate_ratios.pdf", plot = hb_coefs_plt, 
       width = 8, height = 6)
```

```{r}
library(sf)
hb_shp <- st_read("data/SG_NHS_HealthBoards_2019/SG_NHS_HealthBoards_2019.shp")

# Join model results
hb_map_data <- hb_shapefile %>%
  left_join(hb_coefs %>% select(HBT, HBName, rate_ratio),
            by = c("HBCode" = "HBT"))

risk_map_plt <- ggplot(hb_map_data) +
  geom_sf(aes(fill = rate_ratio), colour = "grey30", linewidth = 0.5) +
  scale_fill_gradient2(low = phs_colours("phs-blue"),
                        mid = "grey95",
                        high = phs_colours("phs-rust"),
                        midpoint = 1,
                        limits = c(0.6, 1.1),
                        name = "Rate Ratio\n(vs GGC)") +
  labs(title = "A&E Attendance Rate Ratios by Health Board") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))

risk_map_plt

ggsave("figures/risk_model/hb_map_attendance_rr.pdf",
       plot = risk_map_plt, width = 8, height = 10)

```

## 6.2 GAM robustness check

Refit the final specifications with gam not bam 

### Model A - Deprivation Quintile
```{r}
# Final model with Deprivation quintile
final_risk_model_A_gam <- gam(
  NumberOfAttendances ~
    Age * Sex + Age * Deprivation + oohours_pct + weekend_pct +
    s(t, k = 40) + s(m_num, bs = "cc", k = 12) + s(HBT, bs = "re") +
    offset(log(population)),
  family = nb(),
  data = risk_data,
  method = "REML"
)

final_coefs_A_gam <- as.data.frame(summary(final_risk_model_A_gam)$p.table)

comparison_A <- data.frame(
  term = rownames(final_coefs_A_gam),
  gam_estimate = round(final_coefs_A_gam$Estimate, 5),
  bam_estimate = round(final_coefs_A$Estimate, 5),
  difference = round(final_coefs_A_gam$Estimate - final_coefs_A$Estimate, 5),
  gam_se = round(final_coefs_A_gam$`Std. Error`, 5),
  bam_se = round(final_coefs_A$`Std. Error`, 5))

comparison_A
```

```{r}
# Smooth term comparison
final_smooth_A_gam <- as.data.frame(summary(final_risk_model_A_gam)$s.table)
final_smooth_A <- as.data.frame(summary(final_risk_model_A)$s.table)

smooth_comparison_A <- data.frame(
  term = rownames(final_smooth_A_gam),
  gam_edf = round(final_smooth_A_gam$edf, 2),
  bam_edf = round(final_smooth_A$edf, 2),
  edf_diff = round(final_smooth_A_gam$edf - final_smooth_A$edf, 2)
)

smooth_comparison_A
```

### Model B - Deprivation Indicators
```{r}
# Final model with Deprivation indicators
final_risk_model_B_gam <- gam(
  NumberOfAttendances ~
    Age * Sex + Income_rate + EMERG + crime_rate + drive_GP +
    oohours_pct + weekend_pct +
    s(t, k = 35) + s(m_num, bs = "cc", k = 12) + s(HBT, bs = "re") +
    offset(log(population)),
  family = nb(),
  data = risk_data,
  method = "REML"
)

final_coefs_B_gam <- as.data.frame(summary(final_risk_model_B_gam)$p.table)

comparison_B <- data.frame(
  term = rownames(final_coefs_B_gam),
  gam_estimate = round(final_coefs_B_gam$Estimate, 5),
  bam_estimate = round(final_coefs_B$Estimate, 5),
  difference = round(final_coefs_B_gam$Estimate - final_coefs_B$Estimate, 5),
  gam_se = round(final_coefs_B_gam$`Std. Error`, 5),
  bam_se = round(final_coefs_B$`Std. Error`, 5))

comparison_B
```

```{r}
# Smooth term comparison
final_smooth_B_gam <- as.data.frame(summary(final_risk_model_B_gam)$s.table)
final_smooth_B <- as.data.frame(summary(final_risk_model_B)$s.table)

smooth_comparison_B <- data.frame(
  term = rownames(final_smooth_B_gam),
  gam_edf = round(final_smooth_B_gam$edf, 2),
  bam_edf = round(final_smooth_B$edf, 2),
  edf_diff = round(final_smooth_B_gam$edf - final_smooth_B$edf, 2)
)

smooth_comparison_B
```

## 6.3 INLA robustness check

Refit the final specification A in INLA to confirm consistency across
frequentist (GAM) and Bayesian (INLA) estimation frameworks.

```{r}
risk_data_inla <- risk_data %>%
  mutate(HBT_id = as.numeric(HBT),
         t_rounded = round(t))

inla_risk <- inla(
  NumberOfAttendances ~
    Age * Sex + Age * Deprivation + oohours_pct + weekend_pct +
    f(t_rounded, model = "rw2", constr = TRUE) +
    f(m_num, model = "rw1", cyclic = TRUE, constr = TRUE) +
    f(HBT_id, model = "iid") +
    offset(log(population)),
  family = "nbinomial",
  data = risk_data_inla,
  control.compute = list(dic = TRUE, waic = TRUE),
  control.predictor = list(compute = TRUE)
)

summary(inla_risk)

# INLA hyperparameters (dispersion, random effect variance)
inla_risk$summary.hyperpar
```


```{r}
# Compare point estimates: BAM vs INLA
final_estimates_A <- summary(final_risk_model_A)$p.table[, "Estimate"]

inla_estimates <- inla_risk$summary.fixed$mean

comparison_inla <- data.frame(
  term = rownames(summary(final_risk_model_A)$p.table),
  GAM = round(final_estimates_A, 4),
  INLA = round(inla_estimates, 4),
  difference = round(final_estimates_A - inla_estimates, 4))

comparison_inla
```

The final Model A specification was refitted using INLA as a Bayesian robustness 
check. Point estimates were consistent across frameworks: 36 of 38 coefficients 
agreed to within +-0.001, with the largest difference in the intercept (0.060) 
and weekend share (0.003).

```{r}
# INLA credible intervals vs GAM confidence intervals for key terms
inla_fixed <- inla_risk$summary.fixed

ci_comparison <- data.frame(
  term = rownames(final_coefs_A),
  gam_lower = round(final_coefs_A$Estimate - 1.96 * final_coefs_A$`Std. Error`, 4),
  gam_upper = round(final_coefs_A$Estimate + 1.96 * final_coefs_A$`Std. Error`, 4),
  inla_lower = round(inla_fixed$`0.025quant`, 4),
  inla_upper = round(inla_fixed$`0.975quant`, 4)
)

ci_comparison
```

Credible intervals (INLA) and confidence intervals (GAM) were compared across
all 38 coefficients. Intervals agree to within +- 0.001 at both bounds for 35 of 
38 terms. The largest discrepancy was for weekend_pct, where INLA's credible 
interval (0.005–0.009) was wider than GAM's confidence interval (0.009–0.012).



