---
title: "PHS Risk Analysis"
author: "Natalia Montalvo Cabornero"
date: "2026-08-14"
output: html_document
---

```{r setup, include=FALSE}

source("R/risk_model_data.R")

# Make sure the figure output folders exist
dir.create("figures/risk_model", recursive = TRUE, showWarnings = FALSE)
```

# 1. Data recap

The combined risk model uses `combined_model_data`, built from:

- **Outcome**: `NumberOfAttendances` — monthly count of A&E attendances per
  demographic group
  
- **Offset**: `population` — exact HB x Age x Sex x Deprivation joint
  population, each attendance year matched to its own population year; 2025/2026
  capped to 2024, the most recent available)
  
- **Demographic predictors**: `Age` (6 groups), `Sex`, `Deprivation` (SIMD
  quintile 1-5)
  
- **SIMD domain indicators** (HB x Quintile level): `Income_rate`,
  `Employment_rate`, `crime_rate`, `ALCOHOL`, `DRUG`, `EMERG`, `drive_GP`

- **Operational context** (HB x Month level): `oohours_pct`, `weekend_pct`,
  referral mix percentages

- **Grain**: Month x HBT x Age x Sex x Deprivation

```{r}
dim(combined_model_data)  # 79505 x 30
summary(combined_model_data$NumberOfAttendances)
summary(combined_model_data$population)

# Confirm no zero/NA populations that would break the log offset
sum(is.na(combined_model_data$population) | combined_model_data$population <= 0)

summarise_columns(combined_model_data)
```

5% missing from the merged data with the simd indicators and population from those 
missing quintile info

```{r}
# Filter to rows with valid population (excludes island-board quintile gaps)
risk_data <- combined_model_data %>%
  filter(!is.na(population), population > 0) %>%
  rename(pct_ref_self = `pct_ref_Self Referral`) %>%
  mutate(
    HBT = factor(HBT),
    Age = factor(Age),
    Sex = factor(Sex),
    Deprivation = factor(Deprivation)
  )
cat("Rows after filtering:", nrow(risk_data), "\n")
cat("Rows excluded:", nrow(combined_model_data) - nrow(risk_data), "\n")
```


# 2. Baseline model

```
NumberOfAttendances ~ Age + Sex + Deprivation +
  s(t, k = 30) + s(m_num, bs = "cc", k = 12) + s(HBT, bs = "re") +
  offset(log(population))
family = nb(), method = "REML"
```

```{r}
baseline_model <- bam(NumberOfAttendances ~ 
                        Age + Sex + Deprivation +
                        s(t, k = 35) +
                        s(m_num, bs = "cc", k = 12) +
                        s(HBT, bs = "re") +
                        offset(log(population)),
                      family = nb(),
                      data = risk_data,
                      method = "REML")

summary(baseline_model)
```

## 2.1 Diagnostics

```{r}
par(mfrow = c(2, 2))
gam.check(baseline_model)
par(mfrow = c(1, 1))
```

The k-index test is non-significant for both smooth terms


## 2.2 Overdispersion check

```{r}
poisson_model <- bam(NumberOfAttendances ~
                       Age + Sex + Deprivation +
                       s(t, k = 35) + 
                       s(m_num, bs = "cc", k = 12) + 
                       s(HBT, bs = "re") +
                       offset(log(population)),
                     family = poisson(),
                     data = risk_data,
                     method = "REML")

# Dispersion ratio: should be ~1 for Poisson; >>1 indicates overdispersion
dispersion_ratio <- sum(residuals(poisson_model, type = "deviance")^2) /
  poisson_model$df.residual
cat("Dispersion ratio:", round(dispersion_ratio, 2), "\n")

# AIC comparison
cat("Poisson AIC:", AIC(poisson_model), "\n")
cat("NB AIC:", AIC(baseline_model), "\n")

rm(poisson_model)
```
A Poisson model was initially considered as the natural choice for count
data, but diagnostic checks revealed substantial overdispersion.

## 2.3 Coefficient interpretation

```{r}
# Exponentiate parametric coefficients to get rate ratios
baseline_coefs <- summary(baseline_model)$p.table
baseline_coefs <- as.data.frame(baseline_coefs)
baseline_coefs$rate_ratio <- exp(baseline_coefs$Estimate)
baseline_coefs$rr_lower <- exp(baseline_coefs$Estimate - 1.96 * baseline_coefs$`Std. Error`)
baseline_coefs$rr_upper <- exp(baseline_coefs$Estimate + 1.96 * baseline_coefs$`Std. Error`)
baseline_coefs
```

Cross-check against EDA findings:

- Deprivation show a monotonic gradient (1 --> 5)
- Age75 plus show higher risk than other age bands
- Males show modestly higher risk than females

## 2.4 Smooth term plots

```{r}
# Smooth term plots
plot(baseline_model, pages = 1, shade = TRUE)
```

## 2.5 Record baseline AIC

```{r}
baseline_aic <- AIC(baseline_model)

aic_log <- tribble(
  ~model, ~aic, ~delta_aic, ~note,
  "baseline", baseline_aic, 0, "Age + Sex + Deprivation + smooths"
)
aic_log
```


# 3. Adding contextual predictors

Starting from the baseline, we add one predictor at a time. For each
addition: check whether AIC improves, whether the new coefficient is
significant, and whether existing coefficients change notably.

SIMD domain indicators are components of the SIMD score that determines
the Deprivation quintile, so including them alongside `Deprivation` risks
collinearity, tested below.

## 3.1 Baseline + Income rate

```{r}
model_income <- bam(NumberOfAttendances ~ 
                        Age + Sex + Deprivation + Income_rate +
                        s(t, k = 35) +
                        s(m_num, bs = "cc", k = 12) +
                        s(HBT, bs = "re") +
                        offset(log(population)),
                      family = nb(),
                      data = risk_data,
                      method = "REML")

summary(model_income)

aic_log <- aic_log %>% add_row(
  model = "+ Income_rate", aic = AIC(model_income),
  delta_aic = AIC(model_income) - baseline_aic,
  note = "Income deprivation rate (HB x Quintile level)"
)
aic_log
```

Deprivation coefficients changed direction, a sign of collinearity. Income 
deprivation is a direct component of the SIMD quintile, so cannot use it 
alongside Deprivation.

## 3.2 Baseline + EMERG
```{r}
model_emerg <- bam(NumberOfAttendances ~
                     Age + Sex + Deprivation + EMERG +
                     s(t, k = 35) + 
                     s(m_num, bs = "cc", k = 12) + 
                     s(HBT, bs = "re") +
                     offset(log(population)),
                   family = nb(),
                   data = risk_data,
                  method = "REML")

summary(model_emerg)

aic_log <- aic_log %>% add_row(
  model = "+ EMERG", aic = AIC(model_emerg),
  delta_aic = AIC(model_emerg) - baseline_aic,
  note = "Historical emergency stays ratio (SIMD health domain)"
)

aic_log
```

Similar issue, so cannot use with Deprivation either.

## 3.3 Baseline + External operational context (when data)

```{r}
model_external <- bam(NumberOfAttendances ~
                        Age + Sex + Deprivation + oohours_pct + weekend_pct +
                        s(t, k = 35) + 
                        s(m_num, bs = "cc", k = 12) + 
                        s(HBT, bs = "re") +
                        offset(log(population)),
                      family = nb(),
                      data = risk_data,
                      method = "REML")

summary(model_external)

aic_log <- aic_log %>% add_row(
  model = "+ oohours_pct + weekend_pct",
  aic = AIC(model_external),
  delta_aic = AIC(model_external) - baseline_aic,
  note = "External operational context only (no SIMD indicators)"
)

aic_log
```

Deprivation coefficients stayed perfectly monotonic, these predictors are
external to SIMD and both significant:

- `oohours_pct` negative: HB with higher out-of-hours share have lower
  attendance rates --> areas with better out-of-hours primary care
  alternatives may divert some demand away from A&E
- `weekend_pct` positive: HB with higher weekend share have higher rates, 
  could reflect areas where weekend alternatives (GP, pharmacy) are less
  available
  
Out-of-hours and weekend attendance shares were included as HB level operational
context variables, capturing differences in the timing structure of A&E demand. 
Because these measures are derived from attendance data, they may be partially 
endogenous to the outcome-generating process. However, they may also reflect
broader structural differences between areas, including the availability and 
use of alternative urgent-care services outside standard working hours. They 
were therefore interpreted as contextual indicators rather than individual level
causal predictors.

## 3.4 Domain indicators replacing Deprivation

```{r}
model_indicators <- bam(NumberOfAttendances ~
                          Age + Sex + Income_rate + EMERG + crime_rate + 
                          drive_GP + oohours_pct + weekend_pct +
                          s(t, k = 35) + 
                          s(m_num, bs = "cc", k = 12) + 
                          s(HBT, bs = "re") +
                          offset(log(population)),
                        family = nb(),
                        data = risk_data,
                        method = "REML")

summary(model_indicators)

aic_log <- aic_log %>% add_row(
  model = "indicators (no Deprivation)",
  aic = AIC(model_indicators),
  delta_aic = AIC(model_indicators) - baseline_aic,
  note = "SIMD domain indicators replace Deprivation quintile"
)
aic_log
```

All predictors significant. Income_rate is the strongest driver. crime_rate
shows a negative sign, among areas with similar income deprivation levels,
those with higher crime rates tend to have slightly lower attendance,
possibly because high-crime areas in Scotland tend to be dense urban
locations with better access to both A&E and alternative services.

```{r}
model_indicators_nocrime <- bam(NumberOfAttendances ~
                                  Age + Sex + Income_rate + EMERG + drive_GP +
                                  oohours_pct + weekend_pct +
                                  s(t, k = 35) + 
                                  s(m_num, bs = "cc", k = 12) + 
                                  s(HBT, bs = "re") +
                                  offset(log(population)),
                                family = nb(),
                                data = risk_data,
                                method = "REML")

summary(model_indicators_nocrime)

aic_log <- aic_log %>% add_row(
  model = "indicators (no crime)",
  aic = AIC(model_indicators_nocrime),
  delta_aic = AIC(model_indicators_nocrime) - baseline_aic,
  note = "Domain indicators without crime_rate"
)

aic_log
```

AIC worsened by around 222 points when crime_rate was removed, confirming it
contributes genuine predictive value despite the counter intuitive sign.

## 3.5 Referral mix

Note: referral mix is partly a consequence of attendance volume (if
attendance surges, the proportion arriving via different routes may shift),
so any significant coefficient should be interpreted as associational,
not causal.

```{r}
# Self referral 
model_self_ref <- bam(NumberOfAttendances ~
                        Age + Sex + Deprivation + pct_ref_self +
                        s(t, k = 35) + 
                        s(m_num, bs = "cc", k = 12) + 
                        s(HBT, bs = "re") +
                        offset(log(population)),
                      family = nb(),
                      data = risk_data,
                      method = "REML")

summary(model_self_ref)

aic_log <- aic_log %>% add_row(
  model = "+ Self Referral %",
  aic = AIC(model_self_ref),
  delta_aic = AIC(model_self_ref) - baseline_aic,
  note = "Self-referral share"
)
aic_log
```

Referral mix (self-referral share) was tested as a predictor but excluded from 
the final specification despite statistical significance, because it describes 
how attenders arrived at A&E rather than who is at risk of attending, an 
operational characteristic of the attendance itself, not a population level 
risk factor. 

# 4. Interactions

## 4.1 Age x sex

```{r}
model_interaction_age_sex <- bam(NumberOfAttendances ~
                                   Age * Sex + Deprivation + oohours_pct + 
                                   weekend_pct +
                                   s(t, k = 35) + 
                                   s(m_num, bs = "cc", k = 12) + 
                                   s(HBT, bs = "re") +
                                   offset(log(population)),
                                 family = nb(),
                                 data = risk_data,
                                 method = "REML")

summary(model_interaction_age_sex)

aic_log <- aic_log %>% add_row(
  model = "baseline + Age*Sex",
  aic = AIC(model_interaction_age_sex),
  delta_aic = AIC(model_interaction_age_sex) - baseline_aic,
  note = "Interaction: does age pattern differ by sex?"
)
aic_log
```


## 4.2 Age × Deprivation

```{r}
model_interaction_age_dep <- bam(NumberOfAttendances ~
                                   Age * Deprivation + Sex + oohours_pct + 
                                   weekend_pct +
                                   s(t, k = 35) + 
                                   s(m_num, bs = "cc", k = 12) + 
                                   s(HBT, bs = "re") +
                                   offset(log(population)),
                                 family = nb(),
                                 data = risk_data,
                                 method = "REML")

summary(model_interaction_age_dep)

aic_log <- aic_log %>% add_row(
  model = "baseline + Age*Deprivation",
  aic = AIC(model_interaction_age_dep),
  delta_aic = AIC(model_interaction_age_dep) - baseline_aic,
  note = "Interaction: does deprivation gap differ by age?"
)
aic_log
```

## 4.3 Both interactions combined
```{r}
model_both_interactions <- bam(NumberOfAttendances ~
                                 Age * Sex + Age * Deprivation + oohours_pct + 
                                 weekend_pct +
                                 s(t, k = 35) + 
                                 s(m_num, bs = "cc", k = 12) + 
                                 s(HBT, bs = "re") +
                                 offset(log(population)),
                               family = nb(),
                               data = risk_data,
                               method = "REML")

summary(model_both_interactions)

aic_log <- aic_log %>% add_row(
  model = "Age*Sex + Age*Dep + external",
  aic = AIC(model_both_interactions),
  delta_aic = AIC(model_both_interactions) - baseline_aic,
  note = "Both interactions + operational context"
)
aic_log
```


# 5. Final Model

```{r}
# Final model with Deprivation quintile
final_risk_model_A <- bam(
  NumberOfAttendances ~
    Age * Sex + Age * Deprivation + oohours_pct + weekend_pct +
    s(t, k = 40) + s(m_num, bs = "cc", k = 12) + s(HBT, bs = "re") +
    offset(log(population)),
  family = nb(),
  data = risk_data,
  method = "REML"
)

# Final model with Deprivation indicators
final_risk_model_B <- bam(
  NumberOfAttendances ~
    Age * Sex + Income_rate + EMERG + crime_rate + drive_GP +
    oohours_pct + weekend_pct +
    s(t, k = 35) + s(m_num, bs = "cc", k = 12) + s(HBT, bs = "re") +
    offset(log(population)),
  family = nb(),
  data = risk_data,
  method = "REML"
)
```

## 5.1 Diagnostics

```{r}
par(mfrow = c(2, 2))
gam.check(final_risk_model_A)
par(mfrow = c(1, 1))
```

```{r}
par(mfrow = c(2, 2))
gam.check(final_risk_model_B)
par(mfrow = c(1, 1))
```

## 5.2 Smooth term plots

```{r}
plot(final_risk_model_A, pages = 1, shade = TRUE)
```

```{r}
plot(final_risk_model_B, pages = 1, shade = TRUE)
```

## 5.3 Coefficient interpretation

```{r}
final_coefs_A <- as.data.frame(summary(final_risk_model_A)$p.table)
final_coefs_A$rate_ratio <- exp(final_coefs_A$Estimate)
final_coefs_A$rr_lower <- exp(final_coefs_A$Estimate - 1.96 * final_coefs_A$`Std. Error`)
final_coefs_A$rr_upper <- exp(final_coefs_A$Estimate + 1.96 * final_coefs_A$`Std. Error`)
final_coefs_A
```

```{r}
final_coefs_B <- as.data.frame(summary(final_risk_model_B)$p.table)
final_coefs_B$rate_ratio <- exp(final_coefs_B$Estimate)
final_coefs_B$rr_lower <- exp(final_coefs_B$Estimate - 1.96 * final_coefs_B$`Std. Error`)
final_coefs_B$rr_upper <- exp(final_coefs_B$Estimate + 1.96 * final_coefs_B$`Std. Error`)
final_coefs_B
```


# 6. Supplementary analyses

## 6.1 Fixed-effect HB version

```{r}
# Final model with Deprivation quintile
final_risk_model_hb <- bam(
  NumberOfAttendances ~
    Age * Sex + Age * Deprivation + oohours_pct + weekend_pct + HBT +
    s(t, k = 40) + s(m_num, bs = "cc", k = 12) +
    offset(log(population)),
  family = nb(),
  data = risk_data,
  method = "REML"
)

# Extract and rank HB rate ratios
hb_coefs <- as.data.frame(summary(final_risk_model_hb)$p.table)
hb_coefs$term <- rownames(hb_coefs)
hb_coefs <- hb_coefs %>%
  filter(grepl("^HBT", term)) %>%
  mutate(
    HBT = gsub("^HBT", "", term),
    rate_ratio = exp(Estimate),
    rr_lower = exp(Estimate - 1.96 * `Std. Error`),
    rr_upper = exp(Estimate + 1.96 * `Std. Error`)
  ) %>%
  arrange(desc(rate_ratio))

hb_coefs
```


```{r}
hb_coefs_plt <- ggplot(hb_coefs, aes(x = fct_reorder(HBT, rate_ratio), y = rate_ratio)) +
  geom_point(colour = phs_colours("phs-blue"), size = 3) +
  geom_errorbar(aes(ymin = rr_lower, ymax = rr_upper), width = 0.3,
                colour = phs_colours("phs-blue")) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey30") +
  coord_flip() +
  labs(x = NULL, y = "Rate Ratio (vs. reference HB)",
       title = "Health Board Attendance Rate Ratios\n(adjusted for Age, Sex, Deprivation, interactions)") +
  theme_phs()

hb_coefs_plt

# ggsave("figures/risk_model/hb_rate_ratios.pdf", plot = hb_coefs_plt, width = 8, height = 6)
```


## 6.2 GAM robustness check
Refit the final specifications with gam not bam 

```{r}
# Final model with Deprivation quintile
final_risk_model_A_gam <- gam(
  NumberOfAttendances ~
    Age * Sex + Age * Deprivation + oohours_pct + weekend_pct +
    s(t, k = 40) + s(m_num, bs = "cc", k = 12) + s(HBT, bs = "re") +
    offset(log(population)),
  family = nb(),
  data = risk_data,
  method = "REML"
)

# add comparison of coeffs
```

```{r}
# Final model with Deprivation indicators
final_risk_model_B_gam <- gam(
  NumberOfAttendances ~
    Age * Sex + Income_rate + EMERG + crime_rate + drive_GP +
    oohours_pct + weekend_pct +
    s(t, k = 35) + s(m_num, bs = "cc", k = 12) + s(HBT, bs = "re") +
    offset(log(population)),
  family = nb(),
  data = risk_data,
  method = "REML"
)

# add comparison of coeffs
```


## 6.3 INLA robustness check

Refit the final specification A in INLA to confirm consistency across
frequentist (GAM) and Bayesian (INLA) estimation frameworks.

```{r}
risk_data_inla <- risk_data %>%
  mutate(
    HBT_id = as.numeric(HBT),
    t_rounded = round(t)
  )

inla_risk <- inla(
  NumberOfAttendances ~
    Age * Sex + Age * Deprivation + oohours_pct + weekend_pct +
    f(t_rounded, model = "rw2", constr = TRUE) +
    f(m_num, model = "rw1", cyclic = TRUE, constr = TRUE) +
    f(HBT_id, model = "iid") +
    offset(log(population)),
  family = "nbinomial",
  data = risk_data_inla,
  control.compute = list(dic = TRUE, waic = TRUE),
  control.predictor = list(compute = TRUE)
)

summary(inla_risk)
```





