---
title: "PHS Admissions Analysis"
author: "Natalia Montalvo Cabornero"
date: "2026-08-14"
output: html_document
---

```{r setup, include=FALSE}

source("R/admissions_model_data.R")

# Make sure the figure output folders exist
dir.create("figures/admissions_model", recursive = TRUE, showWarnings = FALSE)
```


# 1. Data recap

Given that someone has already attended A&E, what factors predict whether they 
are admitted to hospital?

- **Outcome**: `cbind(admitted, not_admitted)` with a binomial family

**Predictors available**:

- *Patient-level (within discharge data)*: `Age`, `DepartmentType`
- *Site-level context*: `TreatmentLocation` (as random effect), referral mix
  (`pct_ref_Self Referral`, `pct_ref_Ambulance`, etc.)
- *Operational context*: `oohours_pct`, `weekend_pct`
- *Area-level deprivation context*: SIMD indicators collapsed to HB level
  (Income_rate, Employment_rate, etc.) — note these describe the HB overall,
  not the individual patient's deprivation, since discharges_data has no
  Deprivation column
- *Time*: trend `s(t)` and seasonality `s(m_num, bs = "cc")`

**Key limitation**: no `Sex` or `Deprivation` columns in discharges_data, so
this model cannot test whether admission probability differs by sex or by
individual patient deprivation.

```{r}
dim(admission_model_data)
summarise_columns(admission_model_data)
```

4369 rows (14.4%) have missing Age, which also causes the referral-mix columns 
to fail their join (since referral context is matched at the Month × HBT × 
TreatmentLocation × DepartmentType × Age level). The missingness rate is 
consistent across all referral columns, confirming it originates from the 
missing Age rather than from gaps in the referral source data itself.

## 1.1 Overall admission rates 

```{r}
admission_model_data %>%
  filter(!is.na(Age), Age != "") %>%
  summarise(
    total_attendances = sum(total_attendances),
    total_admitted = sum(admitted),
    admission_rate = total_admitted / total_attendances * 100
  )
```


## 1.2 Admission rates by department type

```{r}
admission_model_data %>%
  filter(!is.na(Age), Age != "") %>%
  group_by(DepartmentType) %>%
  summarise(
    total_attendances = sum(total_attendances),
    total_admitted = sum(admitted),
    admission_rate = total_admitted / total_attendances * 100,
    pct_of_all_attendances = total_attendances /
      sum(admission_model_data %>%
            filter(!is.na(Age), Age != "") %>%
            pull(total_attendances)) * 100
  )
```

Type 1 (major ED) accounts for 88% of all attendances with an admission rate of 
27.3%, while Type 3 (minor injury units) only the 12% of attendances with a 
lower admission rate of 8.9%. 

(Potentially restrict the models to Type 1 department.)


## 1.3 Admission rates by age group 

```{r}
# Both departments
admission_model_data %>%
  filter(!is.na(Age), Age != "") %>%
  group_by(Age) %>%
  summarise(
    total_attendances = sum(total_attendances),
    total_admitted = sum(admitted),
    admission_rate = total_admitted / total_attendances * 100
  ) %>%
  arrange(desc(admission_rate))
```


```{r}
# Only Type 1 department
admission_model_data %>%
  filter(!is.na(Age), Age != "", DepartmentType == "Type 1") %>%
  group_by(Age) %>%
  summarise(
    total_attendances = sum(total_attendances),
    total_admitted = sum(admitted),
    admission_rate = total_admitted / total_attendances * 100
  ) %>%
  arrange(desc(admission_rate))
```


# 2. Initial model exploration (GAM)

A binomial GAM was fitted first as the natural starting point for grouped
binary outcome data.

```{r}
# Prepare GAM model data
admission_data_gam <- admission_model_data %>%
  filter(!is.na(admitted), !is.na(total_attendances),
         total_attendances > 0,
         !is.na(Age), Age != "") %>%
  mutate(
    Age = factor(Age),
    HBT = factor(HBT),
    DepartmentType = factor(DepartmentType),
    TreatmentLocation = factor(TreatmentLocation)
  )

cat("Number of rows in the data:", nrow(admission_data_gam), "\n")
```

## 2.1 Binomial GAM baseline

```{r}
baseline_admission_bin <- gam(
  cbind(admitted, not_admitted) ~
    Age + DepartmentType +
    s(t, k = 30) +
    s(m_num, bs = "cc", k = 12) +
    s(TreatmentLocation, bs = "re"),
  family = binomial(),
  data = admission_data_gam,
  method = "REML"
)

summary(baseline_admission_bin)

# baseline_admission_bin2 <- bam(
#   cbind(admitted, not_admitted) ~
#     Age + DepartmentType +
#     s(t, k = 30) +
#     s(m_num, bs = "cc", k = 12) +
#     s(TreatmentLocation, bs = "re"),
#   family = binomial(),
#   data = admission_data_gam,
#   method = "REML"
# )
# 
# summary(baseline_admission_bin2)
```
Differences with bam 

```{r}
par(mfrow = c(2, 2))
gam.check(baseline_admission_bin)
par(mfrow = c(1, 1))
```

## 2.2 Overdispersion check

```{r}
# Dispersion ratio: residual deviance / residual df
# Should be close to 1 for binomial; >>1 indicates overdispersion
dispersion_ratio_admission <- sum(
                     residuals(baseline_admission_bin, 
                     type = "deviance")^2) / baseline_admission_bin$df.residual
dispersion_ratio_admission
```

## 2.3 Check the Quasi-binomial distribution

```{r}
baseline_admission <- gam(
  cbind(admitted, not_admitted) ~
    Age + DepartmentType +
    s(t, k = 30) +
    s(m_num, bs = "cc", k = 12) +
    s(TreatmentLocation, bs = "re"),
  family = quasibinomial(),
  data = admission_data_gam,
  method = "REML"
)

summary(baseline_admission)
summary(baseline_admission)$dispersion
```

```{r}
# Compare binomial vs quasibinomial side by side
binomial_coefs <- summary(baseline_admission_bin)$p.table
quasi_coefs <- summary(baseline_admission)$p.table

cbind(
  binomial_est = binomial_coefs[, "Estimate"],
  quasi_est = quasi_coefs[, "Estimate"],
  binomial_se = binomial_coefs[, "Std. Error"],
  quasi_se = quasi_coefs[, "Std. Error"]
)
```

The coefficients are very similar except for department type, which shifted from
-6.04 (binomial) to -5.15 (quasibinomial), while all other parametric 
coefficients remained stable to the third decimal place. This suggests numerical 
sensitivity in the quasi-likelihood estimation for near-boundary predictions, 
driven by the disparity in admission rates between department types.

```{r}
par(mfrow = c(2, 2))
gam.check(baseline_admission)
par(mfrow = c(1, 1))
```

```{r}
gam_type1_only <- gam(
  cbind(admitted, not_admitted) ~
    Age +
    s(t, k = 30) +
    s(m_num, bs = "cc", k = 12) +
    s(TreatmentLocation, bs = "re"),
  family = binomial(),
  data = admission_data_gam %>% filter(DepartmentType == "Type 1"),
  method = "REML"
)

# Overdispersion check on Type 1 only
dispersion_type1 <- sum(residuals(gam_type1_only, type = "deviance")^2) /
  gam_type1_only$df.residual
dispersion_type1

# Also worth a quick look at the diagnostics
par(mfrow = c(2, 2))
gam.check(gam_type1_only)
par(mfrow = c(1, 1))
```

```{r}
gam_type1_only2 <- gam(
  cbind(admitted, not_admitted) ~
    Age +
    s(t, k = 30) +
    s(m_num, bs = "cc", k = 12) +
    s(TreatmentLocation, bs = "re"),
  family = quasibinomial(),
  data = admission_data_gam %>% filter(DepartmentType == "Type 1"),
  method = "REML"
)
summary(gam_type1_only2)

# Overdispersion check on Type 1 only
dispersion_type1_2 <- sum(residuals(gam_type1_only2, type = "deviance")^2) /
  gam_type1_only2$df.residual
dispersion_type1_2

# Also worth a quick look at the diagnostics
par(mfrow = c(2, 2))
gam.check(gam_type1_only2)
par(mfrow = c(1, 1))
```



```{r}
# Compare binomial vs quasibinomial side by side TYPE 1
binomial_coefs <- summary(gam_type1_only)$p.table
quasi_coefs <- summary(gam_type1_only2)$p.table

cbind(
  binomial_est = binomial_coefs[, "Estimate"],
  quasi_est = quasi_coefs[, "Estimate"],
  binomial_se = binomial_coefs[, "Std. Error"],
  quasi_se = quasi_coefs[, "Std. Error"]
)
```


## 2.4 Framework decision

1. **Overdispersion** (dispersion ratio = 6.08 under binomial; estimated 
   scale = 6.56 under quasibinomial): the binomial variance assumption 
   underestimates the true variability in admission counts. A quasibinomial()
   was tested: while standard errors were correctly inflated, the DepartmentType 
   coefficient shifted from -6.04 (binomial) to -5.15 (quasibinomial), a change 
   of 0.89 on the log-odds scale, while all other parametric coefficients 
   remained stable to the third decimal place. This suggests numerical 
   sensitivity in the quasi-likelihood estimation for near-boundary predictions. 
   Additionally, the quasibinomial family loses AIC (the primary model 
   comparison metric used throughout the risk model), forcing reliance on 
   less precise alternatives (anova(..., test = "F") or deviance-explained) for 
   predictor comparison.

2. **Structural differences between department types**: Type 1 departments 
   showed a 29.1% admission rate across 88% of all attendances, while Type 3 
   (minor injury units) showed 12.7% across the remaining 12%. Although Type 3's
   rate is not negligible, the two department types represent fundamentally 
   different clinical settings where the factors influencing admission decisions
   likely differ — major EDs with full inpatient capacity versus minor injury units with limited admission pathways. Diagnostic plots confirmed this structural difference produced residual heterogeneity: the QQ plot showed severe S-shaped departure (deviance residuals reaching +30), and the residuals-vs-linear-predictor plot showed a clear fan pattern, with variance increasing markedly across the range of predicted probabilities.


check this again

##################################################################

# 3. INLA baseline model

## 3.1 Prepare INLA data

```{r}
# Filter to Type 1 only, rename columns with spaces for cleaner formula
# syntax, create numeric indices for random effects (INLA requires numeric,
# not factor, for f() terms), and add observation-level ID for the
# overdispersion random effect
admission_inla_data <- admission_model_data %>%
  filter(DepartmentType == "Type 1",
         !is.na(Age), Age != "",
         !is.na(admitted), total_attendances > 0) %>%
  rename(
    pct_ref_self = `pct_ref_Self Referral`,
    pct_ref_ambulance = `pct_ref_Ambulance`
  ) %>%
  mutate(
    Age = factor(Age),
    TreatmentLocation_id = as.numeric(factor(TreatmentLocation)),
    HBT_id = as.numeric(factor(HBT)),
    t_rounded = round(as.numeric(m_date - min(m_date)) / 30),
    obs_id = row_number()
  )

cat("Number of rows:", nrow(admission_inla_data), "\n")
cat("Number of treatment locations:", n_distinct(admission_inla_data$TreatmentLocation), "\n")

# Verify time and month indices are what INLA's rw2/rw1 expect
range(admission_inla_data$t_rounded)
table(admission_inla_data$m_num)

# Location ID mapping (for interpreting site random effects later)
location_mapping <- admission_inla_data %>%
  distinct(TreatmentLocation, TreatmentLocation_id, HBT) %>%
  arrange(TreatmentLocation_id)
location_mapping
```

## 3.2 Model specification justification

```{r}
baseline_admission_inla <- inla(
  admitted ~
    Age +
    f(t_rounded, model = "rw2", constr = TRUE) +
    f(m_num, model = "rw1", cyclic = TRUE, constr = TRUE) +
    f(TreatmentLocation_id, model = "iid") +
    f(obs_id, model = "iid"),
  family = "binomial",
  Ntrials = admission_inla_data$total_attendances,
  data = admission_inla_data,
  control.compute = list(dic = TRUE, waic = TRUE, config = TRUE),
  control.predictor = list(compute = TRUE)
)

summary(baseline_admission_inla)
```


## 3.4 Coefficient interpretation

```{r}
# Fixed effects: INLA reports posterior mean and 95% credible intervals
# Exponentiate to get odds ratios (logit link)
fixed_effects <- baseline_admission_inla$summary.fixed
fixed_effects$odds_ratio <- exp(fixed_effects$mean)
fixed_effects$or_lower <- exp(fixed_effects$`0.025quant`)
fixed_effects$or_upper <- exp(fixed_effects$`0.975quant`)
fixed_effects
```

Cross-check against the descriptive admission rates in Section 1.3:
- 75 plus should show the highest odds of admission
- Under 18 should show lower odds than the 18-24 reference
- The odds should increase monotonically with age


## 3.5 Hyperparameters and random effect summaries

```{r}
# The obs_id precision tells you how much overdispersion the observation-
# level RE is absorbing — low precision (high variance) = substantial
# overdispersion being handled
baseline_admission_inla$summary.hyperpar
```


## 3.6 Smooth term plots

```{r}
par(mfrow = c(1, 3))

# Time trend
plot(baseline_admission_inla$summary.random$t_rounded$ID,
     baseline_admission_inla$summary.random$t_rounded$mean,
     type = "l", xlab = "t (months since start)", ylab = "Effect",
     main = "Time trend f(t)")

# Seasonal cycle
plot(baseline_admission_inla$summary.random$m_num$ID,
     baseline_admission_inla$summary.random$m_num$mean,
     type = "l", xlab = "Month", ylab = "Effect",
     main = "Seasonal cycle f(m_num)")

# Site-level random intercepts (ranked)
site_re <- baseline_admission_inla$summary.random$TreatmentLocation_id
plot(sort(site_re$mean), seq_along(site_re$mean),
     xlab = "Random intercept", ylab = "Site rank",
     main = "Site random effects", pch = 16)
abline(v = 0, lty = 2)

par(mfrow = c(1, 1))
```

```{r}
baseline_no_olre <- inla(
  admitted ~
    Age +
    f(t_rounded, model = "rw2", constr = TRUE) +
    f(m_num, model = "rw1", cyclic = TRUE, constr = TRUE) +
    f(TreatmentLocation_id, model = "iid"),
  family = "binomial",
  Ntrials = admission_inla_data$total_attendances,
  data = admission_inla_data,
  control.compute = list(dic = TRUE, waic = TRUE)
)

summary(baseline_no_olre)

# Compare
tibble(
  model = c("Without OLRE", "With OLRE"),
  WAIC = c(baseline_no_olre$waic$waic, baseline_admission_inla$waic$waic),
  DIC = c(baseline_no_olre$dic$dic, baseline_admission_inla$dic$dic)
)
```


```{r}
baseline_betabin <- inla(
  admitted ~
    Age +
    f(t_rounded, model = "rw2", constr = TRUE) +
    f(m_num, model = "rw1", cyclic = TRUE, constr = TRUE) +
    f(TreatmentLocation_id, model = "iid"),
  family = "betabinomial",
  Ntrials = admission_inla_data$total_attendances,
  data = admission_inla_data,
  control.compute = list(dic = TRUE, waic = TRUE)
)

tibble(
  model = c("Binomial + OLRE", "Beta-Binomial"),
  WAIC = c(baseline_admission_inla$waic$waic, baseline_betabin$waic$waic)
)
```



## 3.7 Record baseline WAIC

```{r}
baseline_waic <- baseline_admission_inla$waic$waic
baseline_waic

waic_log <- tribble(
  ~model, ~waic, ~delta_waic, ~note,
  "baseline", baseline_waic, 0,
  "Age + time + season + site RE + obs RE"
)

waic_log
```


# 4. Adding contextual predictors

## 4.1 Self-referral share

Hypothesis: self-referred patients may be lower acuity on average (chose
to come themselves rather than being sent by a GP or arriving by
ambulance), so a higher self-referral share could predict lower admission
probability.

```{r}
model_adm_referral <- inla(
  admitted ~
    Age + pct_ref_self +
    f(t_rounded, model = "rw2", constr = TRUE) +
    f(m_num, model = "rw1", cyclic = TRUE, constr = TRUE) +
    f(TreatmentLocation_id, model = "iid") +
    f(obs_id, model = "iid"),
  family = "binomial",
  Ntrials = admission_inla_data$total_attendances,
  data = admission_inla_data,
  control.compute = list(dic = TRUE, waic = TRUE)
)

model_adm_referral$summary.fixed
model_adm_referral$waic$waic - baseline_waic

waic_log <- waic_log %>% add_row(
  model = "+ Self Referral %",
  waic = model_adm_referral$waic$waic,
  delta_waic = model_adm_referral$waic$waic - baseline_waic,
  note = "Self-referral share (site x month x age level)"
)
waic_log
```


## 4.2 Ambulance referral share

Hypothesis: ambulance arrivals are typically higher acuity, predicting
higher admission probability.

```{r}
model_adm_ambulance <- inla(
  admitted ~
    Age + pct_ref_ambulance +
    f(t_rounded, model = "rw2", constr = TRUE) +
    f(m_num, model = "rw1", cyclic = TRUE, constr = TRUE) +
    f(TreatmentLocation_id, model = "iid") +
    f(obs_id, model = "iid"),
  family = "binomial",
  Ntrials = admission_inla_data$total_attendances,
  data = admission_inla_data,
  control.compute = list(dic = TRUE, waic = TRUE)
)
summary(model_adm_ambulance)
model_adm_ambulance$summary.fixed
model_adm_ambulance$waic$waic - baseline_waic

waic_log <- waic_log %>% add_row(
  model = "+ Ambulance %",
  waic = model_adm_ambulance$waic$waic,
  delta_waic = model_adm_ambulance$waic$waic - baseline_waic,
  note = "Ambulance referral share (site x month x age level)"
)

waic_log
```


## 4.3 Out-of-hours share

Hypothesis: out-of-hours attendances may have different case mix (fewer
minor complaints, more genuine emergencies), potentially predicting
higher admission rates.

```{r}
model_adm_ooh <- inla(
  admitted ~
    Age + oohours_pct +
    f(t_rounded, model = "rw2", constr = TRUE) +
    f(m_num, model = "rw1", cyclic = TRUE, constr = TRUE) +
    f(TreatmentLocation_id, model = "iid") +
    f(obs_id, model = "iid"),
  family = "binomial",
  Ntrials = admission_inla_data$total_attendances,
  data = admission_inla_data,
  control.compute = list(dic = TRUE, waic = TRUE)
)

model_adm_ooh$summary.fixed
model_adm_ooh$waic$waic - baseline_waic

waic_log <- waic_log %>% add_row(
  model = "+ oohours_pct",
  waic = model_adm_ooh$waic$waic,
  delta_waic = model_adm_ooh$waic$waic - baseline_waic,
  note = "Out-of-hours share (site x month level)"
)
waic_log

```


## 4.4 Income deprivation (HB-level context)

Note: this is the HB's overall income deprivation rate, not the individual
patient's deprivation — no collinearity risk since there is no row-level
Deprivation variable to compete with.

```{r}
model_adm_income <- inla(
  admitted ~
    Age + Income_rate +
    f(t_rounded, model = "rw2", constr = TRUE) +
    f(m_num, model = "rw1", cyclic = TRUE, constr = TRUE) +
    f(TreatmentLocation_id, model = "iid") +
    f(obs_id, model = "iid"),
  family = "binomial",
  Ntrials = admission_inla_data$total_attendances,
  data = admission_inla_data,
  control.compute = list(dic = TRUE, waic = TRUE)
)

model_adm_income$summary.fixed
model_adm_income$waic$waic - baseline_waic

waic_log <- waic_log %>% add_row(
  model = "+ Income_rate",
  waic = model_adm_income$waic$waic,
  delta_waic = model_adm_income$waic$waic - baseline_waic,
  note = "HB-level income deprivation rate"
)

waic_log
```


## 4.5 Compare all single additions

```{r}
waic_log %>% arrange(delta_waic)
```


```{r}
# Within-site deviation from the site's own mean
admission_inla_data <- admission_inla_data %>%
  group_by(TreatmentLocation_id) %>%
  mutate(
    site_mean_self = mean(pct_ref_self, na.rm = TRUE),
    self_ref_deviation = pct_ref_self - site_mean_self,
    site_mean_ambulance = mean(pct_ref_ambulance, na.rm = TRUE),
    ambulance_deviation = pct_ref_ambulance - site_mean_ambulance
  ) %>%
  ungroup()

model_adm_self_dev <- inla(
  admitted ~
    Age + self_ref_deviation +
    f(t_rounded, model = "rw2", constr = TRUE) +
    f(m_num, model = "rw1", cyclic = TRUE, constr = TRUE) +
    f(TreatmentLocation_id, model = "iid") +
    f(obs_id, model = "iid"),
  family = "binomial",
  Ntrials = admission_inla_data$total_attendances,
  data = admission_inla_data,
  control.compute = list(dic = TRUE, waic = TRUE)
)
model_adm_self_dev$waic$waic
model_adm_self_dev$waic$waic - baseline_waic
```

```{r}
admission_inla_data <- admission_inla_data %>%
  group_by(TreatmentLocation_id) %>%
  mutate(
    site_mean_weekend = mean(weekend_pct, na.rm = TRUE),
    weekend_deviation = weekend_pct - site_mean_weekend
  ) %>%
  ungroup()

model_adm_weekend_dev <- inla(
  admitted ~
    Age + weekend_deviation +
    f(t_rounded, model = "rw2", constr = TRUE) +
    f(m_num, model = "rw1", cyclic = TRUE, constr = TRUE) +
    f(TreatmentLocation_id, model = "iid") +
    f(obs_id, model = "iid"),
  family = "binomial",
  Ntrials = admission_inla_data$total_attendances,
  data = admission_inla_data,
  control.compute = list(dic = TRUE, waic = TRUE)
)

model_adm_weekend_dev$waic$waic
model_adm_weekend_dev$waic$waic - baseline_waic
```


# 5. Combined model

Based on the results above, combine the predictors that individually
improved WAIC. Check whether they still help when present together
(correlated predictors may become redundant in combination).

```{r}
# Adjust this specification based on which predictors improved WAIC above
model_adm_combined <- inla(
  admitted ~
    Age + pct_ref_self + pct_ref_ambulance +
    oohours_pct + Income_rate +
    f(t_rounded, model = "rw2", constr = TRUE) +
    f(m_num, model = "rw1", cyclic = TRUE, constr = TRUE) +
    f(TreatmentLocation_id, model = "iid") +
    f(obs_id, model = "iid"),
  family = "binomial",
  Ntrials = admission_inla_data$total_attendances,
  data = admission_inla_data,
  control.compute = list(dic = TRUE, waic = TRUE, config = TRUE),
  control.predictor = list(compute = TRUE)
)

summary(model_adm_combined)
```

```{r}
# Final coefficient table with odds ratios and credible intervals
final_fixed <- model_adm_combined$summary.fixed
final_fixed$odds_ratio <- exp(final_fixed$mean)
final_fixed$or_lower <- exp(final_fixed$`0.025quant`)
final_fixed$or_upper <- exp(final_fixed$`0.975quant`)
final_fixed

waic_log <- waic_log %>% add_row(
  model = "combined",
  waic = model_adm_combined$waic$waic,
  delta_waic = model_adm_combined$waic$waic - baseline_waic,
  note = "All improving predictors together"
)

waic_log %>% arrange(delta_waic)
```

```{r, fig.width=10, fig.height=5}
# Final model smooth term plots
par(mfrow = c(1, 3))

plot(model_adm_combined$summary.random$t_rounded$ID,
     model_adm_combined$summary.random$t_rounded$mean,
     type = "l", xlab = "t (months since start)", ylab = "Effect",
     main = "Time trend f(t)")

plot(model_adm_combined$summary.random$m_num$ID,
     model_adm_combined$summary.random$m_num$mean,
     type = "l", xlab = "Month", ylab = "Effect",
     main = "Seasonal cycle f(m_num)")

site_re <- model_adm_combined$summary.random$TreatmentLocation_id
plot(sort(site_re$mean), seq_along(site_re$mean),
     xlab = "Random intercept", ylab = "Site rank",
     main = "Site random effects", pch = 16)
abline(v = 0, lty = 2)

par(mfrow = c(1, 1))
```

```{r}
# Hyperparameter summaries
model_adm_combined$summary.hyperpar
```


# 6. Sensitivity: narrow vs broad admission definition

Compare the final model under both definitions to check whether results
are robust to including transfers ("Transferred to Other Hospital/Service").

```{r}
# Create broad-definition data with consistent factor levels and IDs
# Using the same factor levels as narrow data ensures TreatmentLocation_id
# mapping is consistent between the two datasets
location_levels <- levels(factor(admission_inla_data$TreatmentLocation))

admission_inla_data_broad <- admission_model_data_broad %>%
  filter(DepartmentType == "Type 1",
         !is.na(Age), Age != "",
         !is.na(admitted), total_attendances > 0) %>%
  rename(
    pct_ref_self = `pct_ref_Self Referral`,
    pct_ref_ambulance = `pct_ref_Ambulance`
  ) %>%
  mutate(
    Age = factor(Age),
    TreatmentLocation_id = as.numeric(factor(TreatmentLocation,
                                              levels = location_levels)),
    t_rounded = round(as.numeric(m_date - min(m_date)) / 30),
    obs_id = row_number()
  )

# Refit final specification with broad definition
model_adm_broad <- inla(
  admitted ~
    Age + pct_ref_self + pct_ref_ambulance +
    oohours_pct + Income_rate +
    f(t_rounded, model = "rw2", constr = TRUE) +
    f(m_num, model = "rw1", cyclic = TRUE, constr = TRUE) +
    f(TreatmentLocation_id, model = "iid") +
    f(obs_id, model = "iid"),
  family = "binomial",
  Ntrials = admission_inla_data_broad$total_attendances,
  data = admission_inla_data_broad,
  control.compute = list(dic = TRUE, waic = TRUE)
)

# Compare coefficients side by side
narrow_fixed <- model_adm_combined$summary.fixed %>%
  as.data.frame() %>%
  mutate(definition = "narrow", or = exp(mean))

broad_fixed <- model_adm_broad$summary.fixed %>%
  as.data.frame() %>%
  mutate(definition = "broad", or = exp(mean))

bind_rows(
  narrow_fixed %>% select(definition, mean, or, `0.025quant`, `0.975quant`),
  broad_fixed %>% select(definition, mean, or, `0.025quant`, `0.975quant`)
)
```


