---
title: "PHS Admissions Analysis"
author: "Natalia Montalvo Cabornero"
date: "2026-08-14"
output: html_document
---

```{r setup, include=FALSE}

source("R/admissions_model_data.R")

# Make sure the figure output folders exist
dir.create("figures/admissions_model", recursive = TRUE, showWarnings = FALSE)
```


# 1. Data recap

Given that someone has already attended A&E, what factors predict whether they 
are admitted to hospital?

- **Outcome**: `cbind(admitted, not_admitted)` with a binomial family

**Predictors available**:

- *Patient-level (within discharge data)*: `Age`, `DepartmentType`
- *Site-level context*: `TreatmentLocation` (as random effect), referral mix
  (`pct_ref_Self Referral`, `pct_ref_Ambulance`, etc.)
- *Operational context*: `oohours_pct`, `weekend_pct`
- *Area-level deprivation context*: SIMD indicators collapsed to HB level
  (Income_rate, Employment_rate, etc.) — note these describe the HB overall,
  not the individual patient's deprivation, since discharges_data has no
  Deprivation column
- *Time*: trend `s(t)` and seasonality `s(m_num, bs = "cc")`

**Key limitation**: no `Sex` or `Deprivation` columns in discharges_data, so
this model cannot test whether admission probability differs by sex or by
individual patient deprivation.

```{r}
dim(admission_model_data)
summarise_columns(admission_model_data)
```

4369 rows (14.4%) have missing Age, which also causes the referral mix columns 
to fail their join (since referral context is matched at the Month × HBT × 
TreatmentLocation × DepartmentType × Age level).

## 1.1 Overall admission rates 

```{r}
admission_model_data %>%
  filter(!is.na(Age), Age != "") %>%
  summarise(
    total_attendances = sum(total_attendances),
    total_admitted = sum(admitted),
    admission_rate = total_admitted / total_attendances * 100
  )
```


## 1.2 Admission rates by department type

```{r}
admission_model_data %>%
  filter(!is.na(Age), Age != "") %>%
  group_by(DepartmentType) %>%
  summarise(
    total_attendances = sum(total_attendances),
    total_admitted = sum(admitted),
    admission_rate = total_admitted / total_attendances * 100,
    pct_of_all_attendances = total_attendances /
      sum(admission_model_data %>%
            filter(!is.na(Age), Age != "") %>%
            pull(total_attendances)) * 100
  )
```

Type 1 (major ED) accounts for 88% of all attendances with an admission rate of 
27.3%, while Type 3 (minor injury units) only the 12% of attendances with a 
lower admission rate of 8.9%. 


## 1.3 Admission rates by age group 

```{r}
# Both departments
admission_model_data %>%
  filter(!is.na(Age), Age != "") %>%
  group_by(Age) %>%
  summarise(
    total_attendances = sum(total_attendances),
    total_admitted = sum(admitted),
    admission_rate = total_admitted / total_attendances * 100
  ) %>%
  arrange(desc(admission_rate))
```


```{r}
# Only Type 1 department
admission_model_data %>%
  filter(!is.na(Age), Age != "", DepartmentType == "Type 1") %>%
  group_by(Age) %>%
  summarise(
    total_attendances = sum(total_attendances),
    total_admitted = sum(admitted),
    admission_rate = total_admitted / total_attendances * 100
  ) %>%
  arrange(desc(admission_rate))
```


# 2. Initial model exploration (GAM)

A binomial GAM was fitted first as the natural starting point for grouped
binary outcome data.

```{r}
# Prepare GAM model data
admission_data_gam <- admission_model_data %>%
  filter(!is.na(admitted), !is.na(total_attendances),
         total_attendances > 0,
         !is.na(Age), Age != "") %>%
  mutate(
    Age = factor(Age),
    HBT = factor(HBT),
    DepartmentType = factor(DepartmentType),
    TreatmentLocation = factor(TreatmentLocation)
  )

cat("Number of rows in the data:", nrow(admission_data_gam), "\n")
```

## 2.1 Binomial GAM baseline

```{r}
baseline_adm_bin_gam <- gam(
  cbind(admitted, not_admitted) ~
    Age + DepartmentType +
    s(t, k = 30) +
    s(m_num, bs = "cc", k = 12) +
    s(TreatmentLocation, bs = "re"),
  family = binomial(),
  data = admission_data_gam,
  method = "REML")

summary(baseline_adm_bin_gam)
```

```{r}
par(mfrow = c(2, 2))
gam.check(baseline_adm_bin_gam)
par(mfrow = c(1, 1))
```

## 2.2 Overdispersion check

```{r}
# Dispersion ratio
dispersion_ratio_bin <- sum(
              residuals(baseline_adm_bin_gam, 
              type = "deviance")^2) / baseline_adm_bin_gam$df.residual

dispersion_ratio_bin
```

## 2.3 Check the Quasi-binomial distribution

To account for the overdispersion

```{r}
baseline_adm_qb_gam <- gam(
  cbind(admitted, not_admitted) ~
    Age + DepartmentType +
    s(t, k = 30) +
    s(m_num, bs = "cc", k = 12) +
    s(TreatmentLocation, bs = "re"),
  family = quasibinomial(),
  data = admission_data_gam,
  method = "REML")

summary(baseline_adm_qb_gam)
summary(baseline_adm_qb_gam)$dispersion
```

```{r}
# Compare binomial vs quasibinomial
binomial_coefs <- summary(baseline_adm_bin_gam)$p.table
quasi_coefs <- summary(baseline_adm_qb_gam)$p.table

cbind(binomial_est = binomial_coefs[, "Estimate"],
      quasi_est = quasi_coefs[, "Estimate"],
      binomial_se = binomial_coefs[, "Std. Error"],
      quasi_se = quasi_coefs[, "Std. Error"])
```

The coefficients are very similar except for department type, which shifted from
-6.04 (binomial) to -5.15 (quasibinomial), while all other parametric 
coefficients remained stable to the third decimal place. 

```{r}
par(mfrow = c(2, 2))
gam.check(baseline_adm_qb_gam)
par(mfrow = c(1, 1))
```

```{r}
gam_type1_bin <- gam(
  cbind(admitted, not_admitted) ~
    Age +
    s(t, k = 30) +
    s(m_num, bs = "cc", k = 12) +
    s(TreatmentLocation, bs = "re"),
  family = binomial(),
  data = admission_data_gam %>% filter(DepartmentType == "Type 1"),
  method = "REML")

summary(gam_type1_bin)

# Overdispersion check on Type 1 only
dispersion_type1_bin <- sum(residuals(gam_type1_bin, type = "deviance")^2) /
  gam_type1_bin$df.residual

dispersion_type1_bin

# Also worth a quick look at the diagnostics
par(mfrow = c(2, 2))
gam.check(gam_type1_bin)
par(mfrow = c(1, 1))
```

```{r}
gam_type1_qb <- gam(
  cbind(admitted, not_admitted) ~
    Age +
    s(t, k = 30) +
    s(m_num, bs = "cc", k = 12) +
    s(TreatmentLocation, bs = "re"),
  family = quasibinomial(),
  data = admission_data_gam %>% filter(DepartmentType == "Type 1"),
  method = "REML"
)
summary(gam_type1_qb)

# Overdispersion check on Type 1 only
dispersion_type1_qb <- sum(residuals(gam_type1_qb, type = "deviance")^2) /
  gam_type1_qb$df.residual

dispersion_type1_qb

# Also worth a quick look at the diagnostics
par(mfrow = c(2, 2))
gam.check(gam_type1_qb)
par(mfrow = c(1, 1))
```



```{r}
# Compare binomial vs quasibinomial side by side TYPE 1
binomial_coefs <- summary(gam_type1_bin)$p.table
quasi_coefs <- summary(gam_type1_qb)$p.table

cbind(
  binomial_est = binomial_coefs[, "Estimate"],
  quasi_est = quasi_coefs[, "Estimate"],
  binomial_se = binomial_coefs[, "Std. Error"],
  quasi_se = quasi_coefs[, "Std. Error"]
)
```


## 2.4 Framework decision

1. **Overdispersion** (dispersion ratio = 6.08 under binomial; estimated 
   scale = 6.56 under quasibinomial): the binomial variance assumption 
   underestimates the true variability in admission counts. A quasibinomial()
   was tested: while standard errors were correctly inflated, the DepartmentType 
   coefficient shifted from -6.04 (binomial) to -5.15 (quasibinomial), a change 
   of 0.89 on the log-odds scale, while all other parametric coefficients 
   remained stable to the third decimal place.

2. **Structural differences between department types**: Type 1 departments 
   showed a 29.1% admission rate across 88% of all attendances, while Type 3 
   (minor injury units) showed 12.7% across the remaining 12%. Although Type 3's
   rate is not negligible, the two department types represent fundamentally 
   different clinical settings where the factors influencing admission decisions
   likely differ.


# 3. INLA model

## 3.1 Prepare INLA data

```{r}
admission_inla_data <- admission_model_data %>%
  filter(DepartmentType == "Type 1",
         !is.na(Age), Age != "",
         !is.na(admitted), total_attendances > 0) %>%
  rename(pct_ref_self = `pct_ref_Self Referral`,
         pct_ref_ambulance = `pct_ref_Ambulance`) %>%
  mutate(Age = factor(Age), 
         TreatmentLocation_id = as.numeric(factor(TreatmentLocation)),
         HBT_id = as.numeric(factor(HBT)),
         t_rounded = round(as.numeric(m_date - min(m_date)) / 30),
         obs_id = row_number())

cat("Number of rows:", nrow(admission_inla_data), "\n")
cat("Number of treatment locations:", 
    n_distinct(admission_inla_data$TreatmentLocation), "\n")

# Verify time and month indices are what INLA's rw2/rw1 expect
range(admission_inla_data$t_rounded)
table(admission_inla_data$m_num)

# Location ID mapping (for interpreting site random effects later)
location_mapping <- admission_inla_data %>%
  distinct(TreatmentLocation, TreatmentLocation_id, HBT) %>%
  arrange(TreatmentLocation_id)

location_mapping
```

## 3.2 Baseline model 

```{r}
# Baseline model age and time trend
baseline_adm_inla <- inla(admitted ~ Age +
                            f(t_rounded, model = "rw2", constr = TRUE),
                          family = "binomial",
                          Ntrials = admission_inla_data$total_attendances,
                          data = admission_inla_data,
                          control.compute = list(waic = TRUE))

summary(baseline_adm_inla)
```

## 3.3 Coefficient interpretation

```{r}
fixed_effects <- baseline_adm_inla$summary.fixed
fixed_effects$odds_ratio <- exp(fixed_effects$mean)
fixed_effects$or_lower <- exp(fixed_effects$`0.025quant`)
fixed_effects$or_upper <- exp(fixed_effects$`0.975quant`)
fixed_effects
```

Check against the descriptive admission rates:
- 75 plus show the highest odds of admission
- Under 18 show lower odds than the 18-24 reference
- The odds increase monotonically with age


## 3.4 Hyperparameters

```{r}
baseline_adm_inla$summary.hyperpar
```

## 3.5 Time trend plot

```{r}
# Time trend
plot(baseline_adm_inla$summary.random$t_rounded$ID,
     baseline_adm_inla$summary.random$t_rounded$mean,
     type = "l", xlab = "t (months since start)", ylab = "Effect",
     main = "Time trend f(t)")
```

## 3.6 Baseline WAIC

```{r}
initial_log <- tribble(
  ~model, ~waic, ~delta_waic, ~note,
  "baseline", baseline_adm_inla$waic$waic, 0,
  "Age + time"
)

initial_log
```


# 4. Building model structure

Starting from the Age + time baseline, add structural model components one at a time.

## 4.1 + Seasonality

```{r}
m_season <- inla(
  admitted ~ Age +
    f(t_rounded, model = "rw2", constr = TRUE) +
    f(m_num, model = "rw1", cyclic = TRUE, constr = TRUE),
  family = "binomial",
  Ntrials = admission_inla_data$total_attendances,
  data = admission_inla_data,
  control.compute = list(waic = TRUE))

summary(m_season)

initial_log <- initial_log %>% add_row(
  model = "+ seasonality",
  waic = m_season$waic$waic,
  delta_waic = m_season$waic$waic - baseline_waic,
  note = "f(m_num, rw1, cyclic)")

initial_log
```


## 4.2 + Site random effect

```{r}
m_site <- inla(
  admitted ~ Age +
    f(t_rounded, model = "rw2", constr = TRUE) +
    f(m_num, model = "rw1", cyclic = TRUE, constr = TRUE) +
    f(TreatmentLocation_id, model = "iid"),
  family = "binomial",
  Ntrials = admission_inla_data$total_attendances,
  data = admission_inla_data,
  control.compute = list(waic = TRUE))

summary(m_site)

initial_log <- initial_log %>% add_row(
  model = "+ site RE",
  waic = m_site$waic$waic,
  delta_waic = m_site$waic$waic - baseline_waic,
  note = "f(TreatmentLocation_id, iid)")

initial_log
```


## 4.3 Observation level random effect

```{r}
m_olre <- inla(
  admitted ~
    Age +
    f(t_rounded, model = "rw2", constr = TRUE) +
    f(m_num, model = "rw1", cyclic = TRUE, constr = TRUE) +
    f(TreatmentLocation_id, model = "iid") +
    f(obs_id, model = "iid"),
  family = "binomial",
  Ntrials = admission_inla_data$total_attendances,
  data = admission_inla_data,
 control.compute = list(dic = TRUE, waic = TRUE, config = TRUE),
  control.predictor = list(compute = TRUE))

summary(m_olre)

initial_log <- initial_log %>% add_row(
  model = "+ obs RE",
  waic = m_olre$waic$waic,
  delta_waic = m_olre$waic$waic - baseline_waic,
  note = "f(obs_id, iid) overdispersion handling"
)

initial_log
```

## 4.4 OLRE check

```{r}
# Beta-Binomial alternative (uniform overdispersion, no obs RE needed)
m_betabin <- inla(
  admitted ~ Age +
    f(t_rounded, model = "rw2", constr = TRUE) +
    f(m_num, model = "rw1", cyclic = TRUE, constr = TRUE) +
    f(TreatmentLocation_id, model = "iid"),
  family = "betabinomial",
  Ntrials = admission_inla_data$total_attendances,
  data = admission_inla_data,
  control.compute = list(waic = TRUE)
)

# Three-way comparison
tibble(
  model = c("Without OLRE (site RE only)", "With OLRE", "Beta-Binomial"),
  WAIC = c(m_site$waic$waic, m_olre$waic$waic, m_betabin$waic$waic)
)
```

## 4.5 Summary

```{r}
initial_log %>% arrange(delta_waic)
```

Each structural component produces a substantial WAIC improvement, with
the site RE and obs RE producing the two largest drops. The full structural
model (Age + time + seasonality + site RE + obs RE) becomes the reference
for testing contextual predictors.

```{r}
# Set the full structural model as reference for Section 5
baseline_admission_inla <- m_olre
baseline_waic_full <- m_olre$waic$waic

summary(baseline_admission_inla)
baseline_admission_inla$summary.hyperpar
```

## 4.6 Full Baseline Model Plots

```{r}
par(mfrow = c(1, 3))

plot(baseline_admission_inla$summary.random$t_rounded$ID,
     baseline_admission_inla$summary.random$t_rounded$mean,
     type = "l", xlab = "t (months since start)", ylab = "Effect",
     main = "Time trend f(t)")

plot(baseline_admission_inla$summary.random$m_num$ID,
     baseline_admission_inla$summary.random$m_num$mean,
     type = "l", xlab = "Month", ylab = "Effect",
     main = "Seasonal cycle f(m_num)")

site_re <- baseline_admission_inla$summary.random$TreatmentLocation_id
plot(sort(site_re$mean), seq_along(site_re$mean),
     xlab = "Random intercept", ylab = "Site rank",
     main = "Site random effects", pch = 16)
abline(v = 0, lty = 2)

par(mfrow = c(1, 1))
```


# 5. Contextual Predictors

Contextual predictors tested individually against the full structural model

```{r}
predictor_log <- tribble(
  ~model, ~waic, ~delta_waic, ~note,
  "full structural (reference)", baseline_waic_full, 0,
  "Age + time + season + site RE + obs RE"
)
```

## 5.1 + Self Referral Share

```{r}
m_self_referral <- inla(
  admitted ~ Age + pct_ref_self +
    f(t_rounded, model = "rw2", constr = TRUE) +
    f(m_num, model = "rw1", cyclic = TRUE, constr = TRUE) +
    f(TreatmentLocation_id, model = "iid") +
    f(obs_id, model = "iid"),
  family = "binomial",
  Ntrials = admission_inla_data$total_attendances,
  data = admission_inla_data,
  control.compute = list(waic = TRUE))

summary(m_self_referral)

predictor_log <- predictor_log %>% add_row(
  model = "+ Self Referral %",
  waic = m_self_referral$waic$waic,
  delta_waic = m_self_referral$waic$waic - baseline_waic,
  note = "Self referral share (site x month x age level)"
)

predictor_log
```

## 5.2 + Ambulance referral

```{r}
m_ambulance <- inla(
  admitted ~
    Age + pct_ref_ambulance +
    f(t_rounded, model = "rw2", constr = TRUE) +
    f(m_num, model = "rw1", cyclic = TRUE, constr = TRUE) +
    f(TreatmentLocation_id, model = "iid") +
    f(obs_id, model = "iid"),
  family = "binomial",
  Ntrials = admission_inla_data$total_attendances,
  data = admission_inla_data,
  control.compute = list(dic = TRUE, waic = TRUE))

summary(m_ambulance)

predictor_log <- predictor_log %>% add_row(
  model = "+ Ambulance %",
  waic = m_ambulance$waic$waic,
  delta_waic = m_ambulance$waic$waic - baseline_waic,
  note = "Ambulance referral share (site x month x age level)"
)

predictor_log
```


## 5.3 + Out of hours share

```{r}
m_ooh <- inla(
  admitted ~
    Age + oohours_pct +
    f(t_rounded, model = "rw2", constr = TRUE) +
    f(m_num, model = "rw1", cyclic = TRUE, constr = TRUE) +
    f(TreatmentLocation_id, model = "iid") +
    f(obs_id, model = "iid"),
  family = "binomial",
  Ntrials = admission_inla_data$total_attendances,
  data = admission_inla_data,
  control.compute = list(dic = TRUE, waic = TRUE)
)

summary(m_ooh)

predictor_log <- predictor_log %>% add_row(
  model = "+ oohours_pct",
  waic = m_ooh$waic$waic,
  delta_waic = m_ooh$waic$waic - baseline_waic,
  note = "Out-of-hours share (site x month level)")

predictor_log
```


## 5.4 + Income deprivation (HB-level context)

```{r}
m_income <- inla(
  admitted ~
    Age + Income_rate +
    f(t_rounded, model = "rw2", constr = TRUE) +
    f(m_num, model = "rw1", cyclic = TRUE, constr = TRUE) +
    f(TreatmentLocation_id, model = "iid") +
    f(obs_id, model = "iid"),
  family = "binomial",
  Ntrials = admission_inla_data$total_attendances,
  data = admission_inla_data,
  control.compute = list(dic = TRUE, waic = TRUE)
)

summary(m_income)

predictor_log <- predictor_log %>% add_row(
  model = "+ Income_rate",
  waic = m_income$waic$waic,
  delta_waic = m_income$waic$waic - baseline_waic,
  note = "HB-level income deprivation rate"
)

predictor_log
```

## 5.5 + Within site referral deviations

Testing whether a site's referral mix or weekend share in a specific month
deviating from its own average predicts admission.

```{r}
# Within-site deviation from the site's own mean
admission_inla_data <- admission_inla_data %>%
  group_by(TreatmentLocation_id) %>%
  mutate(
    self_ref_deviation = pct_ref_self - mean(pct_ref_self, na.rm = TRUE),
    weekend_deviation = weekend_pct - mean(weekend_pct, na.rm = TRUE)
  ) %>%
  ungroup()

m_self_dev <- inla(
  admitted ~
    Age + self_ref_deviation +
    f(t_rounded, model = "rw2", constr = TRUE) +
    f(m_num, model = "rw1", cyclic = TRUE, constr = TRUE) +
    f(TreatmentLocation_id, model = "iid") +
    f(obs_id, model = "iid"),
  family = "binomial",
  Ntrials = admission_inla_data$total_attendances,
  data = admission_inla_data,
  control.compute = list(dic = TRUE, waic = TRUE)
)

summary(m_self_dev)

predictor_log <- predictor_log %>% add_row(
  model = "+ Self ref deviation",
  waic = m_self_dev$waic$waic,
  delta_waic = m_self_dev$waic$waic - baseline_waic,
  note = "Within-site self-referral deviation"
)

predictor_log
```

```{r}
m_weekend_dev <- inla(
  admitted ~
    Age + weekend_deviation +
    f(t_rounded, model = "rw2", constr = TRUE) +
    f(m_num, model = "rw1", cyclic = TRUE, constr = TRUE) +
    f(TreatmentLocation_id, model = "iid") +
    f(obs_id, model = "iid"),
  family = "binomial",
  Ntrials = admission_inla_data$total_attendances,
  data = admission_inla_data,
  control.compute = list(dic = TRUE, waic = TRUE)
)

summary(m_weekend_dev)

predictor_log <- predictor_log %>% add_row(
  model = "+ Weekend deviation",
  waic = m_weekend_dev$waic$waic,
  delta_waic = m_weekend_dev$waic$waic - baseline_waic,
  note = "Within-site weekend deviation"
)

predictor_log
```

## 5.6 Predictor WAIC Summary

```{r}
predictor_log %>% arrange(delta_waic)
```

No contextual predictor improved the full structural model.


# 6. Final model

Based on the results above, the full structural model is the final model.

```{r}
# The baseline is the final model
final_admission <- baseline_admission_inla

# Final coefficient table
final_coefs_adm <- final_admission$summary.fixed
final_coefs_adm$odds_ratio <- exp(final_coefs_adm$mean)
final_coefs_adm$or_lower <- exp(final_coefs_adm$`0.025quant`)
final_coefs_adm$or_upper <- exp(final_coefs_adm$`0.975quant`)
final_coefs_adm
```

## 6.1 Final model diagnostics

```{r}
par(mfrow = c(1, 3))

plot(final_admission$summary.random$t_rounded$ID,
     final_admission$summary.random$t_rounded$mean,
     type = "l", xlab = "t (months since start)", ylab = "Effect",
     main = "Time trend f(t)")

plot(final_admission$summary.random$m_num$ID,
     final_admission$summary.random$m_num$mean,
     type = "l", xlab = "Month", ylab = "Effect",
     main = "Seasonal cycle f(m_num)")

site_re <- final_admission$summary.random$TreatmentLocation_id
plot(sort(site_re$mean), seq_along(site_re$mean),
     xlab = "Random intercept", ylab = "Site rank",
     main = "Site random effects", pch = 16)
abline(v = 0, lty = 2)

par(mfrow = c(1, 1))
```


```{r}
# Hyperparameter summaries
final_admission$summary.hyperpar
```


# 7. Sensitivity: narrow vs broad admission definition

Compare the final model under both definitions to check whether results
are robust to including transfers ("Transferred to Other Hospital/Service")

```{r}
location_levels <- levels(factor(admission_inla_data$TreatmentLocation))

admission_inla_data_broad <- admission_model_data_broad %>%
  filter(DepartmentType == "Type 1",
         !is.na(Age), Age != "",
         !is.na(admitted), total_attendances > 0) %>%
  rename(
    pct_ref_self = `pct_ref_Self Referral`,
    pct_ref_ambulance = `pct_ref_Ambulance`
  ) %>%
  mutate(
    Age = factor(Age),
    TreatmentLocation_id = as.numeric(factor(TreatmentLocation,
                                              levels = location_levels)),
    t_rounded = round(as.numeric(m_date - min(m_date)) / 30),
    obs_id = row_number()
  )

# Refit BASELINE specification with broad definition
model_adm_broad <- inla(
  admitted ~
    Age +
    f(t_rounded, model = "rw2", constr = TRUE) +
    f(m_num, model = "rw1", cyclic = TRUE, constr = TRUE) +
    f(TreatmentLocation_id, model = "iid") +
    f(obs_id, model = "iid"),
  family = "binomial",
  Ntrials = admission_inla_data_broad$total_attendances,
  data = admission_inla_data_broad,
  control.compute = list(dic = TRUE, waic = TRUE)
)

# Compare coefficients side by side
narrow_fixed <- final_admission$summary.fixed %>%
  as.data.frame() %>%
  mutate(term = rownames(final_admission$summary.fixed),
         or_narrow = exp(mean)) %>%
  select(term, mean_narrow = mean, or_narrow)

broad_fixed <- model_adm_broad$summary.fixed %>%
  as.data.frame() %>%
  mutate(term = rownames(model_adm_broad$summary.fixed),
         or_broad = exp(mean)) %>%
  select(term, mean_broad = mean, or_broad)

sensitivity_comparison <- narrow_fixed %>%
  left_join(broad_fixed, by = "term") %>%
  mutate(or_difference = round(or_narrow - or_broad, 3))

sensitivity_comparison
```

The narrow definition produced a steeper age gradient at every level, reflecting
the inclusion of inter hospital transfers in the broad definition.
