---
title: "PHS Multiple Attendance Analysis"
author: "Natalia Montalvo Cabornero"
date: "2026-08-14"
output: html_document
---

```{r setup, include=FALSE}

source("R/mul_att_model_data.R")

# Theme for plotting created in prep.R, sourced through mul_att_model_data.R
theme_set(theme_report())


# Make sure the figure output folders exist
dir.create("figures/mul_attendance", recursive = TRUE, showWarnings = FALSE)
```

# 1. Data Recap

Among people who attended A&E at least once in a 12 month period, what
demographic factors are associated with becoming a frequent attender
(5+ visits)?

**Data**: `mul_att_data`, a single 12 month rolling snapshot, with patient
counts bucketed by visit frequency (1, 2, 3, 4, 5+), broken down by
Age, Sex, Deprivation, and HBT.

```{r}
dim(mul_att_data)
summarise_columns(mul_att_data)
```

## 1.1 Overall Frequent Attender Rate

```{r}
freq_summary <- mul_att_long %>%
  filter(!is.na(Age)) %>%
  group_by(attendance_freq) %>%
  summarise(n_patients = sum(n_patients, na.rm = TRUE), .groups = "drop") %>%
  mutate(pct = n_patients / sum(n_patients) * 100)

freq_summary
```

## 1.2 Frequent Attender Rate by Demographics

```{r}
# By deprivation
mul_att_long %>%
  filter(!is.na(Age), !is.na(Deprivation)) %>%
  group_by(Deprivation) %>%
  summarise(
    total_patients = sum(n_patients, na.rm = TRUE),
    frequent = sum(n_patients[attendance_freq == "5+"], na.rm = TRUE),
    freq_rate = frequent / total_patients * 100,
    .groups = "drop"
  )
```

```{r}
# By age
mul_att_long %>%
  filter(!is.na(Age)) %>%
  group_by(Age) %>%
  summarise(
    total_patients = sum(n_patients, na.rm = TRUE),
    frequent = sum(n_patients[attendance_freq == "5+"], na.rm = TRUE),
    freq_rate = frequent / total_patients * 100,
    .groups = "drop"
  ) %>%
  arrange(desc(freq_rate))
```

```{r}
# By sex
mul_att_long %>%
  filter(!is.na(Age), !is.na(Sex), Sex %in% c("Male", "Female")) %>%
  group_by(Sex) %>%
  summarise(
    total_patients = sum(n_patients, na.rm = TRUE),
    frequent = sum(n_patients[attendance_freq == "5+"], na.rm = TRUE),
    freq_rate = frequent / total_patients * 100,
    .groups = "drop"
  )
```


# 2. Baseline model

Since there is no time dimension, a plain GLM is appropriate. The outcome is 
modelled as `cbind(frequent, not_frequent)` with a binomial family. 

```{r}
mul_baseline_bin <- glm(
  cbind(frequent, not_frequent) ~ Age + Sex + Deprivation,
  family = binomial(),
  data = mul_model_data
)

summary(mul_baseline_bin)
```

## 2.1 Overdispersion Check

```{r}
dispersion_ratio_bin <- sum(residuals(mul_baseline_bin, type ="deviance")^2)/mul_baseline_bin$df.residual

cat("Dispersion ratio:", round(dispersion_ratio_bin, 2), "\n")
```

Dispersion ratio substantially exceeds 1, indicating overdispersion. 
The binomial variance assumption underestimates the true variability, switching 
to `quasibinomial()`.

```{r}
mul_baseline <- glm(
  cbind(frequent, not_frequent) ~ Age + Sex + Deprivation,
  family = quasibinomial(),
  data = mul_model_data
)

summary(mul_baseline)
cat("Estimated dispersion:", round(summary(mul_baseline)$dispersion, 2), "\n")
```

## 2.2 Coefficient Interpretation

```{r}
mul_coefs <- as.data.frame(summary(mul_baseline)$coefficients)
mul_coefs$odds_ratio <- exp(mul_coefs$Estimate)
mul_coefs$or_lower <- exp(mul_coefs$Estimate - 1.96 * mul_coefs$`Std. Error`)
mul_coefs$or_upper <- exp(mul_coefs$Estimate + 1.96 * mul_coefs$`Std. Error`)
mul_coefs
```

Cross-check against descriptive rates:

- **Age**: 75 plus shows highest odds 
- **Sex**: should be non-significant, but sex makes no difference to whether 
           they become frequent attenders
- **Deprivation**: shows monotonic gradient

## 2.3 Diagnostics

```{r}
pdf("figures/mul_attendance/diagnostics_baseline.pdf", width = 8, 
    height = 6)
par(mfrow = c(2, 2))
plot(mul_baseline)
par(mfrow = c(1, 1))
dev.off()
```

## 2.4 Record Baseline deviance

```{r}
# Deviance for comparison
baseline_deviance <- mul_baseline$deviance
baseline_df <- mul_baseline$df.residual

cat("Baseline deviance:", round(baseline_deviance, 1), "\n")
cat("Baseline df:", baseline_df, "\n")
```


# 3. Adding Predictors

## 3.1 + HBT (Health Board)

Test whether board level variation matters for frequent attendance beyond
what demographics explain.

```{r}
mul_model_hbt <- glm(
  cbind(frequent, not_frequent) ~ Age + Sex + Deprivation + HBT,
  family = quasibinomial(),
  data = mul_model_data %>% mutate(HBT = factor(HBT))
)

summary(mul_model_hbt)

anova(mul_baseline, mul_model_hbt, test = "F")
```

HBT significantly improves model fit. Island and rural boards (Shetland,
Orkney, Dumfries & Galloway) show the highest frequent-attender odds after
demographic adjustment, consistent with limited alternative services
driving repeat A&E use in these areas.

## 3.2 + DepartmentType

```{r}
mul_model_dept <- glm(
  cbind(frequent, not_frequent) ~ Age + Sex + Deprivation + HBT + DepartmentType,
  family = quasibinomial(),
  data = mul_model_data)

summary(mul_model_dept)
anova(mul_model_hbt, mul_model_dept, test = "F")
```

Type 3 patients have substantially lower odds of becoming frequent
attenders than Type 1 (OR 0.42), clinically expected, since minor
injuries are typically one-off events rather than recurring conditions.

## 3.3 Domain Indicators Replacing Deprivation (Sensitivity Check)

Same approach as the risk model's Model B: replace the deprivation
quintile with individual SIMD domain indicators to identify which
specific aspect of deprivation drives frequent attendance.

```{r}
mul_model_indicators <- glm(
  cbind(frequent, not_frequent) ~ Age + Sex + Income_rate + EMERG +
    crime_rate + drive_GP + HBT + DepartmentType,
  family = quasibinomial(),
  data = mul_model_data_simd
)

summary(mul_model_indicators)
```

Key contrast with the risk model: income deprivation was the dominant
predictor in Step 1, but in Step 3, EMERG (historical emergency
hospitalisation ratio) is the only significant domain indicator.
This suggests that frequent attendance is driven by chronic health burden
rather than income deprivation per se. However, interpretation is limited
by confounding between HB-level indicators and HBT fixed effects.


## 3.4 Age x Deprivation interaction

```{r}
mul_model_interaction <- glm(
  cbind(frequent, not_frequent) ~ Age * Deprivation + Sex + HBT + DepartmentType,
  family = quasibinomial(),
  data = mul_model_data)

anova(mul_model_dept, mul_model_interaction, test = "F")
summary(mul_model_interaction)
```


# 4. Final Model

The model with HBT and DepartmentType is selected as the final
specification, using the deprivation quintile (not domain indicators)
for interpretability.

```{r}
mul_final <- glm(
  cbind(frequent, not_frequent) ~ Age * Deprivation + Sex + HBT + DepartmentType,
  family = quasibinomial(),
  data = mul_model_data)

summary(mul_final)
cat("Estimated dispersion:", round(summary(mul_final)$dispersion, 2), "\n")
```


## 4.1 Diagnostics

```{r}
pdf("figures/mul_attendance/diagnostics_final.pdf", width = 8, height = 6)
par(mfrow = c(2, 2))
plot(mul_final)
par(mfrow = c(1, 1))
dev.off()
```


## 4.2 Coefficient Interpretation

```{r}
final_coefs_mul <- as.data.frame(summary(mul_final)$coefficients)
final_coefs_mul$odds_ratio <- exp(final_coefs_mul$Estimate)
final_coefs_mul$or_lower <- exp(final_coefs_mul$Estimate - 1.96 * final_coefs_mul$`Std. Error`)
final_coefs_mul$or_upper <- exp(final_coefs_mul$Estimate + 1.96 * final_coefs_mul$`Std. Error`)
final_coefs_mul
```


## 4.3 HBT Ranking

```{r}
hbt_coefs_mul <- final_coefs_mul %>%
  mutate(term = rownames(final_coefs_mul)) %>%
  filter(grepl("^HBT", term)) %>%
  mutate(HBT = gsub("^HBT", "", term)) %>%
  left_join(hb_names, by = "HBT") %>%
  # Add reference board at OR = 1
  bind_rows(tibble(
    HBT = levels(factor(mul_model_data$HBT))[1],
    HBName = hb_names$HBName[hb_names$HBT == levels(factor(mul_model_data$HBT))[1]],
    odds_ratio = 1, or_lower = 1, or_upper = 1
  )) %>%
  arrange(desc(odds_ratio))

hbt_coefs_mul
```


```{r}
hbt_freq_plt <- ggplot(hbt_coefs_mul,
                        aes(x = fct_reorder(HBName, odds_ratio),
                            y = odds_ratio)) +
  geom_point(aes(colour = odds_ratio == 1), size = 3, show.legend = FALSE) +
  scale_colour_manual(values = c("TRUE" = "grey50",
                                  "FALSE" = phs_colours("phs-blue"))) +
  geom_errorbar(aes(ymin = or_lower, ymax = or_upper), width = 0.3,
                colour = phs_colours("phs-blue")) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey30") +
  coord_flip() +
  labs(x = NULL,
       y = "Odds Ratio (vs. reference HB)",
       title = "Health Board Frequent Attender Odds Ratios\n(adjusted for Age, Sex, Deprivation, DepartmentType)") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

hbt_freq_plt

ggsave("figures/mul_attendance/hbt_freq_odds_ratios.pdf",
       plot = hbt_freq_plt, width = 8, height = 6)
```


# 5. Sensitivity: Repeat Attender Definition (2+ vs 1)

As a sensitivity check, redefine the outcome as any repeat attendance
(2+ visits) rather than the clinical frequent-attender threshold (5+).

```{r}
cat("Rows:", nrow(mul_model_data_2), "\n")
cat("Total patients:", sum(mul_model_data_2$total_patients), "\n")
cat("Repeat attenders:", sum(mul_model_data_2$repeat_att), "\n")
cat("Overall repeat rate:",
    round(sum(mul_model_data_2$repeat_att) / 
            sum(mul_model_data_2$total_patients) * 100, 1), "%\n")
```

```{r}
mul_sensitivity <- glm(
  cbind(repeat_att, not_repeat_att) ~ Age + Sex + Deprivation + HBT + 
    DepartmentType,
  family = quasibinomial(),
  data = mul_model_data_2)

summary(mul_sensitivity)
```

```{r}
# Compare deprivation coefficients: 5+ vs 2+ definition
primary_dep <- final_coefs_mul %>%
  mutate(term = rownames(final_coefs_mul)) %>%
  filter(grepl("^Deprivation", term)) %>%
  select(term, or_primary = odds_ratio)

sensitivity_coefs <- as.data.frame(summary(mul_sensitivity)$coefficients)
sensitivity_coefs$odds_ratio <- exp(sensitivity_coefs$Estimate)
sensitivity_dep <- sensitivity_coefs %>%
  mutate(term = rownames(sensitivity_coefs)) %>%
  filter(grepl("^Deprivation", term)) %>%
  select(term, or_sensitivity = odds_ratio)

dep_comparison <- primary_dep %>%
  left_join(sensitivity_dep, by = "term")

dep_comparison
```

The deprivation gradient is steeper under the clinical frequent attender ´
definition (5+) than under the broader repeat attender definition (2+): Q5 vs 
Q1 odds ratio is 0.48 for frequent attenders versus 0.62 for repeat attenders. 
This confirms that deprivation disproportionately concentrates at the highest 
frequency end of the attendance distribution


