---
title: "EDA"
author: "Natalia"
date: "2026-07-14"
output: html_document
---

```{r setup, include=FALSE}
library(readxl)
library(ggplot2)
library(dplyr)
library(tidyr)
library(INLA)
library(mgcv)
library(tidyverse)
library(lubridate)
library(phsstyles)
```

# 1. Load main data sets

```{r}
month_demo_data <- read.csv("datasets/opendata_monthly_ae_demographics_202604.csv")
# month_activity_data <- read.csv("datasets/monthly_ae_activity_202605.csv")
week_activity_data <- read.csv("datasets/weekly_ae_activity_20260628.csv")

# Treatment location data set
trt_loc <- read.csv("datasets/ae_hospital_site_list_09_09_2025.csv")

# SIMD lookup 
simd_lookup <- read_xlsx("datasets/SIMD_2020v2_postcode_lookup_updated_2025.xlsx", sheet = 2)
```

```{r}
# Function to get initial summaries from data sets
summarise_columns <- function(data) {
  
  # Blank and : as NA
  is_missing <- function(x) {
    if (is.character(x)) {
      is.na(x) | trimws(x) %in% c("", ":")
    } else {
      is.na(x)
    }
  }
  
  data %>%
    summarise(
      across(everything(),
             list(
               type        = ~ class(.)[1],
               n_rows      = ~ length(.),
               n_missing   = ~ sum(is_missing(.)),
               pct_missing = ~ round(mean(is_missing(.)) * 100, 2),
               n_unique    = ~ n_distinct(.[!is_missing(.)]),
               options     = ~ paste(sort(unique(as.character(.[!is_missing(.)]))),
                                     collapse = ", ")
             ),
             .names = "{.col}__{.fn}"
      )
    ) %>%
    pivot_longer(everything(),
                 names_to = c("variable", ".value"),
                 names_sep = "__")
}
```

# 2. Monthly demographics data set

```{r}
summarise_columns(month_demo_data)
head(month_demo_data)
```

```{r}
# Convert date 
month_demo_data <- month_demo_data %>%
  mutate(m_date = ym(as.character(Month)),
         year   = format(m_date, "%Y"),
         m_num  = as.integer(format(m_date, "%m")))

# Monthly demographics year trends
month_trend_data <- month_demo_data %>%
  group_by(year, m_num) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop")

# Plot attendances by month over years
m_demo_trend_plt <- ggplot(month_trend_data, 
                           aes(x = m_num, y = total, color = year, 
                               group = year)) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(x = "Month", 
       y = "Total Attendances",
       title = "A&E Attendances by Month Across Years") +
  theme_minimal(base_size = 14) +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

m_demo_trend_plt

ggsave("figures/m_demo/m_demo_trend_by_year.pdf", plot = m_demo_trend_plt,
       width = 8, height = 5)

```

```{r}
# Divide data by sex
sex_data <- month_demo_data %>%
  filter(Sex %in% c("Male", "Female")) %>%
  group_by(Sex) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop")

# Plot sex differences in attendance
sex_plt <- ggplot(sex_data, aes(x = Sex, y = total, fill = Sex)) +
  geom_col() +
  scale_fill_discrete_phs(palette = "main") +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Attendances by Sex", 
       x = "Sex", 
       y = "Total Attendances") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

sex_plt

ggsave("figures/m_demo/attendances_by_sex.pdf", plot = sex_plt, width = 6, 
       height = 5)
```

```{r}
# Monthly demographics year trends by sex
month_trend_sex_data <- month_demo_data %>%
  filter(Sex %in% c("Male", "Female")) %>%
  group_by(year, m_num, Sex) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop")

month_trend_sex_plt <- ggplot(month_trend_sex_data, 
                              aes(x = m_num, y = total,
                                  color = year, group = year)) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  facet_wrap(~ Sex, ncol = 2) +
  labs(x = "Month", y = "Total Attendances",
       title = "A&E Attendances by Month, Year, and Sex") +
  theme_phs() +
  theme(axis.text.x = element_text(size = 11, angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))

month_trend_sex_plt

ggsave("figures/m_demo/montrend_by_year_sex.pdf", plot = month_trend_sex_plt,
       width = 10, height = 5)
```


```{r}
# Divide attendances by deprivation index
deprivation_data <- month_demo_data %>%
  filter(!is.na(Deprivation)) %>%
  group_by(Deprivation) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop") %>%
  mutate(Deprivation = factor(Deprivation, levels = 1:5,
           labels = c("1 (most deprived)", "2", "3", "4", "5 (least deprived)")))

# Plotting attendances by deprivation quintile
deprivation_plt <- ggplot(deprivation_data, aes(x = Deprivation, y = total)) +
  geom_col(fill = phs_colours("phs-graphite")) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Deprivation Quintile", y = "Total Attendances",
       title = "Attendances by Deprivation Quintile") +
  theme_phs() +
  theme(axis.text.x = element_text(size = 11, angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))

deprivation_plt

ggsave("figures/m_demo/attendances_by_deprivation.pdf", plot = deprivation_plt, width = 8, height = 5)
```

```{r}
# Divide attendances by deprivation index and type of department
deprivation_dept_data <- month_demo_data %>%
  filter(!is.na(Deprivation)) %>%
  group_by(Deprivation, DepartmentType) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop") %>%
  mutate(Deprivation = factor(Deprivation, levels = 1:5,
           labels = c("1 (most deprived)", "2", "3", "4", "5 (least deprived)")))

# Plotting attendances by deprivation quintile and department type
deprivation_dept_plt <- ggplot(deprivation_dept_data, 
                               aes(x = Deprivation, y = total)) +
  geom_col(fill = phs_colours("phs-graphite")) +
  facet_wrap(~ DepartmentType, scales = "free_y") +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Deprivation Quintile", 
       y = "Total Attendances",
       title = "Attendances by Deprivation Quintile and Department Type") +
  theme_phs() +
  theme(axis.text.x = element_text(size = 10, angle = 30, hjust = 1),
        plot.title = element_text(hjust = 0.5))

deprivation_dept_plt

ggsave("figures/m_demo/deprivation_by_dept.pdf", plot = deprivation_dept_plt, 
       width = 10, height = 6)
```

```{r}
# Divide attendances by deprivation index and sex
deprivation_sex_data <- month_demo_data %>%
  filter(!is.na(Deprivation), Sex %in% c("Male", "Female")) %>%
  group_by(Deprivation, Sex) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop") %>%
  mutate(Deprivation = factor(Deprivation, levels = 1:5,
           labels = c("1 (most deprived)", "2", "3", "4", "5 (least deprived)")))

# Plotting attendances by deprivation quintile and sex
deprivation_sex_plt <- ggplot(deprivation_sex_data, aes(x = Deprivation, y = total, fill = Sex)) +
  geom_col(position = "dodge") +
  scale_fill_discrete_phs(palette = "main") +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Deprivation Quintile", y = "Total Attendances",
       title = "Attendances by Deprivation Quintile and Sex") +
  theme_phs() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5))

deprivation_sex_plt

ggsave("figures/m_demo/deprivation_by_sex.pdf", plot = deprivation_sex_plt, width = 8, height = 5)
```


```{r}
# Data by age groups
age_data <- month_demo_data %>%
  filter(!is.na(Age), Age != "") %>%
  group_by(Age) %>%
  summarise(total = sum(NumberOfAttendances, na.rm = TRUE), .groups = "drop")

# Plot attendances by age groups
age_plt <- ggplot(age_data, aes(x = fct_reorder(Age, total), y = total)) +
  geom_col(fill = phs_colours("phs-teal")) +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Age Group", y = "Total Attendances", title = "Attendances by Age Group") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

age_plt

ggsave("figures/m_demo/attendances_by_age.pdf", plot = age_plt, width = 8, height = 5)
```

```{r}
# Distribution of attendances
attendances_hist_plt <- ggplot(month_demo_data, 
                               aes(x = NumberOfAttendances)) +
  geom_histogram(bins = 50, fill = phs_colours("phs-purple"), alpha = 0.85) +
  labs(x = "Number of Attendances", 
       y = "Count",
       title = "Distribution of Number Of Attendances") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

attendances_hist_plt

ggsave("figures/m_demo/attendances_histogram.pdf", plot = attendances_hist_plt, width = 8, height = 5)

# Distribution of attendances log scale
attendances_hist_log_plt <- ggplot(month_demo_data, 
                                   aes(x = NumberOfAttendances)) +
  geom_histogram(bins = 50, fill = phs_colours("phs-purple"), alpha = 0.85) +
  scale_x_log10() +
  labs(x = "Number of Attendances (log scale)", 
       y = "Count",
       title = "Distribution of Number Of Attendances (log scale)") +
  theme_phs() +
  theme(plot.title = element_text(hjust = 0.5))

attendances_hist_log_plt

ggsave("figures/m_demo/attendances_histogram_log.pdf", plot = attendances_hist_log_plt, width = 8, height = 5)
```
```{r}

```

# 3. Weekly activity data

This data set is only Type 1 (ED)
And all the rows are duplicated a they have 2 categories: unplanned and planned, 
but there is a 3 category "all" that combines both.

32 treatment locations and 14 HBT


```{r}
summarise_columns(week_activity_data)
head(week_activity_data)
tail(week_activity_data)

week_activity_data <- week_activity_data %>%
  mutate(w_date = ymd(as.character(WeekEndingDate)),
         year    = format(w_date, "%Y"),
         m_num   = as.integer(format(w_date, "%m")))
```

# 4. Treatment location data

que localizaciones estan mas pobladas
cuales tienen mas a&e
cuales tiene mas attendacne
cuales tienen mas espera

```{r}
summarise_columns(trt_loc)
head(trt_loc)
```



```{r}
# Add location info to week activity data set
week_activity_data <- week_activity_data %>%
  left_join(trt_loc, by = c("TreatmentLocation" = "TreatmentLocationCode"))

week_activity_data %>%
  filter(is.na(TreatmentLocationName)) %>%
  distinct(TreatmentLocation)

# Which of 32 codes matched to what?
week_activity_data %>%
  distinct(TreatmentLocation, TreatmentLocationName, HB, CurrentDepartmentType, Status)
```

```{r}
# How many weeks per location in data set
location_nweeks <- week_activity_data %>%
  filter(AttendanceCategory == "All") %>%
  count(TreatmentLocation, name = "n_weeks_reported") %>%
  left_join(trt_loc, by = c("TreatmentLocation" = "TreatmentLocationCode")) %>%
  select(TreatmentLocation, TreatmentLocationName, HB, Status, n_weeks_reported) %>%
  arrange(n_weeks_reported)

location_nweeks
```

- G516H: West Glasgow Ambulatory Care Hospital S08000031 Closed (15 weeks)
- G306H: New Victoria Hospital S08000031Ope n (13 weeks)

```{r}
# Average attendance per location
location_attendance <- week_activity_data %>%
  filter(AttendanceCategory == "All") %>%
  group_by(TreatmentLocation) %>%
  summarise(
    n_weeks_reported = n(),
    total_attendance = sum(NumberOfAttendancesEpisode, na.rm = TRUE),
    avg_weekly_attendance = mean(NumberOfAttendancesEpisode, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(trt_loc, by = c("TreatmentLocation" = "TreatmentLocationCode")) %>%
  select(TreatmentLocation, TreatmentLocationName, HB, Status,
         n_weeks_reported, total_attendance, avg_weekly_attendance) %>%
  arrange(desc(total_attendance))

location_attendance

# Plot locations by number of attendances
location_attendance_plt <- ggplot(location_attendance, 
                                    aes(x = fct_reorder(TreatmentLocationName, avg_weekly_attendance), 
                                        y = avg_weekly_attendance)) +
  geom_col(fill = phs_colours("phs-blue")) +
  coord_flip() +
  scale_y_continuous(labels = scales::comma) +
  labs(x = NULL, y = "Average Weekly Attendances",
       title = "NHS Scotland A&E Locations by Average Weekly Attendance") +
  theme_phs() +
  theme(
    axis.text.y = element_text(size = 9),
    panel.grid.major.y = element_blank()
  )

location_attendance_plt

ggsave("figures/location_avg_attendance.pdf", plot = location_attendance_plt, 
       width = 8, height = 10)
```

```{r}
# Average attendance per location per year
location_year_attendance <- week_activity_data %>%
  filter(AttendanceCategory == "All") %>%
  group_by(year, TreatmentLocation) %>%
  summarise(
    n_weeks_reported = n(),
    total_attendance = sum(NumberOfAttendancesEpisode, na.rm = TRUE),
    avg_weekly_attendance = mean(NumberOfAttendancesEpisode, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(trt_loc, by = c("TreatmentLocation" = "TreatmentLocationCode")) %>%
  select(year, TreatmentLocation, TreatmentLocationName, HB, Status,
         n_weeks_reported, total_attendance, avg_weekly_attendance) %>%
  arrange(desc(total_attendance))

location_year_attendance

# Top 10 locations with most attendances
top10_locations <- location_year_attendance %>%
  group_by(TreatmentLocationName) %>%
  summarise(overall_avg = mean(avg_weekly_attendance, na.rm = TRUE),
            .groups = "drop") %>%
  slice_max(overall_avg, n = 10) %>%
  arrange(desc(overall_avg)) %>%
  pull(TreatmentLocationName)

top10_locations

# Plot locations by number of attendances
top10_trend_overlay_plt <- location_year_attendance %>%
  filter(TreatmentLocationName %in% top10_locations) %>%
  mutate(TreatmentLocationName = factor(TreatmentLocationName, levels = top10_locations)) %>%
  ggplot(aes(x = year, 
             y = avg_weekly_attendance, 
             colour = TreatmentLocationName, group = TreatmentLocationName)) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_y_continuous(labels = scales::comma) +
  labs(x = "Year", y = "Average Weekly Attendances", colour = "Location",
       title = "Top 10 A&E Locations: Attendance Trend by Year") +
  theme_phs() +
  theme(axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
        legend.position = "right",
        legend.text = element_text(size = 9),
        legend.title = element_text(size = 11),
        legend.key.size = unit(0.4, "cm"))

top10_trend_overlay_plt

ggsave("figures/top10_location_trend_overlay.pdf", plot = top10_trend_overlay_plt, width = 10, height = 6)

```


